---
title: Getting started with fmrilss
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Getting started with fmrilss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-options, include=FALSE}
plot_device <- if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(family = params$family, preset = params$preset)
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE,
  warning = TRUE,
  dev = plot_device,
  dpi = 150,
  fig.width = 7,
  fig.height = 4.5,
  fig.bg = "white"
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have a preprocessed fMRI run and an event design with one regressor per
trial. Your goal is a beta estimate for every trial and voxel, even though
nearby haemodynamic responses overlap. `fmrilss` fits each trial in a separate
model and returns the estimates as a trial-by-voxel matrix that can feed an
MVPA, RSA, connectivity, or reliability analysis.

This article follows one complete workflow: build a trial design, fit LSS,
check the result, handle nuisance regressors correctly, and compare LSS with a
simultaneous trial-wise GLM. The example is synthetic so that the truth is
known; in a real analysis, `Y` and the design matrices come from your
preprocessing and design pipeline.

## What are the inputs?

`lss()` uses matrices with time along the rows:

- `Y` contains the observed data: time points by voxels.
- `X` contains one HRF-convolved column per trial: time points by trials.
- `Z` contains regressors retained in every trial-wise model, such as run
  intercepts or trends.
- `Nuisance` contains regressors to project out, such as motion or
  physiological traces.

The function returns one row per trial and one column per voxel. Trial and
voxel names are preserved when `X` and `Y` have column names.

```{r attach}
suppressPackageStartupMessages({
  library(fmrilss)
  library(fmrihrf)
})
```

The following code constructs the trial design used throughout the article.
Each event is an impulse convolved with the same canonical HRF.

```{r build-trial-design}
n_time <- 160
n_trials <- 24
n_voxels <- 8
sframe <- sampling_frame(blocklens = n_time, TR = 1)
onsets <- round(seq(8, 112, length.out = n_trials))
rset <- regressor_set(
  onsets,
  factor(seq_len(n_trials)),
  hrf = HRF_SPMG1,
  duration = 0,
  span = 30,
  summate = FALSE
)
X <- evaluate(
  rset,
  grid = samples(sframe, global = TRUE),
  precision = 0.1,
  method = "conv"
)
X <- as.matrix(X)
colnames(X) <- sprintf("trial_%02d", seq_len(n_trials))
```

We add an intercept, a trend, and two centred nuisance traces. The final lines
create one synthetic `Y` with known trial effects. Replace this simulation with
your own preprocessed time-by-voxel matrix.

```{r simulate-data}
set.seed(2401)
noise_sd <- 2.5
Z <- cbind(
  Intercept = 1,
  Linear_trend = as.numeric(scale(seq_len(n_time), scale = FALSE))
)
Nuisance <- scale(cbind(
  Motion_x = sin(seq_len(n_time) / 9),
  Motion_y = cos(seq_len(n_time) / 13)
), center = TRUE, scale = FALSE)

true_betas <- matrix(
  rnorm(n_trials * n_voxels),
  n_trials,
  n_voxels,
  dimnames = list(colnames(X), sprintf("voxel_%02d", seq_len(n_voxels)))
)
fixed_coef <- matrix(rnorm(ncol(Z) * n_voxels), ncol(Z), n_voxels)
nuisance_coef <- matrix(
  rnorm(ncol(Nuisance) * n_voxels, sd = 1.2),
  ncol(Nuisance),
  n_voxels
)
mean_signal <- X %*% true_betas +
  Z %*% fixed_coef +
  Nuisance %*% nuisance_coef
Y <- mean_signal + matrix(
  rnorm(n_time * n_voxels, sd = noise_sd),
  n_time,
  n_voxels,
  dimnames = list(NULL, colnames(true_betas))
)
```

```{r simulate-check, include=FALSE}
stopifnot(
  all(dim(Y) == c(n_time, n_voxels)),
  all(dim(X) == c(n_time, n_trials)),
  all(is.finite(Y)),
  all(is.finite(X))
)
```

```{r input-shapes, echo=FALSE}
knitr::kable(
  data.frame(
    Object = c("Y", "X", "Z", "Nuisance", "LSS result"),
    Rows = c(nrow(Y), nrow(X), nrow(Z), nrow(Nuisance), ncol(X)),
    Columns = c(ncol(Y), ncol(X), ncol(Z), ncol(Nuisance), ncol(Y)),
    Role = c(
      "observed voxel time series",
      "one HRF-convolved column per trial",
      "shared modeled regressors",
      "projected nuisance regressors",
      "one estimate per trial and voxel"
    )
  ),
  caption = "Matrix shapes in the example workflow."
)
```

## How do I estimate the trial betas?

The default optimized R backend is the clearest first path. Supply all parts of
the model in the same call:

```{r fit-lss}
beta_lss <- lss(Y, X, Z = Z, Nuisance = Nuisance)
dim(beta_lss)
```

`beta_lss` is the promised `r nrow(beta_lss)` by `r ncol(beta_lss)`
trial-by-voxel matrix. A small preview is usually more useful than printing the
whole object:

```{r beta-preview}
round(beta_lss[1:4, 1:4], 2)
```

```{r beta-contract-check, include=FALSE}
direct_betas <- t(vapply(seq_len(n_trials), function(trial_id) {
  direct_design <- cbind(
    target = X[, trial_id],
    other_trials = rowSums(X) - X[, trial_id],
    Z,
    Nuisance
  )
  lm.fit(direct_design, Y)$coefficients[1L, ]
}, numeric(n_voxels)))
stopifnot(
  all(dim(beta_lss) == c(n_trials, n_voxels)),
  identical(rownames(beta_lss), colnames(X)),
  identical(colnames(beta_lss), colnames(Y)),
  all(is.finite(beta_lss)),
  max(abs(beta_lss - direct_betas)) < 1e-8
)
```

Each row comes from a model with three parts: the target trial, the sum of all
other trials, and the shared regressors. The target coefficient becomes the
reported beta. Repeating that model for every trial reduces competition among
individual trial columns, but it does not make a poor or rank-deficient design
informative.

```{r beta-heatmap, echo=FALSE, fig.cap="Estimated single-trial responses for the eight simulated voxels. Red is negative and blue is positive; deeper colour indicates larger magnitude, not statistical significance.", fig.alt="Heatmap with 24 trial rows and 8 voxel columns. Dark red cells are strongly negative and dark blue cells are strongly positive."}
op <- par(mar = c(4, 4, 1, 1), bg = "white")
image(
  x = seq_len(ncol(beta_lss)),
  y = seq_len(nrow(beta_lss)),
  z = t(beta_lss),
  axes = FALSE,
  xlab = "Voxel",
  ylab = "Trial",
  col = hcl.colors(31, "Blue-Red 3", rev = TRUE)
)
axis(1, at = seq_len(ncol(beta_lss)), labels = colnames(beta_lss), las = 2, cex.axis = 0.7)
axis(2, at = c(1, 6, 12, 18, 24), labels = c(1, 6, 12, 18, 24), las = 1)
box()
par(op)
```

## How should I handle nuisance regressors?

The safest route is to pass `Nuisance` directly to `lss()`, as above. The
package applies a rank-aware Frisch--Waugh--Lovell step for the shared `Z` and
`Nuisance` span before fitting the trial models.

If you deliberately cache a projection for repeated analyses, apply it to
every matrix that remains in the model. Here we project only `Nuisance`, so
the same projection must transform `Y`, `X`, and `Z`:

```{r projection-equivalence}
Q_nuisance <- project_confounds(Nuisance)
beta_projected <- lss(
  Q_nuisance %*% Y,
  Q_nuisance %*% X,
  Z = Q_nuisance %*% Z
)
c(maximum_absolute_difference = max(abs(beta_lss - beta_projected)))
```

```{r projection-equivalence-check, include=FALSE}
projection_error <- max(abs(beta_lss - beta_projected))
Nuisance_redundant <- cbind(
  Nuisance,
  Duplicate_motion_x = 2 * Nuisance[, "Motion_x"]
)
beta_redundant <- lss(Y, X, Z = Z, Nuisance = Nuisance_redundant)
Q_redundant <- project_confounds(Nuisance_redundant)
beta_redundant_projected <- lss(
  Q_redundant %*% Y,
  Q_redundant %*% X,
  Z = Q_redundant %*% Z
)
redundant_projection_error <- max(abs(
  beta_redundant - beta_redundant_projected
))
stopifnot(
  is.finite(projection_error),
  projection_error < 1e-8,
  is.finite(redundant_projection_error),
  redundant_projection_error < 1e-8
)
```

Leaving `Z` unprojected changes the model and can change the trial estimates.
Also note that `project_confounds()` materializes an \(n \times n\) matrix;
passing `Nuisance` to `lss()` avoids that storage and should be your default.

## When does LSS differ from LSA?

Least Squares All (LSA) estimates every trial column simultaneously. LSS
replaces the non-target columns with their sum in each trial-specific model.
That creates a bias--variance trade-off rather than a universal winner.

To make the comparison reproducible, the next experiment holds `X`,
`true_betas`, and all nuisance effects fixed. Because both estimators are
linear here, we can calculate conditional squared bias and noise variance
directly. We also generate 100 independent noise realizations as a check on
that calculation.

```{r repeated-comparison, include=FALSE}
shared_projection <- project_confounds(cbind(Z, Nuisance))
X_residual <- shared_projection %*% X
X_norm <- sweep(X_residual, 2, sqrt(colSums(X_residual^2)), "/")
X_cor <- cor(X_residual)
design_diagnostics <- data.frame(
  Diagnostic = c(
    "Residualized trial-design rank",
    "Largest absolute trial correlation",
    "Scaled condition number"
  ),
  Value = c(
    qr(X_residual)$rank,
    max(abs(X_cor[upper.tri(X_cor)])),
    kappa(X_norm, exact = TRUE)
  )
)

# Exact conditional means and unit-noise linear operators.
lss_mean <- lss(mean_signal, X, Z = Z, Nuisance = Nuisance)
lsa_mean <- lsa(mean_signal, X, Z = cbind(Z, Nuisance))
lss_operator <- lss(diag(n_time), X, Z = Z, Nuisance = Nuisance)
lsa_operator <- lsa(diag(n_time), X, Z = cbind(Z, Nuisance))

squared_bias <- c(
  LSS = mean((lss_mean - true_betas)^2),
  LSA = mean((lsa_mean - true_betas)^2)
)
variance_factor <- c(
  LSS = mean(rowSums(lss_operator^2)),
  LSA = mean(rowSums(lsa_operator^2))
)
variance_at_example <- noise_sd^2 * variance_factor
exact_mse <- squared_bias + variance_at_example

set.seed(2402)
n_repetitions <- 100L
lss_repetitions <- array(
  NA_real_,
  dim = c(n_trials, n_voxels, n_repetitions)
)
lsa_repetitions <- array(
  NA_real_,
  dim = c(n_trials, n_voxels, n_repetitions)
)

for (iteration in seq_len(n_repetitions)) {
  Y_iteration <- mean_signal + matrix(
    rnorm(n_time * n_voxels, sd = noise_sd),
    n_time,
    n_voxels
  )
  lss_repetitions[, , iteration] <- lss(
    Y_iteration,
    X,
    Z = Z,
    Nuisance = Nuisance
  )
  lsa_repetitions[, , iteration] <- lsa(
    Y_iteration,
    X,
    Z = cbind(Z, Nuisance)
  )
}

monte_carlo_mse <- function(estimates, truth) {
  errors <- sweep(estimates, c(1, 2), truth, "-")
  mean(errors^2)
}

comparison <- data.frame(
  Estimator = c("LSS", "LSA"),
  `Squared bias` = squared_bias,
  Variance = variance_at_example,
  MSE = exact_mse,
  `Monte Carlo MSE` = c(
    monte_carlo_mse(lss_repetitions, true_betas),
    monte_carlo_mse(lsa_repetitions, true_betas)
  ),
  RMSE = sqrt(exact_mse),
  check.names = FALSE,
  row.names = NULL
)
monte_carlo_lss <- comparison$`Monte Carlo MSE`[comparison$Estimator == "LSS"]
monte_carlo_lsa <- comparison$`Monte Carlo MSE`[comparison$Estimator == "LSA"]

crossover_sd <- sqrt(
  (squared_bias["LSS"] - squared_bias["LSA"]) /
    (variance_factor["LSA"] - variance_factor["LSS"])
)
noise_grid <- c(1, crossover_sd, noise_sd)
sensitivity <- data.frame(
  `Noise SD` = noise_grid,
  `LSS RMSE` = sqrt(squared_bias["LSS"] + noise_grid^2 * variance_factor["LSS"]),
  `LSA RMSE` = sqrt(squared_bias["LSA"] + noise_grid^2 * variance_factor["LSA"]),
  check.names = FALSE
)

stopifnot(
  qr(X_residual)$rank == n_trials,
  all(is.finite(as.matrix(comparison[, -1]))),
  all(comparison$Variance > 0),
  max(abs(comparison$MSE - comparison$`Monte Carlo MSE`) / comparison$MSE) < 0.1,
  sensitivity$`LSA RMSE`[1] < sensitivity$`LSS RMSE`[1],
  tail(sensitivity$`LSS RMSE`, 1) < tail(sensitivity$`LSA RMSE`, 1)
)
```

```{r design-diagnostics, echo=FALSE}
display_diagnostics <- design_diagnostics
display_diagnostics$Value <- round(display_diagnostics$Value, 3)
knitr::kable(
  display_diagnostics,
  caption = "Diagnostics for the residualized trial design used in the comparison."
)
```

```{r comparison-table, echo=FALSE}
display_comparison <- comparison
display_comparison <- display_comparison[, c(
  "Estimator", "Squared bias", "Variance", "MSE", "RMSE"
)]
display_comparison[, -1] <- round(display_comparison[, -1], 3)
knitr::kable(
  display_comparison,
  col.names = c(
    "Estimator",
    "Squared bias (beta^2)",
    "Variance (beta^2)",
    "MSE (beta^2)",
    "RMSE (beta)"
  ),
  caption = paste0("Exact conditional decomposition at noise SD = ", noise_sd, ".")
)
```

The 100-repetition check gave MSE =
`r round(monte_carlo_lss, 3)`
for LSS and
`r round(monte_carlo_lsa, 3)`
for LSA, close to the exact values in the table.

```{r comparison-plot, echo=FALSE, fig.width=7, fig.height=3.8, fig.cap="At noise SD 2.5, LSS has more squared bias but sufficiently lower variance to produce lower mean squared error than LSA.", fig.alt="Grouped bar chart comparing LSS and LSA squared bias, variance, and mean squared error on a common beta-squared scale. LSS has higher squared bias and lower variance and mean squared error."}
plot_values <- t(as.matrix(comparison[, c("Squared bias", "Variance", "MSE")]))
colnames(plot_values) <- comparison$Estimator
op <- par(mar = c(4, 4, 1, 1), bg = "white")
barplot(
  plot_values,
  beside = TRUE,
  col = c("#9ecae1", "#3182bd", "#de2d26"),
  border = NA,
  ylab = expression("Mean squared beta units"),
  legend.text = rownames(plot_values),
  args.legend = list(x = "topleft", bty = "n", cex = 0.8)
)
par(op)
```

The noise scale matters. The calculated crossover for this fixed design and
truth is SD = `r round(crossover_sd, 3)`:

```{r sensitivity-table, echo=FALSE}
display_sensitivity <- sensitivity
display_sensitivity[] <- lapply(display_sensitivity, round, digits = 3)
knitr::kable(
  display_sensitivity,
  caption = "Conditional RMSE on both sides of the design-specific crossover."
)
```

At noise SD 1, LSA has lower RMSE; at the illustrated SD 2.5, LSS has lower
RMSE. The example therefore demonstrates both sides of the trade-off rather
than supplying a general estimator-selection rule. Event spacing, noise, HRF
mismatch, and the distribution of true trial effects can all move the
crossover. Simulate conditions that resemble your own acquisition.

## What should I change for real data?

Real fMRI errors are temporally correlated. Once the design is correct, pass a
validated prewhitening specification so the response and every design matrix
receive the same filter. This call also selects the optimized C++ backend;
prewhitening has the same model meaning across backends:

```{r prewhitening}
beta_ar1 <- lss(
  Y,
  X,
  Z = Z,
  Nuisance = Nuisance,
  method = "cpp_optimized",
  prewhiten = prewhiten_options(method = "ar", p = 1)
)
```

```{r prewhitening-check, include=FALSE}
beta_ar1_reference <- lss(
  Y,
  X,
  Z = Z,
  Nuisance = Nuisance,
  method = "naive",
  prewhiten = prewhiten_options(method = "ar", p = 1)
)
stopifnot(
  identical(dim(beta_ar1), dim(beta_lss)),
  all(is.finite(beta_ar1)),
  max(abs(beta_ar1 - beta_ar1_reference)) < 1e-8
)
```

The simulated errors above are independent, so this call demonstrates the API
and backend agreement, not a benefit from whitening. Treat AR(1) as an example
starting model: choose an order and pooling strategy using residual diagnostics
and acquisition-aware validation.

For multiple runs, use run-aware intercepts and supply run labels to
`prewhiten_options(method = "ar", p = 1, pooling = "run", runs = ...)`. Do not
concatenate runs and let filtering cross a run boundary.

For a large voxel matrix, the optimized C++ backend changes computation, not
the estimand:

```{r backend-equivalence}
beta_cpp <- lss(
  Y,
  X,
  Z = Z,
  Nuisance = Nuisance,
  method = "cpp_optimized"
)
c(maximum_absolute_difference = max(abs(beta_lss - beta_cpp)))
```

```{r backend-equivalence-check, include=FALSE}
backend_error <- max(abs(beta_lss - beta_cpp))
stopifnot(is.finite(backend_error), backend_error < 1e-8)
```

Start with the default backend while developing the analysis, then switch when
runtime or memory justifies it. Backend agreement is a useful regression check,
but agreement alone does not validate the scientific model.

## What are the important limits?

- `X` must already encode the HRF-convolved response for each trial. Use a
  design-aware workflow if you begin from event tables.
- LSS reduces competition among individual trial columns; it cannot recover
  information absent from the acquisition or rescue a singular design.
- The output contains trial coefficients, not significance tests or an
  automatic decision about whether a trial is usable.
- Nuisance projection and prewhitening change the model. Apply them
  consistently to the response and every relevant design matrix.
- LSS versus LSA is a design- and noise-dependent choice. Validate it with
  repeated simulations, not a single attractive plot.

## Where should I go next?

- Read `?lss` and `?prewhiten_options` for the complete core interface.
- Continue to `vignette("oasis_method")` for the OASIS solver, explicit ridge
  regularization, diagnostics, and multi-basis designs.
- Then read `vignette("oasis_theory")` for the derivation and computational
  contract behind that practical guide.
- Next, use `vignette("lss_with_fmridesign")` when your inputs are event tables,
  formulas, and multiple runs.
- Continue to `vignette("voxel-wise-hrf")` for estimated voxel-wise HRF shapes
  and trial coefficients.
- Finish with `vignette("sbhm")` for library-constrained voxel-specific shapes,
  score margins, and trial coefficients.

## Reference

Mumford, J. A., Turner, B. O., Ashby, F. G., & Poldrack, R. A. (2012).
Deconvolving BOLD activation in event-related designs for multivoxel pattern
classification analyses. *NeuroImage*, 59(3), 2636--2643.
<https://doi.org/10.1016/j.neuroimage.2011.08.076>
