---
title: "Getting Started with fmrilss"
author: "fmrilss Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fmrilss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## The Challenge of Event-Related fMRI Analysis

Imagine you're analyzing an fMRI experiment where participants view different images in rapid succession, with only a few seconds between each stimulus. You want to estimate the brain's response to each individual image, but there's a problem: the hemodynamic response unfolds slowly over 10-15 seconds, causing responses from consecutive trials to overlap substantially. Traditional approaches that model all trials simultaneously in a single general linear model (GLM) can suffer from collinearity, especially when trials occur close together. This collinearity makes it difficult to obtain reliable, trial-specific activation estimates that are essential for techniques like multivariate pattern analysis (MVPA) or trial-by-trial connectivity analyses.

The Least Squares Separate (LSS) approach, introduced by Mumford and colleagues in 2012, offers an elegant solution to this problem. Rather than estimating all trials at once, LSS fits a separate GLM for each trial, dramatically reducing collinearity and producing more reliable trial-specific estimates. The `fmrilss` package brings you a fast, flexible implementation of this powerful method, with modern enhancements that make it practical for large-scale analyses.

## Understanding the LSS Approach

Before diving into the implementation, let's build an intuition for how LSS works. This vignette assumes you're familiar with basic fMRI analysis concepts like the general linear model, design matrices, and the hemodynamic response function (HRF). You should also be comfortable with R programming and basic matrix operations, as we'll be working with these throughout.

In a standard GLM analysis (often called Least Squares All or LSA), we create a design matrix with separate columns for each trial and estimate all beta coefficients simultaneously. When trials are closely spaced, these columns become highly correlated, leading to unstable estimates. LSS takes a different approach: for each trial of interest, it creates a simplified model with just two main regressors. The first regressor models the trial we're currently interested in, while the second aggregates all other trials into a single regressor. This process repeats for every trial, yielding a unique, more stable beta estimate for each one.

Beyond these two core regressors, LSS models can include experimental regressors that capture session-wide effects like linear trends or block effects, which we want to model but don't need trial-specific estimates for. The model can also incorporate nuisance regressors such as motion parameters or physiological noise, which are projected out before the analysis begins.

## Your First LSS Analysis

```{r setup}
library(fmrihrf)
library(fmrilss)

set.seed(42)
n_timepoints <- 150
n_trials <- 12
n_voxels <- 25
```

Let's walk through a complete example that demonstrates the power of LSS. We'll start by creating synthetic data that mimics a typical rapid event-related design, then compare different analysis approaches to see how LSS performs.

### Creating the Experimental Design

The foundation of any fMRI analysis is the design matrix, which encodes when stimuli were presented and how we expect the brain to respond. In our example, we'll simulate an experiment with trials occurring at regular intervals, which is common in many cognitive neuroscience studies.

```{r design}
# Trial design matrix (X)
X <- matrix(0, n_timepoints, n_trials)
# Ensure integer onsets within bounds
onsets <- round(seq(from = 10, to = n_timepoints - 12, length.out = n_trials))
for(i in 1:n_trials) {
  X[onsets[i]:(onsets[i] + 5), i] <- 1
}

# Experimental regressors (Z) - intercept and condition effects
# These are regressors we want to model and get estimates for, but not trial-wise
Z <- cbind(Intercept = 1, LinearTrend = scale(1:n_timepoints, center = TRUE, scale = FALSE))

# Nuisance regressors - e.g., 6 motion parameters
Nuisance <- matrix(rnorm(n_timepoints * 6), n_timepoints, 6)
```

The trial matrix `X` represents the expected BOLD response for each trial, typically convolved with an HRF (though we're using a simplified box-car function here for clarity). The experimental regressors in `Z` capture effects that vary across the session but not trial-by-trial, such as baseline activation levels or linear drift. The nuisance matrix contains signals we want to remove, like head motion or breathing artifacts.

### Visualizing Regressors

Before fitting models, it helps to quickly inspect the experimental and nuisance regressors. Below are clean plots of the session‑wide experimental effects (excluding the intercept) and the nuisance regressors, standardized for comparability. If `ggplot2` is available, we use it for polished visuals; otherwise we fall back to base R.

```{r plot-regressors, fig.width=8, fig.height=6}
# Pretty plotting helper
plot_regressors <- function(Z, Nuisance) {
  # Omit the intercept when plotting experimental regressors
  z_names <- setdiff(colnames(Z), "Intercept")
  z_mat <- if (length(z_names)) as.matrix(Z[, z_names, drop = FALSE]) else NULL

  # Standardize (z-score) for comparability
  N_std <- scale(Nuisance)

  if (requireNamespace("ggplot2", quietly = TRUE)) {
    ggplot2::theme_set(ggplot2::theme_minimal(base_size = 12))

    # Experimental regressors (standardized)
    if (!is.null(z_mat) && ncol(z_mat) > 0) {
      z_std <- scale(z_mat)
      z_df <- data.frame(
        Time = rep(seq_len(nrow(z_std)), times = ncol(z_std)),
        Regressor = rep(colnames(z_std), each = nrow(z_std)),
        Value = as.vector(z_std)
      )
      pz <- ggplot2::ggplot(z_df, ggplot2::aes(Time, Value, color = Regressor)) +
        ggplot2::geom_line(linewidth = 0.9) +
        ggplot2::labs(
          title = "Experimental regressors (standardized; intercept omitted)",
          x = "Time (TR)", y = "Z-score"
        ) +
        ggplot2::theme(legend.position = "top", legend.title = ggplot2::element_blank())
      print(pz)
    }

    # Nuisance regressors (standardized), small multiples
    n_df <- data.frame(
      Time = rep(seq_len(nrow(N_std)), times = ncol(N_std)),
      Regressor = rep(paste0("N", seq_len(ncol(N_std))), each = nrow(N_std)),
      Value = as.vector(N_std)
    )
    pn <- ggplot2::ggplot(n_df, ggplot2::aes(Time, Value)) +
      ggplot2::geom_line(color = "#2c7fb8", linewidth = 0.6) +
      ggplot2::facet_wrap(~ Regressor, ncol = 3) +
      ggplot2::labs(
        title = "Nuisance regressors (standardized)",
        x = "Time (TR)", y = "Z-score"
      ) +
      ggplot2::theme(strip.background = ggplot2::element_rect(fill = "#f5f5f5", color = NA))
    print(pn)

  } else {
    # Base plotting fallback
    op <- par(no.readonly = TRUE)
    on.exit(par(op), add = TRUE)

    if (!is.null(z_mat) && ncol(z_mat) > 0) {
      z_std <- scale(z_mat)
      matplot(z_std, type = "l", lwd = 2, lty = 1, col = seq_len(ncol(z_std)),
              xlab = "Time (TR)", ylab = "Z-score",
              main = "Experimental regressors (standardized; intercept omitted)")
      legend("topright", legend = colnames(z_std), col = seq_len(ncol(z_std)),
             lwd = 2, bty = "n")
    }

    par(mfrow = c(3, 2), mar = c(3, 3, 2, 1))
    for (i in seq_len(ncol(N_std))) {
      plot(N_std[, i], type = "l", col = "#2c7fb8", lwd = 1.5,
           xlab = "Time (TR)", ylab = "Z-score", main = paste0("N", i))
      abline(h = 0, col = "#cccccc", lty = 3)
    }
    par(mfrow = c(1, 1))
  }
}

plot_regressors(Z, Nuisance)
```

### Generating Realistic Data

Now we'll create synthetic fMRI data that includes contributions from all these components, plus some noise to make it realistic:

```{r data}
# Simulate effects for each component
true_trial_betas <- matrix(rnorm(n_trials * n_voxels, 0, 1.2), n_trials, n_voxels)
true_fixed_effects <- matrix(rnorm(2 * n_voxels, c(5, -0.1), 0.5), 2, n_voxels)
true_nuisance_effects <- matrix(rnorm(6 * n_voxels, 0, 2), 6, n_voxels)

# Combine signals and add noise
Y <- (Z %*% true_fixed_effects) +
     (X %*% true_trial_betas) +
     (Nuisance %*% true_nuisance_effects) +
     matrix(rnorm(n_timepoints * n_voxels, 0, 1), n_timepoints, n_voxels)
```

### Running the Analysis

The `lss()` function provides a clean, modern interface that adapts to your needs. At its simplest, you can provide just the data and trial matrix, and the function automatically includes an intercept term:

```{r lss_basic}
beta_basic <- lss(Y, X)
# The result is a trials-by-voxels matrix
dim(beta_basic)
```

For a more complete analysis that accounts for experimental effects and removes nuisance signals, we include our `Z` and `Nuisance` matrices. The function handles the nuisance regression efficiently, projecting these signals out of both the data and design matrix before estimating trial-specific betas:

```{r lss_advanced}
beta_full <- lss(Y, X, Z = Z, Nuisance = Nuisance)
# The output dimensions remain the same
dim(beta_full)
```

## Choosing the Right Computational Backend

The `fmrilss` package offers multiple computational backends, each optimized for different scenarios. While the default R implementation is well-optimized and readable, making it excellent for understanding the algorithm and debugging, you'll often want to use the high-performance C++ backend for real analyses, especially with large datasets. The C++ implementation leverages Armadillo for efficient linear algebra and OpenMP for parallel processing across multiple CPU cores:

```{r lss_cpp, eval=require("Rcpp") && require("RcppArmadillo")}
# Run the same analysis with the high-performance C++ engine
beta_fast <- lss(Y, X, Z = Z, Nuisance = Nuisance, method = "cpp_optimized")

# The results are numerically identical to the R version
all.equal(beta_full, beta_fast, tolerance = 1e-8)
```

The beauty of this design is that you can seamlessly switch between backends without changing any other code, allowing you to use the readable R implementation during development and the optimized C++ version for production analyses.

## LSS versus Traditional GLM: When Each Shines

To understand when LSS offers advantages over traditional approaches, let's compare it with the standard Least Squares All (LSA) method that estimates all trials simultaneously:

```{r lsa-comparison}
# LSA: Standard GLM with all trials in one model
beta_lsa <- lsa(Y, X, Z = Z, Nuisance = Nuisance)

# Compare dimensions
cat("LSS beta dimensions:", dim(beta_full), "\n")
cat("LSA beta dimensions:", dim(beta_lsa), "\n")

# Compare variance in beta estimates
var_lss <- apply(beta_full, 2, var)
var_lsa <- apply(beta_lsa, 2, var)

cat("\nMean variance across voxels:\n")
cat("  LSS:", mean(var_lss), "\n")
cat("  LSA:", mean(var_lsa), "\n")

# Plot comparison
par(mfrow = c(1, 2))
hist(beta_full[, 1], main = "LSS: Beta distribution (Voxel 1)",
     xlab = "Beta values", col = "lightblue")
hist(beta_lsa[, 1], main = "LSA: Beta distribution (Voxel 1)",
     xlab = "Beta values", col = "lightgreen")
```

The comparison reveals important differences between the methods. LSS typically produces beta estimates with different variance characteristics than LSA, which can be advantageous for certain analyses. When should you choose one over the other?

For MVPA analyses, LSS generally performs better because it reduces collinearity between trial estimates, leading to more distinguishable patterns. In rapid event-related designs where hemodynamic responses overlap substantially, LSS's approach of isolating each trial helps maintain estimate stability. However, for block designs with well-separated trials, or when you need to compute group-level contrasts that benefit from the stability of simultaneous estimation, LSA might be more appropriate. Connectivity analyses that require trial-specific estimates particularly benefit from LSS's approach.

## Introducing the OASIS Method

While the standard LSS implementation is powerful, the package also includes the OASIS (Optimized Analytic Single-pass Inverse Solution) method, which extends LSS with several sophisticated features. OASIS automatically estimates hemodynamic response functions, applies ridge regularization for enhanced stability, provides efficient computation for complex designs, and includes built-in support for multi-basis HRF models. Here's a taste of what OASIS can do:

```{r oasis-intro}
# Basic OASIS usage
# fmrihrf is now imported automatically
beta_oasis <- lss(
  Y = Y,
  X = NULL,  # OASIS builds design internally
  method = "oasis",
  oasis = list(
    design_spec = list(
      sframe = sampling_frame(blocklens = nrow(Y), TR = 1),
      cond = list(
        onsets = onsets,        # reuse onsets from above
        hrf = HRF_SPMG1,        # HRF model
        span = 30
      )
    ),
    ridge_mode = "fractional",
    ridge_x = 0.01,
    ridge_b = 0.01
  )
)

cat("OASIS beta dimensions:", dim(beta_oasis), "\n")
```

## Working with Different Computational Backends

The package provides multiple backends to match your computational needs and resources. Each backend implements the same algorithm but with different optimization strategies. The naive implementation offers the clearest code for understanding the algorithm, while the vectorized and optimized R versions provide good performance with pure R code. For production work with large datasets, the C++ backend offers the best performance, especially when combined with parallel processing.

```{r backend-comparison, eval=FALSE}
# Benchmark different methods
library(microbenchmark)

methods <- c("naive", "r_vectorized", "r_optimized", "cpp_optimized")
timings <- list()

for (m in methods) {
  timings[[m]] <- system.time({
    lss(Y, X, method = m)
  })[3]
}

# Display timing comparison
timing_df <- data.frame(
  Method = methods,
  Time = unlist(timings)
)
print(timing_df)

# For large datasets, consider threading for C++ backends
if (require("parallel")) {
  n_cores <- parallel::detectCores() - 1
  # Set OpenMP threads for C++ backend if supported
  Sys.setenv(OMP_NUM_THREADS = n_cores)
}
```

## Handling Complex Experimental Designs

Real experiments often involve multiple conditions, parametric modulations, and various covariates. The `fmrilss` package handles these complexities gracefully. When working with multiple conditions, you can create separate design matrices for each condition and include condition labels as experimental regressors. This allows you to model condition-specific effects while still obtaining trial-wise estimates within each condition.

```{r multiple-conditions}
# Create design with multiple conditions
n_cond <- 3
X_multi <- matrix(0, n_timepoints, n_trials * n_cond)

# Generate a simple HRF for demonstration
hrf <- c(0, 0.2, 0.5, 0.8, 1, 0.9, 0.7, 0.5, 0.3, 0.1)

for (c in 1:n_cond) {
  trial_idx <- ((c-1) * n_trials + 1):(c * n_trials)
  for (i in 1:n_trials) {
    onset <- 10 + (trial_idx[i] - 1) * 5
    if (onset + 9 <= n_timepoints) {
      X_multi[onset:(onset + 9), trial_idx[i]] <- hrf
    }
  }
}

# Add condition labels as experimental regressors
Z_cond <- matrix(0, n_timepoints, n_cond)
for (c in 1:n_cond) {
  trial_idx <- ((c-1) * n_trials + 1):(c * n_trials)
  Z_cond[, c] <- rowSums(X_multi[, trial_idx, drop = FALSE])
}

# Run LSS with condition regressors
beta_multi <- lss(Y, X_multi, Z = Z_cond, method = "r_optimized")
```

Parametric modulations, where trial responses are weighted by continuous variables like reaction time or stimulus intensity, are also straightforward to implement. For a quantitative demo, we simulate a parametric effect in the data and show recovery with and without the modulator:

```{r parametric-modulation}
# Add parametric modulator (e.g., reaction time, stimulus intensity)
modulator <- scale(rnorm(n_trials, mean = 0, sd = 1), center = TRUE, scale = FALSE)

# Create parametrically modulated design (scale each trial by its modulator)
X_param <- sweep(X, 2, as.numeric(modulator), `*`)

# Simulate data with a parametric effect (reuse fixed and nuisance parts)
Y_mod <- (Z %*% true_fixed_effects) +
         (X_param %*% true_trial_betas) +
         (Nuisance %*% true_nuisance_effects) +
         matrix(rnorm(n_timepoints * n_voxels, 0, 1), n_timepoints, n_voxels)

# Fit models with and without the parametric modulator
beta_unmod <- lss(Y_mod, X,       Z = Z, method = "r_optimized")
beta_param <- lss(Y_mod, X_param, Z = Z, method = "r_optimized")

# Ground truths for comparison
# - For unmodulated design X, the true coefficients are modulator * true_trial_betas
true_unmod_coefs <- sweep(true_trial_betas, 1, as.numeric(modulator), `*`)
# - For parametrically modulated design X_param, the true coefficients are true_trial_betas

cor_unmod <- cor(as.vector(beta_unmod), as.vector(true_unmod_coefs))
cor_param <- cor(as.vector(beta_param), as.vector(true_trial_betas))

cat("Correlation with true coefficients (parametric effect simulated):\n")
cat("  Using X (no modulator):\t", round(cor_unmod, 3), "\n")
cat("  Using X_param (with modulator):", round(cor_param, 3), "\n")
```

## Optimizing Performance for Large-Scale Analyses

When working with whole-brain data containing hundreds of thousands of voxels, memory management and computational efficiency become critical. The package provides several strategies for handling large datasets effectively.

For very large datasets that exceed available memory, you can process voxels in chunks. This approach maintains reasonable memory usage while still benefiting from vectorized operations within each chunk:

```{r memory-tips, eval=FALSE}
# For very large datasets, process in chunks
chunk_size <- 1000
n_chunks <- ceiling(ncol(Y) / chunk_size)

beta_chunks <- list()
for (chunk in 1:n_chunks) {
  voxel_idx <- ((chunk - 1) * chunk_size + 1):min(chunk * chunk_size, ncol(Y))
  beta_chunks[[chunk]] <- lss(Y[, voxel_idx], X, method = "cpp_optimized")
}

# Combine results
beta_full <- do.call(cbind, beta_chunks)
```

Another optimization strategy involves preprocessing your data to remove nuisance signals before running LSS. While LSS can handle nuisance regressors internally, preprocessing can be more efficient when running multiple analyses:

```{r preprocessing}
# Project out nuisance before LSS (when appropriate)
# project_confounds returns the projection matrix Q; apply it to both Y and X
Q_nuis <- project_confounds(Nuisance)
Y_clean <- Q_nuis %*% Y
X_clean <- Q_nuis %*% X

# This can be faster than including Nuisance in each LSS iteration
beta_preprocessed <- lss(Y_clean, X_clean, Z = Z, method = "r_optimized")
```

Choosing the right backend for your data size is crucial for optimal performance. As a rule of thumb, for small datasets with fewer than 100 voxels, the optimized R implementation is usually fast enough and offers the advantage of being easily debuggable. For medium-sized datasets up to 10,000 voxels, the C++ backend provides significant speedup. For very large whole-brain datasets, consider using OASIS, which implements additional optimizations for handling massive design matrices efficiently.

## Troubleshooting Common Challenges

Even with robust implementations, certain data characteristics can cause issues. Understanding how to diagnose and address these problems will help you get the most out of your analyses.

When design matrices become singular or near-singular due to high collinearity between regressors, standard least squares solutions become unstable. You can detect this by examining the correlation matrix of your design:

```{r troubleshoot-singular, eval=FALSE}
# Check for collinearity
cor_matrix <- cor(X)
high_cor <- which(abs(cor_matrix) > 0.9 & cor_matrix != 1, arr.ind = TRUE)

if (nrow(high_cor) > 0) {
  warning("High correlation between regressors detected")
  # Consider using ridge regression via OASIS
  beta_ridge <- lss(Y, X, method = "oasis",
                    oasis = list(ridge_mode = "absolute", ridge_x = 0.1))
}
```

Memory limitations can also pose challenges with large datasets. Before starting an analysis, it's wise to estimate memory requirements and adjust your approach accordingly:

```{r troubleshoot-memory, eval=FALSE}
# Monitor memory usage
mem_required <- object.size(Y) * n_trials * 2  # Rough estimate
# mem_available <- memory.limit()  # Windows only

# if (mem_required > mem_available * 0.8) {
#   warning("May run out of memory. Consider chunking or using OASIS.")
# }
```

If you encounter unexpectedly slow performance, profiling can help identify bottlenecks:

```{r troubleshoot-speed, eval=FALSE}
# Profile code to find bottlenecks
Rprof("lss_profile.out")
beta_slow <- lss(Y, X, method = "naive")
Rprof(NULL)
summaryRprof("lss_profile.out")
```

## Where to Go From Here

This vignette has introduced you to the core concepts and capabilities of the `fmrilss` package. You've learned how LSS addresses the collinearity problem in rapid event-related designs, how to choose between different computational backends, and strategies for handling complex experimental designs and large datasets.

To deepen your understanding and explore advanced features, we recommend examining the voxel-wise HRF vignette, which demonstrates how to model spatial variation in hemodynamic responses across the brain. The OASIS method vignette provides comprehensive coverage of this powerful extension, including HRF estimation and ridge regression techniques. For those interested in hierarchical analyses, the mixed models vignette shows how to combine LSS with mixed-effects modeling frameworks.

The `fmrilss` package represents a comprehensive toolkit for trial-wise beta estimation, providing options that range from simple, interpretable implementations to highly optimized solutions for large-scale analyses. Whether you're conducting exploratory analyses on a laptop or processing massive datasets on a computing cluster, the package offers the flexibility and performance you need for modern fMRI analysis.

## References

Mumford, J. A., Turner, B. O., Ashby, F. G., & Poldrack, R. A. (2012). Deconvolving BOLD activation in event-related designs for multivoxel pattern classification analyses. *NeuroImage*, 59(3), 2636-2643.
