---
title: "Getting Started with fmrilss"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fmrilss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction to Least Squares Separate (LSS)

The `fmrilss` package provides a fast and flexible implementation of the Least Squares Separate (LSS) modeling approach, first proposed by Mumford et al. (2012). LSS is a powerful method for analyzing event-related fMRI data, especially for studies involving multivariate pattern analysis (MVPA) or functional connectivity.

The core idea of LSS is to estimate a separate GLM for each trial. For a given trial, the model includes:

1.  A regressor for the **trial of interest**.
2.  A single regressor for **all other trials** combined.
3.  **Experimental regressors** that we want to model and estimate, but not trial-wise (e.g., intercept, condition effects, block effects).
4.  **Nuisance regressors** that we want to remove from the data before analysis (e.g., motion parameters, physiological noise).

This process is repeated for every trial, yielding a unique beta estimate for each one.

## The `lss()` Function

The main entry point to the package is the `lss()` function, which has a clean and modern interface:

```r
lss(Y, X, Z = NULL, Nuisance = NULL, method = "r_optimized")
```

-   `Y`: The data matrix (timepoints x voxels).
-   `X`: The trial design matrix (timepoints x trials).
-   `Z`: A matrix of experimental regressors we want to model and get beta estimates for, but not trial-wise (e.g., intercept, condition effects, block effects). Defaults to an intercept if `NULL`.
-   `Nuisance`: A matrix of nuisance regressors to be projected out before analysis (e.g., motion parameters, physiological noise).
-   `method`: The computational backend to use. Options range from a simple `naive` R loop to a highly optimized and parallelized `"cpp_optimized"` engine.

## A Practical Example

Let's walk through a complete example, from generating data to running the analysis.

### 1. Load the Package and Prepare Data

First, we load `fmrilss` and set up our parameters.

```{r setup}
library(fmrilss)

set.seed(42)
n_timepoints <- 150
n_trials <- 12
n_voxels <- 25
```

### 2. Create Design Matrices

Next, we create the core components of our model.

-   **Trial Matrix (`X`)**: A matrix where each column represents the predicted BOLD response for a single trial.
-   **Experimental Regressors (`Z`)**: A matrix for experimental regressors that we want to model and get beta estimates for, but not trial-wise. These might include condition effects, block effects, or other experimental factors that vary across the session but not trial-by-trial.
-   **Nuisance Matrix**: Regressors we want to remove from the data before modeling, such as motion parameters or physiological noise.

```{r design}
# Trial design matrix (X)
X <- matrix(0, n_timepoints, n_trials)
onsets <- seq(from = 10, to = n_timepoints - 12, length.out = n_trials)
for(i in 1:n_trials) {
  X[onsets[i]:(onsets[i] + 5), i] <- 1
}

# Experimental regressors (Z) - intercept and condition effects
# These are regressors we want to model and get estimates for, but not trial-wise
Z <- cbind(Intercept = 1, LinearTrend = scale(1:n_timepoints, center = TRUE, scale = FALSE))

# Nuisance regressors - e.g., 6 motion parameters
Nuisance <- matrix(rnorm(n_timepoints * 6), n_timepoints, 6)
```

### 3. Simulate Data (`Y`)

Now, we'll generate a synthetic data matrix `Y` that includes signals from our regressors plus some random noise.

```{r data}
# Simulate effects for each component
true_trial_betas <- matrix(rnorm(n_trials * n_voxels, 0, 1.2), n_trials, n_voxels)
true_fixed_effects <- matrix(rnorm(2 * n_voxels, c(5, -0.1), 0.5), 2, n_voxels)
true_nuisance_effects <- matrix(rnorm(6 * n_voxels, 0, 2), 6, n_voxels)

# Combine signals and add noise
Y <- (Z %*% true_fixed_effects) + 
     (X %*% true_trial_betas) +
     (Nuisance %*% true_nuisance_effects) +
     matrix(rnorm(n_timepoints * n_voxels, 0, 1), n_timepoints, n_voxels)
```

### 4. Run the LSS Analysis

With our data prepared, we can now run the `lss` analysis.

#### Basic Usage
If you only provide the data `Y` and the trial matrix `X`, the function will automatically include an intercept.

```{r lss_basic}
beta_basic <- lss(Y, X)
# The result is a trials-by-voxels matrix
dim(beta_basic)
```

#### Including Experimental Regressors and Nuisance Regressors
For a more realistic analysis, we include our `Z` and `Nuisance` matrices. The nuisance regressors are projected out of both `Y` and `X` before the trial-wise GLMs are estimated, while the experimental regressors in `Z` are included in every trial-wise GLM to get their beta estimates.

```{r lss_advanced}
beta_full <- lss(Y, X, Z = Z, Nuisance = Nuisance)
# The output dimensions remain the same
dim(beta_full)
```

### 5. Choosing a High-Performance Method

While the default R implementation is well-optimized, the C++ backend offers a significant speedup, especially for large datasets. It is also parallelized with OpenMP to use multiple CPU cores. To use it, simply set `method = "cpp_optimized"`.

```{r lss_cpp, eval=require("Rcpp") && require("RcppArmadillo")}
# Run the same analysis with the high-performance C++ engine
beta_fast <- lss(Y, X, Z = Z, Nuisance = Nuisance, method = "cpp_optimized")

# The results are numerically identical to the R version
all.equal(beta_full, beta_fast, tolerance = 1e-8)
```

This makes it easy to switch between a readable R implementation and a high-performance C++ engine without changing any other code.

## References

Mumford, J. A., Turner, B. O., Ashby, F. G., & Poldrack, R. A. (2012). Deconvolving BOLD activation in event-related designs for multivoxel pattern classification analyses. *NeuroImage*, 59(3), 2636-2643. 