---
title: 'GLMsingle in fmrilss: library HRFs, GLMdenoise and fractional ridge'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{GLMsingle in fmrilss: library HRFs, GLMdenoise and fractional ridge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(family = params$family, preset = params$preset)
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  dpi = 150
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## What GLMsingle estimates

GLMsingle (Prince et al., 2022) estimates one response amplitude per trial
and voxel. It adds three data-driven steps to a single-trial GLM:

1. **A library HRF per voxel.** Each voxel uses whichever of 20 canonical HRF
   shapes best explains its time series (type B).
2. **GLMdenoise.** Principal components of the residual time series in a
   "noise pool" of task-unresponsive voxels become nuisance regressors. The
   number of components is chosen by cross-validation (type C).
3. **Fractional ridge regression.** Each voxel gets its own amount of
   shrinkage, chosen by cross-validation, and the result is rescaled to the
   unregularised betas (type D).

Both cross-validations compare single-trial betas for the **same condition
in different runs**, so the design must repeat conditions across runs.

`glmsingle()` computes the same estimator as the reference GLMsingle
implementation, but organises the work so that each run is solved
separately, data are touched once per stage, and every candidate model is
scored from small summary matrices.

## A simulated experiment

Four runs of 110 TRs. Each run shows trials from 8 conditions; every
condition repeats across runs. Voxels differ in HRF shape, and all voxels
share slow structured noise, which GLMdenoise can learn from voxels that do
not respond to the task.

```{r simulate}
library(fmrilss)
set.seed(2024)
tr <- 1; stimdur <- 3
n_runs <- 4; n_time <- 110; n_vox <- 120; n_cond <- 8
lib <- glmsingle_hrf_library(stimdur, tr)
true_hrf <- sample(ncol(lib), n_vox, replace = TRUE)
responsive <- seq_len(n_vox) <= 80
cond_mean <- matrix(rnorm(n_cond * n_vox, 1.5, 1), n_cond)
load <- matrix(rnorm(3 * n_vox, 0, 1.5), 3)

design <- Y <- true_beta <- vector("list", n_runs)
for (r in seq_len(n_runs)) {
  onsets <- seq(3, n_time - 15, by = 4) + sample(0:1, 1)
  cond <- sample(rep_len(seq_len(n_cond), length(onsets)))
  D <- matrix(0, n_time, n_cond)
  D[cbind(onsets + 1, cond)] <- 1
  amp <- cond_mean[cond, ] + matrix(rnorm(length(onsets) * n_vox, 0, 0.5), length(onsets))
  amp[, !responsive] <- 0
  signal <- vapply(seq_len(n_vox), function(v) {
    s <- numeric(n_time); s[onsets + 1] <- amp[, v]
    stats::convolve(s, rev(lib[, true_hrf[v]]), type = "open")[seq_len(n_time)]
  }, numeric(n_time))
  latent <- apply(matrix(rnorm(3 * n_time), n_time), 2, cumsum) * 0.4
  noise <- matrix(rnorm(n_time * n_vox), n_time) + latent %*% load
  Y[[r]] <- 1000 * (1 + 0.01 * (signal + noise))
  design[[r]] <- D
  true_beta[[r]] <- amp
}
true_beta <- do.call(rbind, true_beta)  # trials x voxels, chronological
```

`design` is in GLMsingle's format: one time x condition 0/1 matrix per run,
with a 1 at each trial onset. A data frame with columns `run`, `onset` (in
seconds) and `condition` works too, as does an fmridesign event model via
`glmsingle_design()`.

## Fitting

```{r fit}
fit <- glmsingle(Y, design, tr = tr, stimdur = stimdur, verbose = FALSE)
fit
```

Each model type is a list with GLMsingle's field names. Betas are trials x
voxels in chronological trial order and, by default, in percent signal
change.

```{r inspect}
dim(coef(fit))            # type D
fit$typed$pcnum           # number of GLMdenoise components
table(fit$typed$FRACvalue)
```

## Did the data-driven steps help?

The truth is known here, so each model type can be scored by how well its
betas correlate with the true trial amplitudes in responsive voxels. Betas
are in percent signal change and the simulation scales signal by 1% of the
baseline, so the scale matches.

```{r accuracy}
score <- function(B) {
  mean(vapply(which(responsive), function(v) stats::cor(B[, v], true_beta[, v]), numeric(1)))
}
round(c(B = score(coef(fit, "b")), C = score(coef(fit, "c")), D = score(coef(fit, "d"))), 3)
```

HRF selection can also be checked directly. Neighbouring library HRFs are
very similar, so a near miss is still a good shape:

```{r hrf-recovery}
mean(abs(fit$typeb$HRFindex[responsive] - true_hrf[responsive]) <= 2)
```

## Choices that differ from the reference implementation

The defaults reproduce GLMsingle (Python, commit `1ab54a6`) except where
GLMsingle is internally inconsistent. Each difference is an argument whose
other value reproduces GLMsingle:

| Argument | Default | GLMsingle behaviour |
|---|---|---|
| `extras_in_denoise` | `"always"`: user nuisance regressors (e.g. motion) are in every GLMdenoise and ridge fit | `"with_pcs"`: dropped when zero PCs are used, including from the cross-validation reference |
| `zero_sd_cv` | `"zero"`: voxels whose reference betas have zero variance carry no cross-validation weight | `"python"`: treated inconsistently across candidates by an in-place division helper |
| `singular` | `"error"`, as GLMsingle | `"pinv"` gives a minimum-norm solution |
| `frac_alpha` | `"fracridge"`: GLMsingle's grid interpolation of the ridge penalty | `"exact"` solves the fraction equation exactly (experimental) |

The automatic ON-OFF R^2 threshold uses a deterministic Gaussian-mixture
fit with a small variance floor (as in GLMsingle's MATLAB code). Pass
`brain_r2` and `pc_r2_cutoff` to fix the thresholds yourself.

`full_glmbadness = TRUE` fills the PC cross-validation diagnostic for every
voxel. By default only the voxels that decide the number of PCs are
evaluated, which gives identical estimates.

## How agreement with GLMsingle is checked

The package tests compare `glmsingle()` with pinned Python GLMsingle on 11
simulated scenarios: default settings, extra regressors, two sessions with
grouped folds, unequal run lengths, unrepeated conditions, rapid events,
near-collinear regressors, an all-zero voxel, a single ridge fraction, and no
HRF library. Every HRF choice, number of PCs and ridge fraction matches.
Betas and R^2 agree to about 1e-6, the precision of GLMsingle's single-precision
arithmetic. For rapid designs the agreement is limited by GLMsingle's float32
normal equations (relative error grows with the square of the design's
condition number). Separate tests check every stage in double precision
against literal re-implementations of GLMsingle's stacked computations.

## Reference

Prince, J. S., Charest, I., Kurzawski, J. W., Pyles, J. A., Tarr, M. J., &
Kay, K. N. (2022). Improving the accuracy of single-trial fMRI response
estimates using GLMsingle. *eLife*, 11, e77599.
