---
title: 'GLMsingle in fmrilss: library HRFs, GLMdenoise and fractional ridge'
output:
  albersdown::albers_vignette:
    family: red
    preset: homage
    fig_width: 8
    fig_height: 4.5
    toc: yes
    toc_depth: 2.0

vignette: |
  %\VignetteIndexEntry{GLMsingle in fmrilss: library HRFs, GLMdenoise and fractional ridge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include=FALSE}
knitr::opts_chunk$set(
  fig.retina = 1,
  collapse = TRUE,
  comment = "#>",
  dpi = 150
)
library(ggplot2)
plot_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
accent <- "#AF3B2E"
```

## What GLMsingle estimates

GLMsingle (Prince et al., 2022) estimates one response amplitude per trial
and voxel. Use `glmsingle()` when you want to select an HRF from a library,
estimate shared noise regressors, and choose ridge shrinkage from repeated
conditions across runs. The procedure builds on a single-trial general linear
model (GLM):

1. **Select an HRF per voxel (type B).** Choose the best-fitting shape from
   a library of 20 haemodynamic response functions (HRFs).
2. **Add GLMdenoise regressors (type C).** Identify a pool of voxels with
   weak task fit, remove polynomial trends from their time series, and use
   principal components (PCs) of the normalized residuals as nuisance
   regressors. Cross-validation selects the number of PCs.
3. **Apply fractional ridge regression (type D).** Cross-validation selects
   shrinkage separately for each voxel. By default, the ridge estimates are
   then rescaled and offset to best match the unregularised estimates.

An initial ON-OFF model (type A) uses one regressor for all trials and helps
define the noise pool. Both default cross-validations compare trial betas for
**the same condition in different runs**. Without those repeats, their
data-driven choices are unavailable; the function warns and disables the
stages that require them.

The implementation solves each run separately and reuses matrix products to
score candidate models. The final sections describe the differences from the
reference implementation and how numerical agreement is tested.

## A simulated experiment

The simulation has four runs of 110 time points, sampled every second. Each
run contains trials from eight conditions, with every condition repeated
across runs. Voxels differ in HRF shape and share slow structured noise.
Eighty voxels respond to the task; the remaining forty provide a potential
noise pool for GLMdenoise.

```{r simulate}
library(fmrilss)
set.seed(2024)
tr <- 1; stim_dur <- 3
n_runs <- 4; n_time <- 110; n_vox <- 120; n_cond <- 8
lib <- glmsingle_hrf_library(stim_dur, tr)
true_hrf <- sample(ncol(lib), n_vox, replace = TRUE)
responsive <- seq_len(n_vox) <= 80
cond_mean <- matrix(rnorm(n_cond * n_vox, 1.5, 1), n_cond)
load <- matrix(rnorm(3 * n_vox, 0, 1.5), 3)

design <- Y <- true_beta <- vector("list", n_runs)
for (r in seq_len(n_runs)) {
  onsets <- seq(3, n_time - 15, by = 4) + sample(0:1, 1)
  cond <- sample(rep_len(seq_len(n_cond), length(onsets)))
  D <- matrix(0, n_time, n_cond)
  D[cbind(onsets + 1, cond)] <- 1
  amp <- cond_mean[cond, ] + matrix(rnorm(length(onsets) * n_vox, 0, 0.5), length(onsets))
  amp[, !responsive] <- 0
  signal <- vapply(seq_len(n_vox), function(v) {
    s <- numeric(n_time); s[onsets + 1] <- amp[, v]
    stats::convolve(s, rev(lib[, true_hrf[v]]), type = "open")[seq_len(n_time)]
  }, numeric(n_time))
  latent <- apply(matrix(rnorm(3 * n_time), n_time), 2, cumsum) * 0.4
  noise <- matrix(rnorm(n_time * n_vox), n_time) + latent %*% load
  Y[[r]] <- 1000 * (1 + 0.01 * (signal + noise))
  design[[r]] <- D
  true_beta[[r]] <- amp
}
true_beta <- do.call(rbind, true_beta)  # trials x voxels, chronological
```

`design` contains one time-by-condition 0/1 matrix per run, with a 1 at each
trial onset. Unlike the trial matrices passed to `lss()`, these are onset
indicators, not HRF-convolved regressors.

You can also supply a data frame with `run`, `onset` (seconds, on the TR grid),
and `condition` columns, or use `glmsingle_design()` with an fmridesign event
model. For list-valued `Y`, ascending event run IDs correspond to the list
order. For a single data matrix, supply scan-level `runs`; event IDs are
matched to the runs in data order.

## Fitting

```{r fit}
fit <- glmsingle(Y, design, tr = tr, stim_dur = stim_dur, verbose = FALSE)
fit
```

The result contains `typea`, `typeb`, `typec`, and `typed` lists using
GLMsingle's field names. Types B--D contain trial-by-voxel betas, ordered by
run and then by onset within each run. Betas are in percent signal change by
default. `coef(fit)` extracts type D; pass `"b"` or `"c"` to extract an earlier
stage.

```{r inspect}
dim(coef(fit))            # type D
fit$typed$pcnum           # number of GLMdenoise components
```

A ridge fraction of 1 means no shrinkage; smaller fractions request more
shrinkage before the final rescaling. `pcnum` is the selected number of noise
components, shared across voxels; `FRACvalue` contains one fraction per voxel.

### What did cross-validation choose?

The left panel shows `xvaltrend`: the negative median cross-validation loss
over the voxels used to choose the PC count. Higher is better; it measures
agreement between betas for repeated conditions across runs, not time-series
R^2. With the default `pc_stop = 1.05`, the algorithm takes the first PC count
whose improvement over zero PCs is at least `1 / 1.05` (about 95%) of the best
improvement. The dashed line marks that threshold, and the red point marks
the selected count.

The right panels show the selected ridge fractions separately for the two
**simulated** voxel groups. These labels come from the generating model, not
from GLMsingle's estimated noise pool. Counts have a common scale; the groups
contain 80 and 40 voxels, respectively.

```{r tuning-data}
pc_scores <- data.frame(
  PCs = seq_along(fit$typec$xvaltrend) - 1L,
  score = fit$typec$xvaltrend
)
pc_threshold <- pc_scores$score[1] +
  (max(pc_scores$score) - pc_scores$score[1]) / fit$settings$pc_stop
ridge_choices <- data.frame(
  fraction = fit$typed$FRACvalue,
  group = ifelse(responsive, "Task-responsive\n(n = 80)", "No task response\n(n = 40)")
)
```

```{r tuning-plot, echo=FALSE, fig.width=9, fig.height=4.5, fig.cap="Cross-validation diagnostics from the same fit. Ridge fractions describe shrinkage before the default final rescaling and offset; they are not the final ratio of estimated to true amplitude.", fig.alt=sprintf("Left: cross-validation score versus noise PC count, with %d PCs selected and a dashed near-optimal threshold. Right: ridge-fraction counts split by simulated task responsiveness; smaller fractions mean more shrinkage before rescaling.", fit$typec$pcnum)}
pc_plot <- ggplot(pc_scores, aes(PCs, score)) +
  geom_hline(yintercept = pc_threshold, linetype = "dashed", colour = "grey55") +
  geom_line(colour = "grey35", linewidth = 0.7) +
  geom_point(colour = "grey35", size = 1.5) +
  geom_point(data = pc_scores[pc_scores$PCs == fit$typec$pcnum, ],
             colour = accent, size = 3.5) +
  scale_x_continuous(breaks = pretty(pc_scores$PCs)) +
  labs(title = "Noise PCs",
       subtitle = sprintf("Selected: %d", fit$typec$pcnum),
       x = "Noise PC count", y = "CV score") +
  plot_theme
ridge_plot <- ggplot(ridge_choices, aes(fraction)) +
  geom_bar(width = 0.04, fill = accent) +
  facet_wrap(~ group, ncol = 1) +
  scale_x_continuous(breaks = c(0.05, 0.5, 1), limits = c(0, 1.025)) +
  labs(title = "Ridge", subtitle = "1 = no shrinkage",
       x = "Ridge fraction", y = "Voxel count") +
  plot_theme
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2)))
print(pc_plot, newpage = FALSE, vp = grid::viewport(layout.pos.col = 1))
print(ridge_plot, newpage = FALSE, vp = grid::viewport(layout.pos.col = 2))
grid::popViewport()
```

## Did the data-driven steps help?

Because the generating amplitudes are known, we can compare estimated and
true betas in the responsive voxels. For each voxel, we correlate estimated
and true amplitudes across all trials. This measures agreement in
trial-to-trial variation, but does not measure amplitude bias or absolute
estimation error. Correlation is undefined for the forty nonresponsive
voxels, whose true amplitudes are all zero, so they are excluded here.

```{r accuracy}
voxel_cor <- sapply(c("b", "c", "d"), function(stage) {
  B <- coef(fit, stage)
  vapply(which(responsive), function(v) {
    stats::cor(B[, v], true_beta[, v])
  }, numeric(1))
})
round(colMeans(voxel_cor), 3)
```

```{r accuracy-plot, echo=FALSE, fig.width=8, fig.height=4.8, fig.cap="Each thin line follows one of the 80 responsive voxels; red points and labels show the mean at each stage. The spread and downward segments remain visible alongside the average improvement.", fig.alt=sprintf("Per-voxel correlations with true trial amplitudes across stages B, C, and D. Mean correlations are %.3f, %.3f, and %.3f, respectively; individual trajectories show heterogeneous recovery.", mean(voxel_cor[, 1]), mean(voxel_cor[, 2]), mean(voxel_cor[, 3]))}
stopifnot(all(is.finite(voxel_cor)), all(abs(voxel_cor) <= 1))
accuracy <- data.frame(
  voxel = rep(which(responsive), times = 3),
  stage = factor(rep(c("B", "C", "D"), each = sum(responsive)),
                 levels = c("B", "C", "D")),
  correlation = as.vector(voxel_cor)
)
stage_means <- aggregate(correlation ~ stage, accuracy, mean)
ggplot(accuracy, aes(stage, correlation)) +
  geom_hline(yintercept = 0, colour = "grey75", linetype = "dotted") +
  geom_line(aes(group = voxel), colour = "grey45", alpha = 0.25, linewidth = 0.4) +
  geom_point(colour = "grey45", alpha = 0.3, size = 1) +
  geom_line(data = stage_means, aes(group = 1), colour = accent, linewidth = 1) +
  geom_point(data = stage_means, colour = accent, size = 3) +
  geom_label(data = stage_means, aes(label = sprintf("%.3f", correlation)),
             nudge_y = 0.13, colour = accent, linewidth = 0, size = 3.8) +
  scale_x_discrete(labels = c(B = "B\nHRF selection", C = "C\n+ GLMdenoise",
                              D = "D\n+ ridge")) +
  scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
  labs(title = "Trial-amplitude recovery",
       x = NULL, y = "Correlation with true amplitudes") +
  plot_theme
```

In this simulation, mean recovery improves from B to C and again from C to D.
The voxel-level lines show why the mean alone is incomplete:
`r sum(voxel_cor[, "c"] < voxel_cor[, "b"])` voxels have lower correlations
after denoising, and `r sum(voxel_cor[, "d"] < voxel_cor[, "c"])` after ridge.
A useful step on average need not help every voxel. This is one fixed
simulated experiment, not a guarantee of improvement for another acquisition.

### Which HRF shapes were recovered?

The library contains responses to a `r stim_dur`-second stimulus, sampled every
`r tr` second and normalized to a peak of one. Its 20 shapes differ in their
timing, width, and undershoot. The first panel shows the full library; the
remaining panels compare the generating and selected shapes for voxels 1, 2,
and 3, chosen by index before inspecting recovery.

```{r hrf-recovery}
example_voxels <- 1:3
data.frame(voxel = example_voxels,
           true_index = true_hrf[example_voxels],
           selected_index = fit$typeb$HRFindex[example_voxels])
```

```{r hrf-plot, echo=FALSE, fig.width=8, fig.height=6, fig.cap="All curves use the same time and peak-normalized amplitude scales. Solid blue curves are the generating HRFs; dashed red curves are the selected library shapes. Coincident curves indicate an exact library match.", fig.alt="Four panels show all 20 HRFs and true versus selected shapes for the first three responsive voxels. Panel labels give both library indices; shape differences can be read directly rather than inferred from index distance."}
hrf_time <- (seq_len(nrow(lib)) - 1) * tr
hrf_curves <- data.frame(
  time = rep(hrf_time, times = ncol(lib)),
  amplitude = as.vector(lib),
  curve = rep(seq_len(ncol(lib)), each = nrow(lib)),
  source = "Library", panel = "All 20 library HRFs"
)
for (v in example_voxels) {
  indices <- c(true_hrf[v], fit$typeb$HRFindex[v])
  hrf_curves <- rbind(hrf_curves, data.frame(
    time = rep(hrf_time, 2), amplitude = as.vector(lib[, indices]),
    curve = c(rep(1, length(hrf_time)), rep(2, length(hrf_time))),
    source = rep(c("True", "Selected"), each = length(hrf_time)),
    panel = sprintf("Voxel %d\ntrue %d / selected %d", v, indices[1], indices[2])
  ))
}
hrf_curves$panel <- factor(hrf_curves$panel, levels = unique(hrf_curves$panel))
ggplot(hrf_curves, aes(time, amplitude, group = curve,
                      colour = source, linetype = source)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ panel, ncol = 2) +
  scale_x_continuous(breaks = c(0, 20, 40)) +
  scale_colour_manual(values = c(Library = "grey65", True = "#246586", Selected = accent),
                      breaks = c("True", "Selected")) +
  scale_linetype_manual(values = c(Library = "solid", True = "solid", Selected = "dashed"),
                        breaks = c("True", "Selected")) +
  labs(x = "Time since onset (s)", y = "Peak-normalized response") +
  plot_theme
```

Nearby library indices do not necessarily imply equally small waveform
differences, so the curves are more informative than index distance alone.
These examples illustrate the selection step; the recovery plot above uses
all responsive voxels.

## Argument names

Arguments use R-style names grouped by stage: `fit_hrf` and `hrf_library`;
`denoise`, `max_pcs`, `n_pcs`, `pc_stop`, `pc_voxel_*` and `noise_pool_*` for
GLMdenoise; `ridge` and `ridge_*` for fractional ridge; `cv_folds` and
`sessions` for cross-validation. `?glmsingle` lists the GLMsingle option each
one corresponds to (for example `wantlibrary` is `fit_hrf`, `brainR2` is
`noise_pool_r2`, `pcR2cutoff` is `pc_voxel_r2`, `extra_regressors` is
`nuisance`). A fixed number of PCs, GLMsingle's `pcstop = -k`, is `n_pcs = k`.

## Choices that differ from the reference implementation

Two defaults differ from the pinned Python reference (commit `1ab54a6`):

| Argument | fmrilss default | Option that follows the Python reference |
|---|---|---|
| `nuisance_in_denoise` | `"always"`: retain user nuisance regressors, such as motion, in every GLMdenoise and ridge fit | `"with_pcs"`: omit them when zero PCs are used, including from the cross-validation reference |
| `cv_zero_variance` | `"ignore"`: give zero cross-validation weight to voxels whose reference betas have zero variance | `"glmsingle"`: reproduce the reference helper's candidate-dependent treatment of zero variance |

Two further options control singular designs and the ridge calculation.
`singular = "error"` follows GLMsingle's default; `"pinv"` instead returns a
minimum-norm solution with a warning. `ridge_alpha = "grid"` uses the
reference grid interpolation to obtain ridge penalties; `"exact"` solves the
fraction equation directly and is experimental.

The automatic ON-OFF R^2 threshold uses a deterministic Gaussian-mixture
fit with a small variance floor, following the MATLAB approach; the Python
mixture fit is unseeded. Pass `noise_pool_r2` and `pc_voxel_r2` to set the thresholds
explicitly. Empty or rank-zero noise pools use zero PCs; otherwise the maximum
PC count is capped at the available rank across runs.

fmrilss computes in double precision, uses 1-based HRF indices, and applies
`percent_signal` to all model types. The Python reference uses single
precision and always scales types C and D to percent signal change.

`pc_cv_all_voxels = TRUE` fills the PC cross-validation diagnostic for every
voxel. By default only the voxels that decide the number of PCs are
evaluated, which gives identical estimates.

## How agreement with GLMsingle is checked

The package tests compare `glmsingle()` with pinned Python GLMsingle on 11
simulated scenarios: default settings, extra regressors, two sessions with
grouped folds, unequal run lengths, unrepeated conditions, rapid events,
near-collinear regressors, an all-zero voxel, a single ridge fraction, and no
HRF library. The tests require matching HRF choices and PC counts. Beta and
ridge-fraction comparisons exclude cases classified as numerically unstable
in the Python reference. Beta tolerances depend on the design's condition
number, accounting for the loss of precision in float32 normal equations;
they do not impose a uniform 1e-6 bound.

Separate tests compare the computational stages in double precision with
dense implementations of GLMsingle's stacked calculations. Together, these
checks test implementation agreement in the specified scenarios; they do not
establish which estimator is best for a new acquisition.

## Reference

Prince, J. S., Charest, I., Kurzawski, J. W., Pyles, J. A., Tarr, M. J., &
Kay, K. N. (2022). Improving the accuracy of single-trial fMRI response
estimates using GLMsingle. *eLife*, 11, e77599.
