---
title: 'Run-aware LSS with fmridesign'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Run-aware LSS with fmridesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (!requireNamespace("fmridesign", quietly = TRUE)) {
  stop(
    "This vignette requires the suggested package 'fmridesign'; ",
    "install it before building the article."
  )
}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

`lss_design()` is the run-aware adapter between `fmridesign` event and baseline
models and the OASIS estimator. Read this after `vignette("fmrilss")` and
`vignette("oasis_method")`. The adapter is useful when event tables have
run-relative onsets, the baseline is structured by run, or non-trial event
terms must remain in the common model.

This article requires the suggested `fmridesign` package. Load the three
packages used below before copying the workflow into a fresh session.

```{r packages}
suppressPackageStartupMessages({
  library(fmrilss)
  library(fmridesign)
  library(fmrihrf)
})
```

## Know the adapter contract

```{r contract-table, echo = FALSE}
contracts <- data.frame(
  Question = c(
    "Which columns become LSS targets?",
    "What happens to other event terms?",
    "How are baseline terms mapped?",
    "What does multi-basis output mean?"
  ),
  Answer = c(
    "Exactly one trialwise event term",
    "They are fixed common regressors, not extra trials",
    "drift + block go to Z; nuisance is projected as Nuisance",
    "K coefficient rows per trial, in trial-major order"
  )
)
knitr::kable(
  contracts,
  align = c("l", "l"),
  caption = "The fmridesign-to-LSS mapping used by lss_design()."
)
```

The function requires one—and only one—trialwise event term. A parametric or
condition-level event term can accompany it, but that term is part of every
trial's common design. It does not create another set of trial targets.

## Start from a two-run event table

Onsets are relative to the start of each run. Here, the same five onset times
occur in both runs; the `run` column supplies their distinct identities.

```{r event-table}
set.seed(20260821)

run_lengths <- c(110L, 130L)
TR <- 1
sframe <- sampling_frame(blocklens = run_lengths, TR = TR)

events <- data.frame(
  event_id = seq_len(10L),
  onset = rep(c(10, 30, 50, 70, 90), times = 2),
  run = rep(1:2, each = 5),
  RT = seq(0.4, 0.9, length.out = 10)
)
events$RT_c <- events$RT - mean(events$RT)
events
```

Build one trialwise target term and one reaction-time event term. The latter is
a common regressor: it adjusts every LSS model for the specified amplitude
modulation but is not reported as a trial beta.

```{r event-model}
emod <- event_model(
  onset ~ trialwise(basis = "spmg1") + hrf(RT_c),
  data = events,
  block = ~run,
  sampling_frame = sframe
)

event_dm <- design_matrix(emod)
event_meta <- attr(event_dm, "col_metadata")
table(event_meta$term_tag)
```

The ten `trial` columns are the LSS targets; the one `RT_c` column is fixed.
`lss_design()` uses the design metadata and stable column identities rather
than assuming that columns arrived in a particular order.

## Verify the run boundary

The target matrix is block diagonal by run: trials 1--5 have no energy in run
2, and trials 6--10 have no energy in run 1.

```{r public-design-reconstruction, include = FALSE}
# Reconstruct the adapter contract from public fmridesign outputs. This court
# deliberately does not call fmrilss's internal mapper.
reconstruct_public_lss <- function(event_model, trial_tag = "trial") {
  dm <- as.matrix(design_matrix(event_model))
  term <- event_model$terms[[trial_tag]]
  trial_levels <- as.character(term$condition_levels)
  K <- fmrihrf::nbasis(attr(term, "hrfspec")$hrf)
  suffix <- if (K == 1L) "" else sprintf("_b%02d", seq_len(K))
  expected <- unlist(lapply(
    paste0(trial_tag, "_", trial_levels),
    function(trial_name) paste0(trial_name, suffix)
  ), use.names = FALSE)
  target_columns <- match(expected, colnames(dm))
  stopifnot(!anyNA(target_columns), !anyDuplicated(target_columns))
  fixed_columns <- setdiff(seq_len(ncol(dm)), target_columns)
  list(
    X = dm[, target_columns, drop = FALSE],
    fixed = if (length(fixed_columns)) dm[, fixed_columns, drop = FALSE] else NULL,
    target_columns = target_columns,
    fixed_columns = fixed_columns,
    K = K,
    ntrials = length(trial_levels)
  )
}

public_design <- reconstruct_public_lss(emod)
X_trial <- public_design$X
X_event_fixed <- public_design$fixed
n_trials <- public_design$ntrials
n_time <- sum(run_lengths)
run_rows <- split(seq_len(n_time), rep(seq_along(run_lengths), run_lengths))
```

```{r run-boundary-table, echo = FALSE}
run_boundary <- data.frame(
  Check = c("run-1 trials inside run 2", "run-2 trials inside run 1"),
  Maximum_absolute_design_value = c(
    max(abs(X_trial[run_rows[[2]], 1:5, drop = FALSE])),
    max(abs(X_trial[run_rows[[1]], 6:10, drop = FALSE]))
  )
)
knitr::kable(
  run_boundary,
  digits = 12,
  caption = "Cross-run leakage in the trialwise design."
)
```

```{r run-boundary-plot, echo = FALSE, fig.cap="The trialwise design is block diagonal by run; the vertical line marks the boundary after scan 110.", fig.alt="Heatmap with activity for trials 1 through 5 only before scan 110 and trials 6 through 10 only after scan 110; no trial regressor crosses the run boundary."}
old_par <- par(bg = "white")
image(
  x = seq_len(nrow(X_trial)),
  y = seq_len(ncol(X_trial)),
  z = X_trial,
  col = hcl.colors(40, "Blues 3", rev = TRUE),
  xlab = "Scan",
  ylab = "Trial column",
  axes = FALSE
)
axis(1, at = c(1, 55, 110, 175, 240))
axis(2, at = seq_len(n_trials), las = 1)
abline(v = run_lengths[1] + 0.5, col = "#b2182b", lwd = 2, lty = 2)
box()
par(old_par)
```

```{r run-boundary-contract, include = FALSE}
stopifnot(max(run_boundary$Maximum_absolute_design_value) < 1e-12)
```

## Add a structured baseline

`baseline_model()` keeps run intercepts, drift, and nuisance inputs attached to
the same sampling frame. The example uses two synthetic motion columns per run.

```{r baseline-model}
motion <- list(
  matrix(rnorm(run_lengths[1] * 2, sd = 0.15), run_lengths[1], 2),
  matrix(rnorm(run_lengths[2] * 2, sd = 0.15), run_lengths[2], 2)
)

bmodel <- baseline_model(
  basis = "poly",
  degree = 1,
  sframe = sframe,
  intercept = "runwise",
  nuisance_list = motion
)

baseline_terms <- term_matrices(bmodel)
vapply(baseline_terms, ncol, integer(1))
```

The adapter sends `drift` and `block` to `Z`, sends `nuisance` to
`Nuisance`, and appends the fixed `RT_c` event column to `Z`. If no baseline
model is supplied, it inserts run-wise intercepts instead.

## Fit and check every trial against a full GLM

The simulation below is correctly specified for every LSS model: target trials
share a coefficient within each voxel, while drift, run intercepts, motion, and
the RT modulator all have their own common coefficients. This is an adapter
court, not evidence about recovery in arbitrary designs.

```{r simulate-data, include = FALSE}
Z_common <- as.matrix(cbind(
  baseline_terms$drift,
  baseline_terms$block,
  X_event_fixed
))
N_common <- as.matrix(baseline_terms$nuisance)
n_voxels <- 4L

common_trial_beta <- matrix(seq(0.7, 1.3, length.out = n_voxels), 1L)
Z_beta <- matrix(
  rnorm(ncol(Z_common) * n_voxels, sd = 0.3),
  ncol(Z_common),
  n_voxels
)
N_beta <- matrix(
  rnorm(ncol(N_common) * n_voxels, sd = 0.2),
  ncol(N_common),
  n_voxels
)
Y <- matrix(rowSums(X_trial), n_time, 1L) %*% common_trial_beta +
  Z_common %*% Z_beta +
  N_common %*% N_beta +
  matrix(rnorm(n_time * n_voxels), n_time, n_voxels)
```

Use explicit zero ridge when the goal is ordinary LSS. The default OASIS
configuration is penalized.

```{r fit}
fit <- lss_design(
  Y,
  emod,
  bmodel,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0
  ),
  validate = FALSE
)
dim(fit)
fit[1:4, ]
```

`validate = FALSE` above skips the adapter's optional full-design
condition-number warning; it does not disable the mandatory trial/basis
identity mapping. The full trialwise matrix contains `RT_c` as a weighted sum
of trial columns, so that screening matrix is rank deficient even though each
target-specific LSS model below is full rank. The independent court checks the
actual per-target designs and every returned cell.

```{r direct-oracle, include = FALSE}
direct <- matrix(NA_real_, n_trials, n_voxels)
direct_rank <- direct_columns <- integer(n_trials)
for (j in seq_len(n_trials)) {
  D <- cbind(
    X_trial[, j],
    rowSums(X_trial[, -j, drop = FALSE]),
    Z_common,
    N_common
  )
  qr_D <- qr(D)
  direct_rank[j] <- qr_D$rank
  direct_columns[j] <- ncol(D)
  direct[j, ] <- qr.coef(qr_D, Y)[1L, ]
}
direct_error <- max(abs(unname(fit) - direct))
stopifnot(
  direct_error < 1e-10,
  all(direct_rank == direct_columns),
  identical(
    attr(fit, "trial_basis_map")$source_column,
    public_design$target_columns
  )
)
```

```{r direct-result, echo = FALSE}
knitr::kable(
  data.frame(
    Maximum_absolute_error = direct_error,
    Minimum_target_model_rank = min(direct_rank),
    Target_model_columns = unique(direct_columns)
  ),
  digits = 12,
  caption = "lss_design() versus independently assembled full GLMs."
)
```

The zero discrepancy also proves that the RT event term and baseline nuisance
span entered the model without becoming extra trial rows.

```{r fixed-term-mutation, include = FALSE}
direct_without_rt <- matrix(NA_real_, n_trials, n_voxels)
Z_without_rt <- as.matrix(cbind(baseline_terms$drift, baseline_terms$block))
for (j in seq_len(n_trials)) {
  D_without_rt <- cbind(
    X_trial[, j],
    rowSums(X_trial[, -j, drop = FALSE]),
    Z_without_rt,
    N_common
  )
  direct_without_rt[j, ] <- qr.coef(qr(D_without_rt), Y)[1L, ]
}
rt_omission_error <- max(abs(direct_without_rt - direct))
stopifnot(rt_omission_error > 1e-3)
```

## Decide whether other-trial effects are pooled across runs

Run-aware onset placement does not make the default coefficient estimator
run-local. A single `lss_design()` call passes all target columns to one OASIS
fit, whose other-trial regressor aggregates trials from both runs. A run-1
target can therefore change when the run-2 response changes. This is a pooled
multi-run estimand, not leakage in the event convolution.

The next mutation adds signal only in run 2. It changes the joint fit's run-1
coefficients, while two explicitly separate run fits leave run 1 unchanged.

```{r run-pooling-contract, include = FALSE}
Y_shifted <- Y
Y_shifted[run_rows[[2]], ] <- Y_shifted[run_rows[[2]], ] +
  matrix(rowSums(X_trial[run_rows[[2]], 6:10, drop = FALSE]), run_lengths[2], 1L) %*%
  matrix(rep(20, n_voxels), 1L, n_voxels)

fit_shifted <- lss_design(
  Y_shifted,
  emod,
  bmodel,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute", ridge_x = 0, ridge_b = 0
  ),
  validate = FALSE
)

fit_one_run <- function(response, run) {
  scan_start <- sum(run_lengths[seq_len(run - 1L)]) + 1L
  scan_rows <- scan_start:(scan_start + run_lengths[run] - 1L)
  events_run <- events[events$run == run, , drop = FALSE]
  events_run$run <- 1L
  sframe_run <- sampling_frame(blocklens = run_lengths[run], TR = TR)
  emod_run <- event_model(
    onset ~ trialwise(basis = "spmg1") + hrf(RT_c),
    data = events_run,
    block = ~run,
    sampling_frame = sframe_run
  )
  bmodel_run <- baseline_model(
    basis = "poly",
    degree = 1,
    sframe = sframe_run,
    intercept = "runwise",
    nuisance_list = list(motion[[run]])
  )
  beta_run <- lss_design(
    response[scan_rows, , drop = FALSE],
    emod_run,
    bmodel_run,
    method = "oasis",
    oasis = oasis_options(
      ridge_mode = "absolute", ridge_x = 0, ridge_b = 0
    ),
    validate = FALSE
  )
  map_run <- attr(beta_run, "trial_basis_map")
  map_run$local_trial <- map_run$trial
  map_run$local_trial_name <- map_run$trial_name
  map_run$local_source_column <- map_run$source_column
  map_run$global_event_id <- events_run$event_id[map_run$trial]
  map_run$run <- run
  global_name <- sprintf(
    "event_%02d_basis_%02d", map_run$global_event_id, map_run$basis
  )
  map_run$trial <- map_run$global_event_id
  map_run$trial_name <- sprintf("event_%02d", map_run$global_event_id)
  map_run$source_column <- public_design$target_columns[map_run$global_event_id]
  map_run$column <- global_name
  map_run$name <- global_name
  map_run$output_name <- global_name
  rownames(beta_run) <- global_name
  attr(beta_run, "trial_basis_map") <- map_run
  list(beta = beta_run, map = map_run)
}

combine_run_fits <- function(parts) {
  beta <- do.call(rbind, lapply(parts, `[[`, "beta"))
  map <- do.call(rbind, lapply(parts, `[[`, "map"))
  rownames(map) <- NULL
  map$output_row <- seq_len(nrow(map))
  attr(beta, "trial_basis_map") <- map
  beta
}

local_fit <- combine_run_fits(lapply(seq_along(run_lengths), function(run) {
  fit_one_run(Y, run)
}))
local_shifted <- combine_run_fits(lapply(seq_along(run_lengths), function(run) {
  fit_one_run(Y_shifted, run)
}))

local_map <- attr(local_fit, "trial_basis_map")

run_perturbation <- data.frame(
  Estimator = c("one pooled two-run fit", "two separate run fits"),
  Maximum_change_in_run1_beta = c(
    max(abs(fit_shifted[1:5, ] - fit[1:5, ])),
    max(abs(local_shifted[1:5, ] - local_fit[1:5, ]))
  )
)
stopifnot(
  run_perturbation$Maximum_change_in_run1_beta[1] > 1e-6,
  run_perturbation$Maximum_change_in_run1_beta[2] < 1e-12,
  identical(local_map$global_event_id, events$event_id),
  !anyDuplicated(local_map$global_event_id),
  !anyDuplicated(rownames(local_fit)),
  identical(local_map$trial, events$event_id),
  identical(local_map$source_column, public_design$target_columns),
  identical(local_map$column, rownames(local_fit)),
  identical(local_map$name, rownames(local_fit)),
  identical(local_map$output_name, rownames(local_fit)),
  !anyDuplicated(paste(local_map$trial, local_map$basis))
)
```

```{r run-pooling-result, echo = FALSE}
knitr::kable(
  run_perturbation,
  digits = 12,
  caption = "Effect of a run-2-only response mutation on run-1 coefficients."
)
```

Use one joint call only when that cross-run pooling is the intended LSS model.
For run-local coefficients, fit each run with its own event and baseline models,
then restore global event identifiers in both row names and the trial/basis map
before combining rows, as the helper above does. Run-specific prewhitening does
not by itself change the pooled other-trial estimand.

## Preserve output identity

The result carries the event model, baseline model, sampling frame, and an
explicit trial/basis map. Use the map or row names; do not reconstruct trial
identity from a presumed source-column order.

```{r output-identity}
attributes_kept <- c(
  event_model = !is.null(attr(fit, "event_model")),
  baseline_model = !is.null(attr(fit, "baseline_model")),
  sampling_frame = !is.null(attr(fit, "sampling_frame")),
  trial_basis_map = !is.null(attr(fit, "trial_basis_map"))
)
attributes_kept
identity_map <- attr(fit, "trial_basis_map")
stopifnot(identical(identity_map$trial, events$event_id))
identity_map$input_event_id <- events$event_id
identity_map[1:3, c("input_event_id", "trial", "basis", "output_row")]
```

## Multi-basis designs return coefficients, not amplitudes

With SPMG3, `lss_design()` detects three basis columns per trial and returns
rows in trial-major, basis-minor order. The next simulation is correctly
specified in that three-dimensional basis and is checked against independent
full GLMs.

```{r multibasis-data, include = FALSE}
emod_spmg3 <- event_model(
  onset ~ trialwise(basis = "spmg3", nbasis = 3),
  data = events,
  block = ~run,
  sampling_frame = sframe
)
public_spmg3 <- reconstruct_public_lss(emod_spmg3)
X_spmg3 <- public_spmg3$X
K <- public_spmg3$K
run_intercepts <- cbind(
  rep(c(1, 0), times = run_lengths),
  rep(c(0, 1), times = run_lengths)
)
summed_spmg3 <- vapply(seq_len(K), function(k) {
  rowSums(X_spmg3[, seq(k, ncol(X_spmg3), by = K), drop = FALSE])
}, numeric(n_time))
theta_spmg3 <- rbind(
  seq(0.8, 1.0, length.out = n_voxels),
  seq(0.05, 0.15, length.out = n_voxels),
  seq(-0.12, -0.06, length.out = n_voxels)
)
Y_spmg3 <- summed_spmg3 %*% theta_spmg3 +
  run_intercepts %*% matrix(rnorm(2 * n_voxels), 2L, n_voxels) +
  matrix(rnorm(n_time * n_voxels), n_time, n_voxels)
```

```{r multibasis-fit}
fit_spmg3 <- lss_design(
  Y_spmg3,
  emod_spmg3,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0
  ),
  validate = FALSE
)
c(rows = nrow(fit_spmg3), trials = n_trials, basis_dimension = K)
rownames(fit_spmg3)[1:6]
attr(fit_spmg3, "trial_basis_map")[
  1:6, c("trial", "basis", "source_column", "output_row")
]
```

```{r multibasis-oracle, include = FALSE}
direct_spmg3 <- matrix(NA_real_, n_trials * K, n_voxels)
for (j in seq_len(n_trials)) {
  target_rows <- ((j - 1L) * K + 1L):(j * K)
  other_trials <- setdiff(seq_len(n_trials), j)
  B <- vapply(seq_len(K), function(k) {
    columns <- (other_trials - 1L) * K + k
    rowSums(X_spmg3[, columns, drop = FALSE])
  }, numeric(n_time))
  D <- cbind(X_spmg3[, target_rows, drop = FALSE], B, run_intercepts)
  direct_spmg3[target_rows, ] <- qr.coef(qr(D), Y_spmg3)[seq_len(K), ]
}
spmg3_error <- max(abs(unname(fit_spmg3) - direct_spmg3))
stopifnot(
  all(dim(fit_spmg3) == c(n_trials * K, n_voxels)),
  spmg3_error < 1e-10,
  identical(
    attr(fit_spmg3, "trial_basis_map")$source_column,
    public_spmg3$target_columns
  )
)
```

```{r multibasis-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_absolute_error = spmg3_error),
  digits = 12,
  caption = "Multi-basis lss_design() versus independent full GLMs."
)
```

Each trial now has canonical, temporal-derivative, and dispersion-derivative
coefficients. The canonical coefficient alone is not a normalized response
amplitude. Keep all three rows unless a separately defined shape and amplitude
estimand justifies a reduction.

## Ridge, standard errors, and whitening retain OASIS semantics

The adapter does not change the inference contract:

- `oasis_options()` uses fractional ridge by default.
- `return_se = TRUE` requires zero ridge, a fixed full-rank design, positive
  residual degrees of freedom, and no estimated prewhitening.
- A non-trial event term remains common whether or not ridge is used.

For multiple runs, whitening needs scan-level segmentation. `lss_design()`
infers it from the sampling frame. The explicit vector below is equivalent and
shows the underlying contract; if supplied, it must encode the same boundaries.
The event model's `blockids` describe events, not scans, so do not pass that
shorter vector as `prewhiten$runs`.

```{r whitening-recipe, eval = FALSE}
scan_run <- rep(seq_along(run_lengths), times = run_lengths)

fit_whitened <- lss_design(
  Y,
  emod,
  bmodel,
  method = "oasis",
  oasis = oasis_options(),
  prewhiten = prewhiten_options(
    method = "ar",
    p = 1,
    pooling = "run",
    runs = scan_run
  )
)
```

This is a recipe, not a recommendation that AR(1) fits every dataset. Choose the
noise model from residual diagnostics. Shared-design OASIS calls reject voxel-
or parcel-specific whitening operators because those require correspondingly
voxel- or parcel-specific filtered designs.

## Validation and failure modes

With `validate = TRUE`, the adapter checks time dimensions and sampling-frame
agreement. It computes a scale-dependent condition number for the full
assembled design and emits a warning only when that number exceeds 30 and the
effective ridge penalties are all zero; it does not return the number as a
diagnostic. Treat the warning as screening, not as a characterization of every
residualized LSS target model. Mandatory semantic checks remain active
regardless of the flag: the event design must have metadata, exactly one
trialwise term, a complete trial/basis rectangle, and stable identities.

Common failures are direct:

- `nrow(Y)` must equal `sum(blocklens(sframe))`.
- The event and baseline models must use the same sampling frame.
- Multi-basis metadata must agree with the HRF basis dimension.
- Parametric and condition-level terms are fixed regressors; they are not
  additional trialwise targets.

## Next steps

- `vignette("voxel-wise-hrf")` --- normalized voxel-specific HRF shapes
- `vignette("sbhm")` --- library-constrained voxel-specific HRFs
