---
title: 'Practical OASIS: Design, Regularization, and Diagnostics'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Practical OASIS: Design, Regularization, and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
library(fmrilss)
library(fmrihrf)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

OASIS is the `lss()` backend to use when trial-wise estimation needs more than
a fixed one-basis design: event-based construction, multiple HRF bases, ridge
regularization, blocked voxel products, or model-based standard errors. This
guide follows `vignette("fmrilss")`, keeps the public workflow visible, and
routes the derivation to `vignette("oasis_theory")`.

## Choose the contract before fitting

```{r contract-table, echo = FALSE}
contracts <- data.frame(
  Goal = c(
    "Exact ordinary LSS",
    "Regularized coefficients",
    "Model-based coefficient SEs",
    "Multiple HRF bases"
  ),
  Required_choice = c(
    "ridge_x = ridge_b = 0; absolute mode makes the zero scale explicit",
    "absolute or fractional ridge; default is fractional 0.05",
    "zero ridge, fixed full-rank design, no estimated prewhitening",
    "interpret K coefficients per trial; do not call them one amplitude"
  )
)
knitr::kable(
  contracts,
  align = c("l", "l"),
  caption = "OASIS choices that change the estimand or the available uncertainty."
)
```

The bare call `oasis_options()` is penalized. It is not the configuration for
an exact comparison with an unpenalized LSS backend. Likewise,
`return_se = TRUE` is intentionally unavailable with ridge, estimated
prewhitening, HRF-grid selection, or voxel-adaptive HRFs.

## Build a two-condition example

The example has one target condition, one explicitly modeled other condition,
a shared intercept and drift, heterogeneous target-trial coefficients, and a
fixed iid noise scale. Keeping the mean signal separate lets us calculate the
ridge bias--variance trade-off exactly for this design.

```{r event-spec}
set.seed(20260811)

n_time <- 180
n_voxels <- 6
TR <- 1
sframe <- fmrihrf::sampling_frame(blocklens = n_time, TR = TR)
times <- fmrihrf::samples(sframe, global = TRUE)

onsets_a <- c(10, 20, 31, 43, 56, 68, 81, 95, 108, 121, 135, 149)
onsets_b <- c(15, 37, 61, 86, 112, 141)
n_trials <- length(onsets_a)

design_spec <- list(
  sframe = sframe,
  cond = list(
    onsets = onsets_a,
    hrf = fmrihrf::HRF_SPMG1,
    duration = 0,
    amplitude = 1,
    span = 30
  ),
  others = list(list(
    onsets = onsets_b,
    hrf = fmrihrf::HRF_SPMG1,
    duration = 0,
    amplitude = 1,
    span = 30
  )),
  precision = 0.1,
  method = "conv"
)
```

For the simulation and independent oracles, the same trial and other-condition
regressors are also evaluated explicitly. In an analysis, `Y` would normally
be the observed time-by-voxel response rather than a simulated matrix.

```{r design-evaluation, include = FALSE}

trial_set <- fmrihrf::regressor_set(
  onsets = onsets_a,
  fac = factor(seq_len(n_trials)),
  hrf = fmrihrf::HRF_SPMG1,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_trials <- fmrihrf::evaluate(
  trial_set, grid = times, precision = 0.1, method = "conv"
)
X_trials <- as.matrix(X_trials)
colnames(X_trials) <- sprintf("trial_%02d", seq_len(n_trials))

other_reg <- fmrihrf::regressor(
  onsets = onsets_b,
  hrf = fmrihrf::HRF_SPMG1,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_other <- fmrihrf::evaluate(
  other_reg, grid = times, precision = 0.1, method = "conv"
)
X_other <- matrix(as.numeric(X_other), n_time, 1L,
                  dimnames = list(NULL, "condition_B"))

Z <- cbind(
  intercept = 1,
  linear_trend = seq(-1, 1, length.out = n_time)
)
```

```{r simulate-response}
true_beta <- matrix(
  rnorm(n_trials * n_voxels, mean = 1, sd = 0.35),
  n_trials,
  n_voxels
)
other_beta <- matrix(seq(0.4, 0.9, length.out = n_voxels), 1L)
fixed_beta <- rbind(
  seq(-0.3, 0.3, length.out = n_voxels),
  seq(0.2, -0.2, length.out = n_voxels)
)
mean_signal <- X_trials %*% true_beta +
  X_other %*% other_beta +
  Z %*% fixed_beta
noise_sd <- 1.5
Y <- mean_signal + matrix(rnorm(n_time * n_voxels, sd = noise_sd),
                          n_time, n_voxels)
```

```{r dimensions, echo = FALSE}
dimensions <- data.frame(
  Object = c("Y", "target trial design", "other-condition aggregate", "fixed design"),
  Rows = c(nrow(Y), nrow(X_trials), nrow(X_other), nrow(Z)),
  Columns = c(ncol(Y), ncol(X_trials), ncol(X_other), ncol(Z))
)
knitr::kable(dimensions, caption = "Objects used in the practical example.")
```

## Fit from events

Put the condition whose trial coefficients you want in `cond`. Put modeled
conditions that should not become trial targets in `others`; OASIS adds their
aggregate basis columns to the common span.

```{r basic-fit}
fit_default <- lss(
  Y = Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(design_spec = design_spec)
)
dim(fit_default)
fit_default[1:4, 1:3]
```

The default result is an `n_trials` by `n_voxels` matrix because this is a
one-basis design. These are fractionally penalized coefficients. The simulation
really contains condition B, so `others` is doing identifiable work rather than
decorating the call.

## Match ordinary LSS exactly when that is the target

Use explicit zero ridge. The verification assembles every corresponding GLM
independently as `[target trial, other target trials, Z, condition B]` and
checks all trials and voxels.

```{r unpenalized-fit}
fit_unpenalized <- lss(
  Y = Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    design_spec = design_spec,
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0
  )
)
```

```{r direct-oracle, include = FALSE}
direct_lss <- matrix(NA_real_, n_trials, n_voxels)
for (j in seq_len(n_trials)) {
  direct_design <- cbind(
    X_trials[, j],
    rowSums(X_trials[, -j, drop = FALSE]),
    Z,
    X_other
  )
  direct_lss[j, ] <- qr.coef(qr(direct_design), Y)[1L, ]
}
direct_error <- max(abs(unname(fit_unpenalized) - direct_lss))
stopifnot(direct_error < 1e-10)
```

```{r direct-oracle-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_absolute_error = direct_error),
  digits = 12,
  caption = "Unpenalized OASIS versus independent trial-wise GLMs."
)
```

For a fixed one-basis `X`, this is the same estimand as the other unpenalized
LSS backends. OASIS is useful here because the event and other-condition design
can be constructed and checked in the same workflow.

## Evaluate ridge on the scale you care about

Ridge reduces sampling variance by accepting bias. A smaller spread of fitted
trials is not, by itself, evidence of improvement. For this fixed design and
truth, OASIS is linear in `Y`, so applying each fit to the noise-free signal and
to the identity matrix gives exact conditional bias and variance under the iid
noise model.

```{r ridge-analysis, include = FALSE}
ridge_levels <- c(0, 0.01, 0.05)

fit_fractional <- function(response, eta) {
  lss(
    response,
    X_trials,
    Z = Z,
    Nuisance = X_other,
    method = "oasis",
    oasis = oasis_options(
      ridge_mode = "fractional",
      ridge_x = eta,
      ridge_b = eta
    )
  )
}

ridge_metrics <- lapply(ridge_levels, function(eta) {
  expected_beta <- fit_fractional(mean_signal, eta)
  operator <- fit_fractional(diag(n_time), eta)
  bias <- unname(expected_beta) - true_beta
  coefficient_variance <- noise_sd^2 * rowSums(operator^2)
  mean_variance <- mean(coefficient_variance)
  mean_squared_bias <- mean(bias^2)
  data.frame(
    Fractional_ridge = eta,
    Mean_absolute_bias = mean(abs(bias)),
    Mean_squared_bias = mean_squared_bias,
    Mean_variance = mean_variance,
    Mean_MSE = mean_squared_bias + mean_variance,
    RMSE = sqrt(mean_squared_bias + mean_variance)
  )
})
ridge_metrics <- do.call(rbind, ridge_metrics)

set.seed(20260812)
mc_repetitions <- 100
mc_mse <- vapply(ridge_levels, function(eta) {
  squared_error <- replicate(mc_repetitions, {
    noisy_y <- mean_signal + matrix(
      rnorm(n_time * n_voxels, sd = noise_sd), n_time, n_voxels
    )
    mean((unname(fit_fractional(noisy_y, eta)) - true_beta)^2)
  })
  mean(squared_error)
}, numeric(1))
ridge_metrics$Monte_Carlo_MSE <- mc_mse
```

```{r ridge-table, echo = FALSE}
knitr::kable(
  ridge_metrics,
  digits = 4,
  caption = paste0(
    "Exact conditional ridge metrics and a ", mc_repetitions,
    "-repetition check for this design, truth, and noise scale."
  )
)
```

```{r ridge-plot, echo = FALSE, fig.cap="Exact squared bias, variance, and MSE for three fractional-ridge settings in the fixed simulation; all bars use squared-beta units.", fig.alt="Grouped bars show mean MSE falling from about 0.210 at zero ridge to 0.194 at fractional ridge 0.05 as variance falls and squared bias rises."}
plot_values <- t(as.matrix(ridge_metrics[, c(
  "Mean_squared_bias", "Mean_variance", "Mean_MSE"
)]))
colnames(plot_values) <- format(ridge_levels, trim = TRUE)
old_par <- par(bg = "white")
barplot(
  plot_values,
  beside = TRUE,
  col = c("#9ecae1", "#3182bd", "#de2d26"),
  border = NA,
  xlab = "Fractional ridge",
  ylab = "Mean squared beta units",
  legend.text = c("Squared bias", "Variance", "MSE"),
  args.legend = list(bty = "n", x = "topleft")
)
par(old_par)
```

```{r ridge-contract, include = FALSE}
stopifnot(
  all(is.finite(as.matrix(ridge_metrics[, -1]))),
  all(ridge_metrics$Mean_MSE >= ridge_metrics$Mean_squared_bias),
  all(ridge_metrics$Mean_MSE >= ridge_metrics$Mean_variance),
  max(abs(ridge_metrics$Mean_MSE - ridge_metrics$Monte_Carlo_MSE)) < 0.03
)
```

This is one scenario, not a universal ranking. Choose a ridge policy using a
simulation that matches the planned timing, basis, nuisance structure, signal
scale, and downstream loss. The package does not justify generic penalty
cutoffs from inter-stimulus interval alone. Even the zero-penalty row has a
small bias against the generating trial effects: ordinary LSS gives all other
target trials one shared coefficient, while this simulation gives them
heterogeneous coefficients. The table therefore evaluates the full LSS
estimator against the generating effects, not merely ridge against an
unpenalized LSS reference.

## Inspect design diagnostics

`return_diag = TRUE` changes the return to a list. For a one-basis fit, `d`,
`alpha`, and `s` are the target energy, target--other cross-product, and
other-trial energy after projection.

```{r diagnostics}
fit_diag <- lss(
  Y,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(return_diag = TRUE)
)
diagnostics <- with(fit_diag$diag, data.frame(
  target_energy = d,
  other_energy = s,
  target_other_correlation = alpha / sqrt(d * s)
))
head(diagnostics, 4)
```

These are unpenalized cross-products on the residualized design scale. They are
useful for finding low-energy or highly correlated trials, but they are not a
condition number, a ridge calibration, or an inferential test.

## Multi-basis output is not one amplitude

An event-built SPMG3 design has three rows per trial: canonical, temporal
derivative, and dispersion-derivative coefficients in that basis. Their scale
depends on the basis definition.

```{r multi-basis}
spec_spmg3 <- design_spec
spec_spmg3$cond$hrf <- fmrihrf::HRF_SPMG3
spec_spmg3$others[[1]]$hrf <- fmrihrf::HRF_SPMG3

fit_spmg3 <- lss(
  Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(design_spec = spec_spmg3, return_diag = TRUE)
)
K_spmg3 <- 3L
c(rows = nrow(fit_spmg3$beta), trials = n_trials, basis_dimension = K_spmg3)
rownames(fit_spmg3$beta)[1:6]
```

The `NK` by `V` output is in canonical trial-major, basis-minor order. Keep the
row names or reshape explicitly; do not infer a scalar response amplitude from
the canonical row alone unless a separate normalization and estimand justify
that operation.

If you already have a raw multi-basis matrix, identity is part of the input
contract. `X_multi` must have unique, non-empty column names, and the map must
identify every column with one exact-integer trial and basis pair. OASIS
canonicalizes the matrix to trial-major, basis-minor order and attaches the
canonical map to the returned beta matrix.

```{r raw-multi-basis-design, include = FALSE}
raw_spmg3_set <- fmrihrf::regressor_set(
  onsets = onsets_a,
  fac = factor(seq_len(n_trials)),
  hrf = fmrihrf::HRF_SPMG3,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_spmg3_source <- as.matrix(fmrihrf::evaluate(
  raw_spmg3_set, grid = times, precision = 0.1, method = "conv"
))
raw_spmg3_other <- fmrihrf::regressor(
  onsets = onsets_b,
  hrf = fmrihrf::HRF_SPMG3,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_other_spmg3 <- as.matrix(fmrihrf::evaluate(
  raw_spmg3_other, grid = times, precision = 0.1, method = "conv"
))
```

For this controlled source construction, we know the trial-major, basis-minor
order and capture those identities *before* deliberately permuting its columns.
For a different upstream design, use that producer's metadata; never infer
identity from the order in which an arbitrary matrix happens to arrive.

```{r raw-multi-basis-contract}
K <- K_spmg3
source_map <- data.frame(
  column = sprintf("source_%02d", seq_len(ncol(X_spmg3_source))),
  trial = rep(seq_len(n_trials), each = K),
  basis = rep(seq_len(K), times = n_trials)
)
colnames(X_spmg3_source) <- source_map$column

column_order <- rev(seq_len(ncol(X_spmg3_source)))
X_multi <- X_spmg3_source[, column_order, drop = FALSE]

fit_raw_multi <- lss(
  Y,
  X_multi,
  Z = Z,
  Nuisance = X_other_spmg3,
  method = "oasis",
  oasis = oasis_options(
    K = K,
    ntrials = n_trials,
    trial_basis_map = source_map,
    return_diag = TRUE
  )
)
raw_permutation_error <- max(abs(fit_raw_multi$beta - fit_spmg3$beta))
c(maximum_beta_error = raw_permutation_error)
head(attr(fit_raw_multi$beta, "trial_basis_map"), 3)
```

Without optional returns, OASIS returns the `NK` by `V` beta matrix. Setting
`return_diag = TRUE` or `return_se = TRUE` returns a list containing `beta`,
then `diag` when requested, then `se` when requested. `beta` and `se` are both
`NK` by `V`. For `K = 1`, `diag` contains vectors `d`, `alpha`, and `s`, each of
length `N`. For `K > 1`, it contains `D`, `C`, and `E` arrays of shape
`K` by `K` by `N`. Diagnostics describe the residualized design, or the
whitened-and-residualized design when whitening is active.

```{r multi-basis-contract, include = FALSE}
stopifnot(
  raw_permutation_error < 1e-10,
  all(dim(fit_spmg3$beta) == c(n_trials * K_spmg3, n_voxels)),
  all(dim(fit_spmg3$diag$D) == c(K_spmg3, K_spmg3, n_trials)),
  all(dim(fit_spmg3$diag$C) == c(K_spmg3, K_spmg3, n_trials)),
  all(dim(fit_spmg3$diag$E) == c(K_spmg3, K_spmg3, n_trials))
)
```

## Standard errors require the unpenalized fixed-design model

For a judgeable SE example, generate data that exactly satisfy each LSS model:
all target trials share a common coefficient vector, condition B and `Z` are in
the fitted common span, and the temporal errors are iid Gaussian.

```{r se-data, include = FALSE}
set.seed(20260813)
common_trial_beta <- matrix(seq(0.6, 1.1, length.out = n_voxels), 1L)
Y_se <- matrix(rowSums(X_trials), n_time, 1L) %*% common_trial_beta +
  X_other %*% other_beta +
  Z %*% fixed_beta +
  matrix(rnorm(n_time * n_voxels, sd = 1), n_time, n_voxels)
```

```{r se-fit}
fit_se <- lss(
  Y_se,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  )
)
dim(fit_se$beta)
dim(fit_se$se)
```

```{r se-oracle, include = FALSE}
direct_beta <- direct_se <- matrix(NA_real_, n_trials, n_voxels)
for (j in seq_len(n_trials)) {
  D <- cbind(
    X_trials[, j],
    rowSums(X_trials[, -j, drop = FALSE]),
    Z,
    X_other
  )
  gram_inv <- solve(crossprod(D))
  coef_j <- gram_inv %*% crossprod(D, Y_se)
  residual_j <- Y_se - D %*% coef_j
  sigma2_j <- colSums(residual_j^2) / (n_time - ncol(D))
  direct_beta[j, ] <- coef_j[1L, ]
  direct_se[j, ] <- sqrt(gram_inv[1L, 1L] * sigma2_j)
}
se_oracle <- data.frame(
  Maximum_beta_error = max(abs(unname(fit_se$beta) - direct_beta)),
  Maximum_SE_error = max(abs(unname(fit_se$se) - direct_se))
)
```

```{r se-result, echo = FALSE}
knitr::kable(
  se_oracle,
  digits = 12,
  caption = "OASIS beta and SE agreement with independent full GLMs."
)
```

```{r se-contract, include = FALSE}
stopifnot(
  all(dim(fit_se$beta) == c(n_trials, n_voxels)),
  all(dim(fit_se$se) == c(n_trials, n_voxels)),
  max(as.matrix(se_oracle)) < 1e-10
)
```

These are conditional, coefficient-scale model SEs under spherical temporal
errors. They are not robust to autocorrelation or heteroskedasticity. The
result does not carry a ready-made multiple-testing or population-inference
procedure, and estimated whitening cannot be combined with `return_se = TRUE`.

## FIR coefficients and their uncertainty

FIR output also has one row per basis coefficient, trial, and voxel. To keep
the uncertainty unit honest, the next plot shows one trial and one voxel with
its coefficient-wise model SE. It is not an across-voxel or across-trial SE for
an average HRF. The data generator uses a correctly specified FIR
target-plus-other-trials model with iid Gaussian errors, and the verification
assembles every corresponding GLM independently.

```{r fir-data, include = FALSE}
K_fir <- 6L
fir_hrf <- fmrihrf::hrf_fir_generator(nbasis = K_fir, span = 24)
spec_fir <- design_spec
spec_fir$cond$hrf <- fir_hrf
spec_fir$cond$span <- 24
spec_fir$others[[1]]$hrf <- fir_hrf
spec_fir$others[[1]]$span <- 24

# Generate the oracle design independently through fmrihrf's public API.
fir_trial_set <- fmrihrf::regressor_set(
  onsets = onsets_a,
  fac = factor(seq_len(n_trials)),
  hrf = fir_hrf,
  duration = 0,
  amplitude = 1,
  span = 24,
  summate = TRUE
)
X_fir <- as.matrix(fmrihrf::evaluate(
  fir_trial_set, grid = times, precision = 0.1, method = "conv"
))
fir_other_reg <- fmrihrf::regressor(
  onsets = onsets_b,
  hrf = fir_hrf,
  duration = 0,
  amplitude = 1,
  span = 24,
  summate = TRUE
)
X_other_fir <- as.matrix(fmrihrf::evaluate(
  fir_other_reg, grid = times, precision = 0.1, method = "conv"
))
summed_fir <- vapply(seq_len(K_fir), function(k) {
  rowSums(X_fir[, seq(k, ncol(X_fir), by = K_fir), drop = FALSE])
}, numeric(n_time))
theta_fir <- outer(
  seq(0.8, -0.2, length.out = K_fir),
  seq(0.8, 1.2, length.out = n_voxels)
)
other_theta_fir <- outer(
  seq(0.3, -0.1, length.out = ncol(X_other_fir)),
  seq(0.9, 1.1, length.out = n_voxels)
)
set.seed(20260815)
Y_fir <- summed_fir %*% theta_fir +
  X_other_fir %*% other_theta_fir +
  Z %*% fixed_beta +
  matrix(rnorm(n_time * n_voxels), n_time, n_voxels)
```

```{r fir-fit}
fit_fir <- lss(
  Y_fir,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    design_spec = spec_fir,
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  )
)
dim(fit_fir$beta)
dim(fit_fir$se)
```

```{r fir-oracle, include = FALSE}
# Assemble every corresponding full GLM independently.
direct_fir_beta <- direct_fir_se <- matrix(
  NA_real_, n_trials * K_fir, n_voxels
)
for (j in seq_len(n_trials)) {
  target_rows <- ((j - 1L) * K_fir + 1L):(j * K_fir)
  other_trial_ids <- setdiff(seq_len(n_trials), j)
  other_trials <- vapply(seq_len(K_fir), function(k) {
    columns <- (other_trial_ids - 1L) * K_fir + k
    rowSums(X_fir[, columns, drop = FALSE])
  }, numeric(n_time))
  D <- cbind(
    X_fir[, target_rows, drop = FALSE], other_trials, Z, X_other_fir
  )
  gram_inv <- solve(crossprod(D))
  coef_j <- gram_inv %*% crossprod(D, Y_fir)
  residual_j <- Y_fir - D %*% coef_j
  sigma2_j <- colSums(residual_j^2) / (n_time - ncol(D))
  direct_fir_beta[target_rows, ] <- coef_j[seq_len(K_fir), , drop = FALSE]
  direct_fir_se[target_rows, ] <- sqrt(
    outer(diag(gram_inv)[seq_len(K_fir)], sigma2_j)
  )
}

fir_oracle <- data.frame(
  Maximum_beta_error = max(abs(unname(fit_fir$beta) - direct_fir_beta)),
  Maximum_SE_error = max(abs(unname(fit_fir$se) - direct_fir_se))
)

fir_rows <- seq_len(K_fir)
fir_example <- data.frame(
  basis = fir_rows,
  coefficient = fit_fir$beta[fir_rows, 1L],
  se = fit_fir$se[fir_rows, 1L]
)
```

```{r fir-oracle-result, echo = FALSE}
knitr::kable(
  fir_oracle,
  digits = 12,
  caption = "FIR beta and SE agreement with independent full GLMs."
)
```

```{r fir-plot, echo = FALSE, fig.cap="FIR basis coefficients plus or minus one conditional model SE for trial 1, voxel 1; the y-axis is coefficient scale, not a normalized HRF amplitude.", fig.alt="Six FIR estimates for one trial and voxel: the first is positive, the second is near zero, and the remaining four are negative; one-SE intervals are wide, with the second crossing zero."}
old_par <- par(bg = "white")
plot(
  fir_example$basis,
  fir_example$coefficient,
  type = "b",
  pch = 19,
  xlab = "FIR basis index",
  ylab = "FIR basis coefficient",
  ylim = range(
    fir_example$coefficient - fir_example$se,
    fir_example$coefficient + fir_example$se
  )
)
arrows(
  fir_example$basis,
  fir_example$coefficient - fir_example$se,
  fir_example$basis,
  fir_example$coefficient + fir_example$se,
  angle = 90,
  code = 3,
  length = 0.04
)
abline(h = 0, col = "grey70", lty = 2)
par(old_par)
```

```{r fir-contract, include = FALSE}
stopifnot(
  all(dim(fit_fir$beta) == c(n_trials * K_fir, n_voxels)),
  all(dim(fit_fir$se) == c(n_trials * K_fir, n_voxels)),
  all(is.finite(fir_example$se)),
  all(fir_example$se >= 0),
  max(as.matrix(fir_oracle)) < 1e-10
)
```

FIR models can be poorly conditioned because each trial contributes many
columns. Ridge may be scientifically useful, but ridge and `return_se = TRUE`
are different contracts; the package does not report inferential FIR SEs for a
penalized fit.

## HRF-grid selection is exploratory

`hrf_grid` chooses one candidate from the observed `Y` using a matched-filter
score after residualizing both the response and candidate against the full
common span (`Z`, `Nuisance`, and modeled `others`), then fits that design. The
call shown returns the beta matrix. Even when design diagnostics are requested,
the public `lss()` route does not expose the winning candidate or candidate
scores, and it does not return selection-adjusted uncertainty.

```{r hrf-grid}
grid_spec <- design_spec
grid_spec$hrf_grid <- list(fmrihrf::HRF_SPMG1, fmrihrf::HRF_GAMMA)

fit_grid <- lss(
  Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    design_spec = grid_spec,
    ridge_mode = "fractional",
    ridge_x = 0.01,
    ridge_b = 0.01
  )
)
dim(fit_grid)
```

The following known-truth check makes that conditioning rule judgeable. Its
target signal is Gaussian, while a much stronger other-condition signal uses
SPMG1. The grid fit must match the otherwise identical fit with the Gaussian
target fixed in advance.

```{r hrf-grid-contract, include = FALSE}
grid_target_onsets <- c(10, 40, 70, 100, 130)
grid_other_onsets <- c(20, 50, 80, 110, 140)
grid_aggregate <- function(onsets, hrf) {
  reg <- fmrihrf::regressor(
    onsets = onsets,
    hrf = hrf,
    duration = 0,
    amplitude = 1,
    span = 30,
    summate = TRUE
  )
  value <- fmrihrf::evaluate(
    reg, times, precision = 0.1, method = "conv"
  )
  if (is.matrix(value)) rowSums(value) else as.numeric(value)
}
Y_grid_contract <- matrix(
  grid_aggregate(grid_target_onsets, fmrihrf::HRF_GAUSSIAN) +
    8 * grid_aggregate(grid_other_onsets, fmrihrf::HRF_SPMG1),
  n_time,
  1L
)
grid_contract_spec <- list(
  sframe = sframe,
  cond = list(
    onsets = grid_target_onsets,
    duration = 0,
    amplitude = 1,
    span = 30
  ),
  others = list(list(
    onsets = grid_other_onsets,
    hrf = fmrihrf::HRF_SPMG1,
    duration = 0,
    amplitude = 1,
    span = 30
  )),
  precision = 0.1,
  method = "conv"
)
selected_spec <- grid_contract_spec
selected_spec$hrf_grid <- list(
  fmrihrf::HRF_SPMG1,
  fmrihrf::HRF_GAUSSIAN
)
gaussian_spec <- grid_contract_spec
gaussian_spec$cond$hrf <- fmrihrf::HRF_GAUSSIAN
grid_contract_options <- list(
  ridge_mode = "absolute",
  ridge_x = 0,
  ridge_b = 0
)
fit_selected_contract <- lss(
  Y_grid_contract,
  X = NULL,
  Z = matrix(1, n_time, 1L),
  method = "oasis",
  oasis = utils::modifyList(
    grid_contract_options,
    list(design_spec = selected_spec)
  )
)
fit_gaussian_contract <- lss(
  Y_grid_contract,
  X = NULL,
  Z = matrix(1, n_time, 1L),
  method = "oasis",
  oasis = utils::modifyList(
    grid_contract_options,
    list(design_spec = gaussian_spec)
  )
)
grid_contract_error <- max(abs(
  unname(fit_selected_contract) - unname(fit_gaussian_contract)
))
stopifnot(grid_contract_error < 1e-10)
```

```{r hrf-grid-contract-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_Gaussian_reference_error = grid_contract_error),
  digits = 12,
  caption = paste(
    "Grid fit versus the fixed-Gaussian reference after conditioning on a",
    "strong SPMG1 other condition."
  )
)
```

Because selection and estimation use the same data, treat this as exploratory
or validate the selection out of sample. `return_se = TRUE` fails closed on
this route. For explicitly normalized voxel-specific shapes, continue to
`vignette("voxel-wise-hrf")`.

## Prewhitening changes the fitted space

Use the top-level `prewhiten` argument. The same estimated operator is applied
to the response, target design, and common design before OASIS fits
coefficients.

```{r prewhiten}
set.seed(20260814)
ar_error <- apply(matrix(rnorm(n_time * n_voxels), n_time), 2, function(x) {
  as.numeric(stats::filter(x, filter = 0.6, method = "recursive"))
})
Y_ar <- mean_signal + ar_error

pw_ar <- prewhiten_options(method = "ar", p = 1, pooling = "global")

fit_ar <- lss(
  Y_ar,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0
  ),
  prewhiten = pw_ar
)
dim(fit_ar)
```

```{r prewhiten-contract, include = FALSE}
fit_ar_naive <- lss(
  Y_ar,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "naive",
  prewhiten = pw_ar
)
prewhiten_backend_error <- max(abs(unname(fit_ar) - unname(fit_ar_naive)))

se_after_whitening <- try(lss(
  Y_ar,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  ),
  prewhiten = pw_ar
), silent = TRUE)

stopifnot(
  prewhiten_backend_error < 1e-10,
  inherits(se_after_whitening, "try-error"),
  grepl(
    "not calibrated for estimated prewhitening",
    as.character(se_after_whitening),
    fixed = TRUE
  )
)
```

```{r prewhiten-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_backend_error = prewhiten_backend_error),
  digits = 12,
  caption = "Unpenalized OASIS versus the naive backend after the same estimated whitening."
)
```

This demonstrates the coefficient path, not that AR(1) is adequate for every
dataset. Choose the order and pooling from residual diagnostics and the study
design. A shared OASIS design supports global or run-level operators; voxel-
and parcel-specific operators are rejected because they cannot be applied to
one shared design matrix.

```{r run-aware-whitening}
run_lengths <- c(90L, 90L)
run_id <- rep(seq_along(run_lengths), times = run_lengths)
runwise <- prewhiten_options(
  method = "ar",
  p = 1,
  pooling = "run",
  runs = run_id
)
length(run_id)
```

Always provide run labels when data span multiple runs so filtering does not
cross run boundaries. `pooling = "global"` with `runs` uses one shared set of
noise coefficients while still respecting those boundaries; `pooling = "run"`
estimates a separate set for each run. Estimated-prewhitening uncertainty is
not calibrated, so this coefficient route cannot also request OASIS model SEs.

## A compact decision guide

- Use a standard LSS backend for a fixed one-basis `X` when you want the
  leanest unpenalized path.
- Use OASIS for event construction, modeled other conditions, multiple bases,
  ridge, blocked products, or fixed-design model SEs.
- Treat ridge, HRF selection, and estimated whitening as changes to the
  estimation contract, not harmless speed options.
- Preserve row names and trial/basis identity whenever $K>1$.

## Next steps

- `vignette("oasis_theory")` --- derivation, complexity, memory, and inference
  boundaries
- `vignette("lss_with_fmridesign")` --- run-aware event-table construction
- `vignette("voxel-wise-hrf")` --- normalized voxel-specific HRF shapes
- `vignette("sbhm")` --- library-constrained voxel-specific HRFs

## Reference

Mumford, J. A., Turner, B. O., Ashby, F. G., & Poldrack, R. A. (2012).
Deconvolving BOLD activation in event-related designs for multivoxel pattern
classification analyses. *NeuroImage*, 59(3), 2636--2643.
[https://doi.org/10.1016/j.neuroimage.2011.08.076](https://doi.org/10.1016/j.neuroimage.2011.08.076)
