---
title: 'OASIS Theory: What Is Reused, Solved, and Returned'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{OASIS Theory: What Is Reused, Solved, and Returned}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
library(fmrilss)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

OASIS is an algebraic implementation of least-squares-separate (LSS). It
does not define a different unpenalized estimator. It identifies the work that
all trial-wise LSS models share, computes that work once, and solves the small
trial-specific systems in batches. With ridge penalties, it instead computes a
penalized LSS estimator.

This article establishes the estimand, derives the one- and multi-basis
systems, and maps each mathematical object to the current implementation. Read
it after `vignette("fmrilss")` and `vignette("oasis_method")`; those articles
introduce the practical meaning of LSS and the user-facing OASIS API.

## Scope and notation

The dimensions below are used throughout.

```{r notation-table, echo = FALSE}
notation <- data.frame(
  Symbol = c("T", "V", "N", "K", "K_z", "p", "B"),
  Meaning = c(
    "time points",
    "response columns or voxels",
    "trials in the condition being estimated",
    "HRF basis columns per trial",
    "columns supplied in the common nuisance design",
    "rank of the common nuisance design; p <= K_z",
    "voxel block size used for data products"
  )
)
knitr::kable(
  notation,
  align = c("c", "l"),
  caption = "Symbols and dimensions used in the OASIS derivation."
)
```

Let $Y \in \mathbb{R}^{T \times V}$ be the data and
$C \in \mathbb{R}^{T \times K_z}$ the common design: intercepts,
drifts, nuisance variables, and aggregates for conditions not being estimated.
Only the column space of $C$ matters. If $Q \in \mathbb{R}^{T \times p}$
is an orthonormal basis for that space, then

\[
R = I_T - QQ^\mathsf{T}
\]

removes it. The implementation rank-reduces $C$ before passing it to the
compiled kernels, so redundant nuisance columns do not add arbitrary projection
directions.

For trial $j$, let $A_j \in \mathbb{R}^{T \times K}$ contain its HRF
basis columns and define

\[
B_j = \sum_{i \ne j} A_i.
\]

After applying the Frisch--Waugh--Lovell reduction, the trial-wise model is

\[
RY = RA_j\,\boldsymbol{\beta}_j + RB_j\,\boldsymbol{\gamma}_j + R\varepsilon_j,
\]

where $\boldsymbol{\beta}_j,\boldsymbol{\gamma}_j \in \mathbb{R}^{K \times V}$.
OASIS returns $\boldsymbol{\beta}_j$. For $K=1$, each trial contributes one
scalar coefficient per voxel. For $K>1$, the $K$ returned rows are basis coefficients; they are
not automatically a single response-amplitude estimate.

## One-basis system

Write $a_j=Rx_j$, $b_j=\sum_{i\ne j}a_i$, and $s=\sum_i a_i$.
The reusable design scalars are

\[
d_j=a_j^\mathsf{T}a_j,\qquad
c_j=a_j^\mathsf{T}b_j,\qquad
e_j=b_j^\mathsf{T}b_j.
\]

The source names these quantities `d`, `alpha`, and `s`, respectively. For all
voxels at once, define $n_{1j}=a_j^\mathsf{T}RY$ and
$n_{2j}=b_j^\mathsf{T}RY=s^\mathsf{T}RY-n_{1j}$. The penalized normal
equations are

\[
\begin{bmatrix}
d_j+\lambda_x & c_j \\
c_j & e_j+\lambda_b
\end{bmatrix}
\begin{bmatrix}\beta_j\\\gamma_j\end{bmatrix}
=
\begin{bmatrix}n_{1j}\\n_{2j}\end{bmatrix}.
\]

Here each right-hand-side row has length $V$. Inverting the $2\times2$
matrix gives

\[
\widehat\beta_j =
\frac{(e_j+\lambda_b)n_{1j}-c_jn_{2j}}
{(d_j+\lambda_x)(e_j+\lambda_b)-c_j^2}.
\]

When both penalties are zero and the two-column trial model has full rank, this
is exactly the ordinary LSS coefficient. A rank-deficient unpenalized model is
not identifiable, and the public OASIS path fails rather than silently adding
jitter. This $2\times2$ form assumes $N>1$; with one trial there is no
“other trials” column, and OASIS solves the one-column target model instead.

## Multi-basis system

For $K>1$, define the residualized blocks
$\widetilde A_j=RA_j$, $\widetilde S=\sum_j\widetilde A_j$, and
$\widetilde B_j=\widetilde S-\widetilde A_j$. The cached $K\times K$ blocks are

\[
D_j=\widetilde A_j^\mathsf{T}\widetilde A_j,\qquad
C_j=\widetilde A_j^\mathsf{T}\widetilde B_j,\qquad
E_j=\widetilde B_j^\mathsf{T}\widetilde B_j.
\]

With $N_{1j}=\widetilde A_j^\mathsf{T}RY$ and
$N_{2j}=\widetilde S^\mathsf{T}RY-N_{1j}$, OASIS solves

\[
\begin{bmatrix}
D_j+\lambda_xI_K & C_j \\
C_j^\mathsf{T} & E_j+\lambda_bI_K
\end{bmatrix}
\begin{bmatrix}\boldsymbol{\beta}_j\\\boldsymbol{\gamma}_j\end{bmatrix}
=
\begin{bmatrix}N_{1j}\\N_{2j}\end{bmatrix}.
\]

The implementation factorizes one $2K\times2K$ Gram matrix per trial and
uses all $V$ voxel columns as right-hand sides. The unpenalized public path
requires every such Gram matrix to have rank $2K$. For $N=1$, the corresponding
target-only system is $K\times K$.

## An executable equality check

The strongest check of the algebra is not agreement between two OASIS helpers;
it is agreement with independently assembled trial-wise GLMs. The following
fixed example checks every trial, basis coefficient, and voxel for both
$K=1$ and $K=3$.

```{r oracle-data, include = FALSE}
set.seed(20260810)

T <- 96
V <- 5
N <- 6
t <- seq_len(T)
Z <- cbind(
  intercept = 1,
  linear_trend = (t - mean(t)) / T
)

direct_lss_oracle <- function(Y, X, Z, K) {
  n_trials <- ncol(X) %/% K
  beta <- matrix(NA_real_, n_trials * K, ncol(Y))
  se <- matrix(NA_real_, n_trials * K, ncol(Y))

  trial_columns <- split(seq_len(ncol(X)), rep(seq_len(n_trials), each = K))
  for (j in seq_len(n_trials)) {
    Aj <- X[, trial_columns[[j]], drop = FALSE]
    Bj <- Reduce(
      `+`,
      lapply(setdiff(seq_len(n_trials), j), function(i) {
        X[, trial_columns[[i]], drop = FALSE]
      })
    )
    design_j <- cbind(Aj, Bj, Z)
    gram_inv <- solve(crossprod(design_j))
    coef_j <- gram_inv %*% crossprod(design_j, Y)
    residual_j <- Y - design_j %*% coef_j
    sigma2_j <- colSums(residual_j^2) / (nrow(Y) - ncol(design_j))
    rows_j <- trial_columns[[j]]
    beta[rows_j, ] <- coef_j[seq_len(K), , drop = FALSE]
    se[rows_j, ] <- sqrt(diag(gram_inv)[seq_len(K)] %o% sigma2_j)
  }
  list(beta = beta, se = se)
}

X1 <- matrix(rnorm(T * N), T, N,
             dimnames = list(NULL, sprintf("trial_%02d", seq_len(N))))
theta1 <- matrix(rnorm(V), 1L, V)
Y1 <- matrix(rowSums(X1), T, 1L) %*% theta1 +
  Z %*% matrix(rnorm(ncol(Z) * V), ncol(Z), V) +
  matrix(rnorm(T * V), T, V)

K <- 3
X3 <- matrix(rnorm(T * N * K), T, N * K)
colnames(X3) <- sprintf(
  "trial_%02d_basis_%d",
  rep(seq_len(N), each = K),
  rep(seq_len(K), times = N)
)
map3 <- data.frame(
  column = colnames(X3),
  trial = rep(seq_len(N), each = K),
  basis = rep(seq_len(K), times = N)
)
trial_columns3 <- split(seq_len(ncol(X3)), rep(seq_len(N), each = K))
S3 <- Reduce(`+`, lapply(trial_columns3, function(idx) {
  X3[, idx, drop = FALSE]
}))
theta3 <- matrix(rnorm(K * V), K, V)
Y3 <- S3 %*% theta3 +
  Z %*% matrix(rnorm(ncol(Z) * V), ncol(Z), V) +
  matrix(rnorm(T * V), T, V)
```

The exact public call for ordinary, unpenalized LSS is short; the zeros are
important because the OASIS default is fractionally penalized.

```{r oasis-public-call}
unpenalized <- oasis_options(
  ridge_mode = "absolute",
  ridge_x = 0,
  ridge_b = 0,
  return_se = TRUE
)
fit1 <- lss(Y1, X1, Z = Z, method = "oasis", oasis = unpenalized)
dim(fit1$beta)
```

```{r oracle-computation, include = FALSE}
ref1 <- direct_lss_oracle(Y1, X1, Z, K = 1)
fit3 <- lss(
  Y3, X3, Z = Z, method = "oasis",
  oasis = oasis_options(
    K = K,
    ntrials = N,
    trial_basis_map = map3,
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  )
)
ref3 <- direct_lss_oracle(Y3, X3, Z, K = K)

oracle_error <- data.frame(
  Basis_dimension = c(1, 3),
  Maximum_beta_error = c(
    max(abs(unname(fit1$beta) - ref1$beta)),
    max(abs(unname(fit3$beta) - ref3$beta))
  ),
  Maximum_SE_error = c(
    max(abs(unname(fit1$se) - ref1$se)),
    max(abs(unname(fit3$se) - ref3$se))
  )
)
```

```{r oracle-summary, echo = FALSE}
knitr::kable(
  oracle_error,
  digits = 12,
  caption = "Maximum discrepancy from independently assembled trial-wise GLMs."
)
```

```{r oasis-oracle-contract, include = FALSE}
stopifnot(
  max(oracle_error$Maximum_beta_error) < 1e-10,
  max(oracle_error$Maximum_SE_error) < 1e-10
)
```

These checks use a correctly specified fixed design with common coefficients
for the summed trial signal and independent Gaussian errors. They establish
implementation equality with the corresponding GLMs. They do not establish
uncertainty for ridge, estimated whitening, or data-adaptive HRF selection.

## Ridge changes the estimator

For each trial and voxel, ridge minimizes

\[
\lVert RY-RA_j\boldsymbol{\beta}_j-RB_j\boldsymbol{\gamma}_j\rVert_F^2
+\lambda_x\lVert\boldsymbol{\beta}_j\rVert_F^2
+\lambda_b\lVert\boldsymbol{\gamma}_j\rVert_F^2.
\]

With `ridge_mode = "absolute"`, `ridge_x` and `ridge_b` are the penalties in
that expression. With `ridge_mode = "fractional"`, the implementation
multiplies them by mean residualized design energies:

- for $K=1$, the means of $d_j$ and $e_j$;
- for $K>1$, the mean diagonal entries of $D_j$ and $E_j$, averaged over
  trials.

Fractional scaling preserves the intended relative penalty under a common
rescaling of the whole trial design. It does not make arbitrary, separate
rescalings of multi-basis columns equivalent. Ridge can stabilize a poorly
conditioned system, but its coefficients are penalized estimates rather than
ordinary LSS coefficients.

The package default is fractional ridge with `ridge_x = ridge_b = 0.05`.
Request zero penalties explicitly when exact unpenalized LSS is the target.
The public path checks that each penalized multi-trial Gram matrix is positive
definite before solving it. When $N=1$, there is no identifiable “other trials”
coefficient, so OASIS solves only the target block; `ridge_b` is then irrelevant.

## What the standard errors mean

OASIS returns model-based standard errors only when all of the following hold:

- `ridge_x = ridge_b = 0`;
- for $N>1$, the trial-specific $2K$-column model is full rank; for $N=1$,
  the target $K$-column model is full rank;
- prewhitening is not estimated in the same call;
- the chosen HRF design is treated as fixed;
- conditional errors have spherical covariance,
  $\operatorname{Var}(\varepsilon_{\cdot v}\mid A_j,B_j,C)=\sigma_{jv}^2I_T$
  for each voxel $v$.

For trial $j$, let $G_j$ be its identifiable unpenalized model Gram matrix:
$2K\times2K$ when $N>1$ and $K\times K$ when $N=1$.
The implementation computes

\[
\widehat\sigma^2_{jv}=\frac{\operatorname{SSE}_{jv}}
{T-p-\operatorname{rank}(G_j)}
\]

and takes the relevant diagonal of
$\widehat\sigma^2_{jv}G_j^{-1}$. Thus the reported values are the same
conditional, homoskedastic-model standard errors as in the corresponding full
GLM under uncorrelated, constant-variance temporal errors. Gaussianity is an
additional requirement for exact finite-sample t inference. The values are not
ridge standard errors, not heteroskedasticity- or autocorrelation-robust errors,
and not calibrated for uncertainty introduced by estimating a whitening model
or selecting an HRF from the same data. `oasis_options()` rejects ridge together
with `return_se = TRUE`; the backend also rejects estimated prewhitening and
voxel-adaptive HRF modes for this request.

## Prewhitening belongs outside `oasis=`

Temporal whitening is handled by the top-level `prewhiten` argument and the
shared `fmriAR` integration. The same estimated linear transformation is
applied to $Y$, the trial design, and the common design before the OASIS
algebra runs.

```{r prewhiten-syntax, eval = FALSE}
fit_ar1 <- lss(
  Y,
  X,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(),
  prewhiten = prewhiten_options(method = "ar", p = 1)
)
```

For multiple runs, supply `pooling = "run"` and the run labels. The legacy
`oasis$whiten` field is deprecated, ignored, and emits a note; it must not be
used to describe the current algorithm. Coefficients remain available after
estimated whitening, but `return_se = TRUE` fails closed because feasible-GLS
uncertainty is not calibrated here.

## Design construction and identity

There are two supported routes into the same solver:

1. `X` supplies an already convolved trial design.
2. `oasis$design_spec` asks the backend to construct one with `fmrihrf`.

The event route preserves event duration and amplitude and adds aggregates for
other conditions to the common nuisance span. In multi-run low-level design
specifications, onsets are global seconds; `lss_design()` with a
`fmridesign::event_model()` is the safer run-aware interface.

A raw multi-basis `X` cannot reveal trial identity from dimensions alone. It
therefore requires explicit `K`, `ntrials`, and a `trial_basis_map` identifying
every uniquely named column. The output contains $NK$ rows in canonical
trial-by-basis order. Supplying an incorrect grouping would change the
estimand, so ambiguous inference is rejected.

If `design_spec$hrf_grid` is supplied, the implementation chooses a candidate
using the observed data before fitting. That is a model-selection step. Treat
the resulting coefficients as conditional on the selected design. Because
ordinary post-selection standard errors are not established by the fixed-design
formula above, `return_se = TRUE` fails closed on this route.

```{r boundary-contracts, include = FALSE}
capture_error <- function(expr) {
  tryCatch({
    force(expr)
    NA_character_
  }, error = function(e) conditionMessage(e))
}

default_fit <- lss(Y1, X1, Z = Z, method = "oasis", oasis = oasis_options())
diag1 <- lss(
  Y1, X1, Z = Z, method = "oasis",
  oasis = oasis_options(return_diag = TRUE)
)
diag3 <- lss(
  Y3, X3, Z = Z, method = "oasis",
  oasis = oasis_options(
    K = K, ntrials = N, trial_basis_map = map3, return_diag = TRUE
  )
)

permutation <- sample(seq_len(ncol(X3)))
fit3_permuted <- lss(
  Y3, X3[, permutation, drop = FALSE], Z = Z, method = "oasis",
  oasis = oasis_options(
    K = K, ntrials = N, trial_basis_map = map3,
    ridge_mode = "absolute", ridge_x = 0, ridge_b = 0
  )
)

sframe <- fmrihrf::sampling_frame(blocklens = T, TR = 1)
event_onsets <- c(10, 30, 50, 70)
event_fit3 <- lss(
  Y1, X = NULL, method = "oasis",
  oasis = oasis_options(design_spec = list(
    sframe = sframe,
    cond = list(onsets = event_onsets, hrf = fmrihrf::HRF_SPMG3, span = 20)
  ))
)

missing_map_error <- capture_error(lss(
  Y3, X3, Z = Z, method = "oasis",
  oasis = list(K = K, ntrials = N)
))
ridge_se_error <- capture_error(oasis_options(return_se = TRUE))
whiten_se_error <- capture_error(lss(
  Y1, X1, Z = Z, method = "oasis", oasis = unpenalized,
  prewhiten = prewhiten_options(method = "ar", p = 1)
))
rank_deficient_error <- capture_error(lss(
  Y1,
  cbind(trial_1 = X1[, 1], trial_2 = X1[, 1]),
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute", ridge_x = 0, ridge_b = 0
  )
))
grid_se_error <- capture_error(lss(
  Y1, X = NULL, method = "oasis",
  oasis = list(
    design_spec = list(hrf_grid = list(fmrihrf::HRF_SPMG1)),
    ridge_mode = "absolute", ridge_x = 0, ridge_b = 0,
    return_se = TRUE
  )
))

scale_errors <- vapply(c(1e-6, 1, 1e6), function(scale) {
  scaled_fit <- lss(
    Y1, scale * X1, Z = Z, method = "oasis", oasis = unpenalized
  )
  scaled_ref <- direct_lss_oracle(Y1, scale * X1, Z, K = 1)
  max(abs(unname(scaled_fit$beta) - scaled_ref$beta) /
        pmax(abs(scaled_ref$beta), 1e-12))
}, numeric(1))

stopifnot(
  is.matrix(default_fit),
  all(dim(default_fit) == c(N, V)),
  identical(rownames(default_fit), colnames(X1)),
  identical(names(diag1), c("beta", "diag")),
  identical(names(diag1$diag), c("d", "alpha", "s")),
  all(vapply(diag1$diag, length, integer(1)) == N),
  all(dim(diag3$beta) == c(N * K, V)),
  all(dim(diag3$diag$D) == c(K, K, N)),
  all(dim(diag3$diag$C) == c(K, K, N)),
  all(dim(diag3$diag$E) == c(K, K, N)),
  max(abs(as.numeric(fit3_permuted) - as.numeric(fit3$beta))) < 1e-10,
  all(dim(event_fit3) == c(length(event_onsets) * K, V)),
  grepl("trial_basis_map", missing_map_error, fixed = TRUE),
  grepl("return_se requires", ridge_se_error, fixed = TRUE),
  grepl("not calibrated for estimated prewhitening", whiten_se_error, fixed = TRUE),
  grepl("rank-deficient trial model", rank_deficient_error, fixed = TRUE),
  grepl("data-adaptive HRF grid selection", grid_se_error, fixed = TRUE),
  max(scale_errors) < 1e-8
)
```

Hidden executable contracts also bind output shapes, diagnostic shapes,
permutation-safe multi-basis identity, event-built multi-basis dimensions,
scale equivariance, and the principal fail-closed inference boundaries.

## Work and memory inventory

The former shorthand “one $T\times T$ QR instead of $N$ such QRs” is not
the relevant comparison. An LSS design has $p+2K$ effective columns, not
$T$ columns. The table below exposes every leading term instead. It assumes
$K_z\le T$, a dense design, and a nuisance rank $p$.

```{r complexity-table, echo = FALSE}
complexity <- data.frame(
  Stage = c(
    "Rank-reveal common design",
    "Residualize all trial columns",
    "Build trial Gram terms",
    "Residualize all voxel data",
    "Trial-data cross-products",
    "Solve all trial systems"
  ),
  `K = 1` = c(
    "O(T K_z^2)",
    "O(T p N)",
    "O(T N)",
    "O(T p V)",
    "O(T N V)",
    "O(N V)"
  ),
  `general K` = c(
    "O(T K_z^2)",
    "O(T p N K)",
    "O(T N K^2)",
    "O(T p V)",
    "O(T N K V)",
    "O(N K^3 + N K^2 V)"
  ),
  check.names = FALSE
)
knitr::kable(
  complexity,
  align = c("l", "c", "c"),
  caption = "Leading arithmetic terms in the current dense OASIS implementation."
)
```

The $O(TNKV)$ cross-product is usually the largest OASIS term when both the
trial and voxel counts are large. `block_cols = B` limits the temporary voxel
block to $O(TB)$, but it does not remove the $NK\times V$ products or the
$NK\times V$ output. Other persistent storage includes $O(Tp)$ for the
nuisance basis, $O(TNK)$ for the residualized trial design, and
$O(NK^2)$ for multi-basis Gram blocks.

A direct implementation that refits each trial model separately performs
$N$ factorizations of $T\times(p+2K)$ designs and repeatedly forms
trial-specific products with $Y$. OASIS reuses the nuisance projection,
aggregate design, Gram terms, and batched products. That structural reuse is
exact; a universal wall-clock speedup is not. Runtime depends on $T,N,V,K,p$,
BLAS, memory bandwidth, block size, and the competing implementation, so this
article makes no synthetic timing claim.

## Implementation map

These names are internal implementation landmarks, not additional public APIs.

```{r implementation-map, echo = FALSE}
implementation <- data.frame(
  Responsibility = c(
    "Validate options, assemble common span, dispatch",
    "Build event-based trial and other-condition designs",
    "Resolve absolute or fractional ridge",
    "One-basis design cache and blocked products",
    "One-basis coefficient solve",
    "Multi-basis design cache and blocked products",
    "Multi-basis coefficient and SE solves"
  ),
  Location = c(
    "R/oasis_backend.R: .lss_oasis",
    "R/oasis_design.R: .oasis_build_X_from_events",
    "R/oasis_ridge_se.R: .oasis_resolve_ridge",
    "src/oasis_core.cpp: oasis_precompute_design, oasis_AtY_SY_blocked",
    "src/oasis_core.cpp: oasis_betas_closed_form",
    "src/oasis_core.cpp: oasisk_precompute_design, oasisk_products",
    "src/oasis_core.cpp: oasisk_betas, oasisk_betas_se"
  )
)
knitr::kable(
  implementation,
  align = c("l", "l"),
  caption = "Internal implementation landmarks for each algebraic stage."
)
```

By default, the public result is an $NK\times V$ coefficient matrix (an
$N\times V$ matrix when $K=1$). Setting `return_se = TRUE` or
`return_diag = TRUE` changes the result to `list(beta, diag?, se?)`; `beta` and
`se` have the same $NK\times V$ shape. For $K=1$, `d`, `alpha`, and `s` are
length-$N$ diagnostic vectors. For $K>1$, `D`, `C`, and `E` are
$K\times K\times N$ arrays. These are unpenalized cross-products on the
residualized design scale, after any requested whitening. They can expose low
energy or collinearity, but they are not a condition-number report and do not
by themselves validate a chosen ridge.

## Boundaries of the result

- Exact equality with classical LSS requires zero ridge, a fixed design, and
  full-rank trial models.
- Multi-basis output is a vector of basis coefficients per trial and voxel;
  any scalar amplitude needs an explicit, identified normalization rule.
- Estimated whitening changes the fitted space and adds uncertainty not covered
  by the conditional standard-error formula.
- HRF grid selection and voxel-adaptive HRFs are data-adaptive procedures;
  fixed-design uncertainty does not automatically survive selection.
- Blocking controls temporary memory. It cannot make storage smaller than the
  requested $NK\times V$ result.

## Next steps

- Continue to `vignette("lss_with_fmridesign")` for run-aware event-table
  construction and trial/basis identity.
- For a backward reference, `vignette("oasis_method")` contains the practical
  fitting and diagnostics workflow, and `vignette("fmrilss")` contains the
  foundational LSS introduction.
- Later articles cover explicitly normalized voxel-specific HRF shapes in
  `vignette("voxel-wise-hrf")` and library-constrained shapes and trial
  coefficients in `vignette("sbhm")`.

## Reference

Mumford, J. A., Turner, B. O., Ashby, F. G., & Poldrack, R. A. (2012).
Deconvolving BOLD activation in event-related designs for multivoxel pattern
classification analyses. *NeuroImage*, 59(3), 2636--2643.
[https://doi.org/10.1016/j.neuroimage.2011.08.076](https://doi.org/10.1016/j.neuroimage.2011.08.076)
