---
title: 'Practical OASIS: Design, Regularization, and Diagnostics'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Practical OASIS: Design, Regularization, and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
library(fmrilss)
library(fmrihrf)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

OASIS is an `lss()` backend that obtains all the trial-wise fits from shared
matrix products instead of fitting a separate GLM for each trial. Use it when
trial-wise estimation needs more than a fixed one-basis design: building the
design from event onsets, multiple HRF bases, ridge regularization, processing
voxels in blocks, or model-based standard errors. This guide builds on
`vignette("fmrilss")` and uses only the public `lss()` interface; the derivation
is in `vignette("oasis_theory")`.

## Decide what you are estimating before fitting

Some OASIS settings change what the coefficients estimate, or whether standard
errors (SEs) are available.

```{r contract-table, echo = FALSE}
contracts <- data.frame(
  Goal = c(
    "Exact ordinary LSS",
    "Regularized coefficients",
    "Model-based coefficient SEs",
    "Multiple HRF bases"
  ),
  Required_choice = c(
    "ridge_x = ridge_b = 0; absolute mode makes the zero scale explicit",
    "absolute or fractional ridge; default is fractional 0.05",
    "zero ridge, fixed full-rank design, no estimated prewhitening",
    "interpret K coefficients per trial; do not call them one amplitude"
  )
)
knitr::kable(
  contracts,
  align = c("l", "l"),
  caption = "OASIS choices that change the estimand or the available uncertainty."
)
```

Calling `oasis_options()` with no arguments gives a penalized fit, so it does
not reproduce an unpenalized LSS backend exactly. Likewise, `return_se = TRUE`
is intentionally unavailable when the fit uses ridge, estimated prewhitening,
HRF-grid selection, or voxel-adaptive HRFs.

## Build a two-condition example

The simulated data have one target condition (A), whose trials each get their
own coefficient, and one other condition (B), which is modeled but not
estimated trial by trial. The model also includes an intercept and a linear
drift. The true coefficients of the target trials vary from trial to trial,
and the noise is iid with a fixed standard deviation. Because the noise-free
signal (`mean_signal`) is kept separate from the noise, the ridge section below
can calculate the bias--variance trade-off exactly for this design.

```{r event-spec}
set.seed(20260811)

n_time <- 180
n_voxels <- 6
TR <- 1
sframe <- fmrihrf::sampling_frame(blocklens = n_time, TR = TR)
times <- fmrihrf::samples(sframe, global = TRUE)
# Canonical HRF scaled to a unit peak over a 30 s window, so coefficients are
# on the peak-response scale.
hrf_unit <- fmrihrf::gen_hrf(fmrihrf::HRF_SPMG1, normalize = TRUE, span = 30)

onsets_a <- c(10, 20, 31, 43, 56, 68, 81, 95, 108, 121, 135, 149)
onsets_b <- c(15, 37, 61, 86, 112, 141)
n_trials <- length(onsets_a)

design_spec <- list(
  sframe = sframe,
  cond = list(
    onsets = onsets_a,
    hrf = hrf_unit,
    duration = 0,
    amplitude = 1,
    span = 30
  ),
  others = list(list(
    onsets = onsets_b,
    hrf = hrf_unit,
    duration = 0,
    amplitude = 1,
    span = 30
  )),
  precision = 0.1,
  method = "conv"
)
```

To simulate the data and to build independent reference fits, a hidden chunk
also evaluates the same trial and condition-B regressors directly as matrices
(`X_trials` and `X_other`). In an analysis, `Y` would normally be the observed
time-by-voxel response rather than a simulated matrix.

```{r design-evaluation, include = FALSE}

trial_set <- fmrihrf::regressor_set(
  onsets = onsets_a,
  fac = factor(seq_len(n_trials)),
  hrf = hrf_unit,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_trials <- fmrihrf::evaluate(
  trial_set, grid = times, precision = 0.1, method = "conv"
)
X_trials <- as.matrix(X_trials)
colnames(X_trials) <- sprintf("trial_%02d", seq_len(n_trials))

other_reg <- fmrihrf::regressor(
  onsets = onsets_b,
  hrf = hrf_unit,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_other <- fmrihrf::evaluate(
  other_reg, grid = times, precision = 0.1, method = "conv"
)
X_other <- matrix(as.numeric(X_other), n_time, 1L,
                  dimnames = list(NULL, "condition_B"))

Z <- cbind(
  intercept = 1,
  linear_trend = seq(-1, 1, length.out = n_time)
)
```

```{r simulate-response}
true_beta <- matrix(
  rnorm(n_trials * n_voxels, mean = 1, sd = 0.35),
  n_trials,
  n_voxels
)
other_beta <- matrix(seq(0.4, 0.9, length.out = n_voxels), 1L)
fixed_beta <- rbind(
  seq(-0.3, 0.3, length.out = n_voxels),
  seq(0.2, -0.2, length.out = n_voxels)
)
mean_signal <- X_trials %*% true_beta +
  X_other %*% other_beta +
  Z %*% fixed_beta
noise_sd <- 1.5
Y <- mean_signal + matrix(rnorm(n_time * n_voxels, sd = noise_sd),
                          n_time, n_voxels)
```

```{r dimensions, echo = FALSE}
dimensions <- data.frame(
  Object = c("Y", "target trial design", "other-condition aggregate", "fixed design"),
  Rows = c(nrow(Y), nrow(X_trials), nrow(X_other), nrow(Z)),
  Columns = c(ncol(Y), ncol(X_trials), ncol(X_other), ncol(Z))
)
knitr::kable(dimensions, caption = "Objects used in the practical example.")
```

## Fit from events

Put the condition whose trial coefficients you want in `cond`. Put modeled
conditions that should not get trial-wise coefficients in `others`. OASIS
models each of these with one aggregate column per basis function and adds
those columns to the *common span*: the regressors shared by every trial's
model, together with `Z` and `Nuisance`.

```{r basic-fit}
fit_default <- lss(
  Y = Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(design_spec = design_spec)
)
dim(fit_default)
fit_default[1:4, 1:3]
```

Because this design uses one basis function, the result is an `n_trials` by
`n_voxels` matrix. The coefficients are penalized by the default fractional
ridge. The simulated data really contain a condition B response, so `others`
models real signal rather than decorating the call.

## Reproduce ordinary LSS exactly

To match ordinary (unpenalized) LSS, set both ridge parameters to zero
explicitly. To verify the fit, a hidden chunk builds each trial's GLM
independently as `[target trial, other target trials, Z, condition B]`, with
the other target trials summed into one column, and compares the coefficients
for all trials and voxels.

```{r unpenalized-fit}
fit_unpenalized <- lss(
  Y = Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    design_spec = design_spec,
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0
  )
)
```

```{r direct-oracle, include = FALSE}
direct_lss <- matrix(NA_real_, n_trials, n_voxels)
for (j in seq_len(n_trials)) {
  direct_design <- cbind(
    X_trials[, j],
    rowSums(X_trials[, -j, drop = FALSE]),
    Z,
    X_other
  )
  direct_lss[j, ] <- qr.coef(qr(direct_design), Y)[1L, ]
}
direct_error <- max(abs(unname(fit_unpenalized) - direct_lss))
stopifnot(direct_error < 1e-10)
```

```{r direct-oracle-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_absolute_error = direct_error),
  digits = 12,
  caption = "Unpenalized OASIS versus independent trial-wise GLMs."
)
```

For a fixed one-basis `X`, this is the same estimand as the other unpenalized
LSS backends. OASIS is useful here because the event and other-condition design
can be constructed and checked in the same workflow.

## Evaluate ridge by bias, variance, and error against the truth

Ridge reduces sampling variance by accepting bias. A smaller spread of fitted
trials is not, by itself, evidence of improvement. For this fixed design and
truth, the OASIS estimate is linear in `Y`. Fitting the noise-free signal
therefore gives the expected coefficients, and so the bias; fitting the
identity matrix gives the linear operator, and so the variance under the iid
noise model. Both are exact, conditional on this design and truth. A Monte
Carlo simulation checks the resulting MSE.

```{r ridge-analysis, include = FALSE}
ridge_levels <- c(0, 0.01, 0.05)

fit_fractional <- function(response, eta) {
  lss(
    response,
    X_trials,
    Z = Z,
    Nuisance = X_other,
    method = "oasis",
    oasis = oasis_options(
      ridge_mode = "fractional",
      ridge_x = eta,
      ridge_b = eta
    )
  )
}

ridge_metrics <- lapply(ridge_levels, function(eta) {
  expected_beta <- fit_fractional(mean_signal, eta)
  operator <- fit_fractional(diag(n_time), eta)
  bias <- unname(expected_beta) - true_beta
  coefficient_variance <- noise_sd^2 * rowSums(operator^2)
  mean_variance <- mean(coefficient_variance)
  mean_squared_bias <- mean(bias^2)
  data.frame(
    Fractional_ridge = eta,
    Mean_absolute_bias = mean(abs(bias)),
    Mean_squared_bias = mean_squared_bias,
    Mean_variance = mean_variance,
    Mean_MSE = mean_squared_bias + mean_variance,
    RMSE = sqrt(mean_squared_bias + mean_variance)
  )
})
ridge_metrics <- do.call(rbind, ridge_metrics)

set.seed(20260812)
mc_repetitions <- 100
mc_mse <- vapply(ridge_levels, function(eta) {
  squared_error <- replicate(mc_repetitions, {
    noisy_y <- mean_signal + matrix(
      rnorm(n_time * n_voxels, sd = noise_sd), n_time, n_voxels
    )
    mean((unname(fit_fractional(noisy_y, eta)) - true_beta)^2)
  })
  mean(squared_error)
}, numeric(1))
ridge_metrics$Monte_Carlo_MSE <- mc_mse
```

```{r ridge-table, echo = FALSE}
knitr::kable(
  ridge_metrics,
  digits = 4,
  caption = paste0(
    "Exact conditional ridge metrics and a ", mc_repetitions,
    "-repetition check for this design, truth, and noise scale."
  )
)
```

```{r ridge-plot, echo = FALSE, fig.cap="Exact squared bias, variance, and MSE for three fractional-ridge settings in the fixed simulation; all bars use squared-beta units.", fig.alt=sprintf("Grouped bars show mean MSE changing from about %.3f at zero ridge to %.3f at fractional ridge %s as variance falls and squared bias rises.", ridge_metrics$Mean_MSE[1], ridge_metrics$Mean_MSE[nrow(ridge_metrics)], format(ridge_levels[length(ridge_levels)]))}
plot_values <- t(as.matrix(ridge_metrics[, c(
  "Mean_squared_bias", "Mean_variance", "Mean_MSE"
)]))
colnames(plot_values) <- format(ridge_levels, trim = TRUE)
old_par <- par(bg = "white")
barplot(
  plot_values,
  beside = TRUE,
  col = c("#9ecae1", "#3182bd", "#de2d26"),
  border = NA,
  xlab = "Fractional ridge",
  ylab = "Mean squared beta units",
  legend.text = c("Squared bias", "Variance", "MSE"),
  args.legend = list(bty = "n", x = "topleft")
)
par(old_par)
```

```{r ridge-contract, include = FALSE}
stopifnot(
  all(is.finite(as.matrix(ridge_metrics[, -1]))),
  all(ridge_metrics$Mean_MSE >= ridge_metrics$Mean_squared_bias),
  all(ridge_metrics$Mean_MSE >= ridge_metrics$Mean_variance),
  # the Monte Carlo check agrees with the exact values to within 10%
  max(abs(ridge_metrics$Mean_MSE - ridge_metrics$Monte_Carlo_MSE) /
        ridge_metrics$Mean_MSE) < 0.10,
  # ridge trades bias for variance in this design
  ridge_metrics$Mean_variance[3] < ridge_metrics$Mean_variance[1],
  ridge_metrics$Mean_squared_bias[3] > ridge_metrics$Mean_squared_bias[1]
)
```

This is one scenario, not a universal ranking. Choose a ridge policy using a
simulation that matches the planned timing, basis, nuisance structure, signal
scale, and downstream loss. The package does not endorse generic penalty
cutoffs based on the inter-stimulus interval alone. Even the zero-penalty row has a
small bias against the generating trial effects: ordinary LSS gives all other
target trials one shared coefficient, while this simulation gives them
heterogeneous coefficients. The table therefore evaluates the full LSS
estimator against the generating effects, not merely ridge against an
unpenalized LSS reference.

## Inspect design diagnostics

With `return_diag = TRUE`, `lss()` returns a list instead of a matrix, with
the coefficients in `beta` and design diagnostics in `diag`. The diagnostics
are computed after the common span has been projected out of the design. For a
one-basis fit, `diag` holds one value per trial in each of three vectors: `d`,
the energy (sum of squares) of the target-trial regressor; `s`, the energy of
the summed other-trial regressor; and `alpha`, their cross-product. The code
below converts `alpha` to a correlation.

```{r diagnostics}
fit_diag <- lss(
  Y,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(return_diag = TRUE)
)
diagnostics <- with(fit_diag$diag, data.frame(
  target_energy = d,
  other_energy = s,
  target_other_correlation = alpha / sqrt(d * s)
))
head(diagnostics, 4)
```

These are unpenalized cross-products on the residualized design scale. They are
useful for finding low-energy or highly correlated trials, but they are not a
condition number, a ridge calibration, or an inferential test.

## With multiple bases, each trial has K coefficients, not one amplitude

The SPMG3 basis (`fmrihrf::HRF_SPMG3`) is the SPM canonical HRF plus its
temporal and dispersion derivatives. When OASIS builds the design from events
with this basis, the output has three rows per trial: the canonical,
temporal-derivative, and dispersion-derivative coefficients. Their scale
depends on the basis definition.

```{r multi-basis}
spec_spmg3 <- design_spec
spec_spmg3$cond$hrf <- fmrihrf::HRF_SPMG3
spec_spmg3$others[[1]]$hrf <- fmrihrf::HRF_SPMG3

fit_spmg3 <- lss(
  Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(design_spec = spec_spmg3, return_diag = TRUE)
)
K_spmg3 <- 3L
c(rows = nrow(fit_spmg3$beta), trials = n_trials, basis_dimension = K_spmg3)
rownames(fit_spmg3$beta)[1:6]
```

The output has `NK` rows (N trials times K basis functions) and `V` columns
(voxels). Rows are in OASIS's standard trial-major, basis-minor order: all K
coefficients for trial 1, then all K for trial 2, and so on. Keep the row names
or reshape explicitly; do not infer a scalar response amplitude from the
canonical-HRF row alone unless a separate normalization and estimand justify
that operation.

If you already have a precomputed multi-basis matrix, you must tell OASIS which
trial and basis function each column belongs to: pass `K`, `ntrials`, and
`trial_basis_map` to `oasis_options()`. The matrix (`X_multi` below) must have unique, non-empty
column names, and the map must assign every column to one exact-integer trial
and basis pair. OASIS reorders the matrix into trial-major, basis-minor order
and attaches the reordered map to the returned beta matrix as the
`trial_basis_map` attribute.

```{r raw-multi-basis-design, include = FALSE}
raw_spmg3_set <- fmrihrf::regressor_set(
  onsets = onsets_a,
  fac = factor(seq_len(n_trials)),
  hrf = fmrihrf::HRF_SPMG3,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_spmg3_source <- as.matrix(fmrihrf::evaluate(
  raw_spmg3_set, grid = times, precision = 0.1, method = "conv"
))
raw_spmg3_other <- fmrihrf::regressor(
  onsets = onsets_b,
  hrf = fmrihrf::HRF_SPMG3,
  duration = 0,
  amplitude = 1,
  span = 30,
  summate = TRUE
)
X_other_spmg3 <- as.matrix(fmrihrf::evaluate(
  raw_spmg3_other, grid = times, precision = 0.1, method = "conv"
))
```

Here we built the source matrix ourselves, so we know it is in trial-major,
basis-minor order. The code records each column's trial and basis *before*
deliberately reversing the column order; the fit should still match the
event-built fit above. For a matrix produced elsewhere, take each column's
trial and basis from the producer's metadata; never infer them from the order
in which an arbitrary matrix happens to arrive.

```{r raw-multi-basis-contract}
K <- K_spmg3
source_map <- data.frame(
  column = sprintf("source_%02d", seq_len(ncol(X_spmg3_source))),
  trial = rep(seq_len(n_trials), each = K),
  basis = rep(seq_len(K), times = n_trials)
)
colnames(X_spmg3_source) <- source_map$column

column_order <- rev(seq_len(ncol(X_spmg3_source)))
X_multi <- X_spmg3_source[, column_order, drop = FALSE]

fit_raw_multi <- lss(
  Y,
  X_multi,
  Z = Z,
  Nuisance = X_other_spmg3,
  method = "oasis",
  oasis = oasis_options(
    K = K,
    ntrials = n_trials,
    trial_basis_map = source_map,
    return_diag = TRUE
  )
)
raw_permutation_error <- max(abs(fit_raw_multi$beta - fit_spmg3$beta))
c(maximum_beta_error = raw_permutation_error)
head(attr(fit_raw_multi$beta, "trial_basis_map"), 3)
```

Without optional returns, OASIS returns the `NK` by `V` beta matrix. Setting
`return_diag = TRUE` or `return_se = TRUE` returns a list containing `beta`,
then `diag` when requested, then `se` when requested. `beta` and `se` are both
`NK` by `V`. For `K = 1`, `diag` contains vectors `d`, `alpha`, and `s`, each of
length `N`. For `K > 1`, it contains `D`, `C`, and `E` arrays of shape
`K` by `K` by `N`. Diagnostics describe the residualized design, or the
whitened-and-residualized design when whitening is active.

```{r multi-basis-contract, include = FALSE}
stopifnot(
  raw_permutation_error < 1e-10,
  all(dim(fit_spmg3$beta) == c(n_trials * K_spmg3, n_voxels)),
  all(dim(fit_spmg3$diag$D) == c(K_spmg3, K_spmg3, n_trials)),
  all(dim(fit_spmg3$diag$C) == c(K_spmg3, K_spmg3, n_trials)),
  all(dim(fit_spmg3$diag$E) == c(K_spmg3, K_spmg3, n_trials))
)
```

## Standard errors require the unpenalized fixed-design model

To check the SEs against a known answer, the example generates data that
exactly satisfy each trial's LSS model: all target trials share a common
coefficient vector, condition B and `Z` are in the fitted common span, and the
temporal errors are iid Gaussian. A hidden chunk compares the OASIS betas and
SEs with those from separately fitted full GLMs.

```{r se-data, include = FALSE}
set.seed(20260813)
common_trial_beta <- matrix(seq(0.6, 1.1, length.out = n_voxels), 1L)
Y_se <- matrix(rowSums(X_trials), n_time, 1L) %*% common_trial_beta +
  X_other %*% other_beta +
  Z %*% fixed_beta +
  matrix(rnorm(n_time * n_voxels, sd = 1), n_time, n_voxels)
```

```{r se-fit}
fit_se <- lss(
  Y_se,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  )
)
dim(fit_se$beta)
dim(fit_se$se)
```

```{r se-oracle, include = FALSE}
direct_beta <- direct_se <- matrix(NA_real_, n_trials, n_voxels)
for (j in seq_len(n_trials)) {
  D <- cbind(
    X_trials[, j],
    rowSums(X_trials[, -j, drop = FALSE]),
    Z,
    X_other
  )
  gram_inv <- solve(crossprod(D))
  coef_j <- gram_inv %*% crossprod(D, Y_se)
  residual_j <- Y_se - D %*% coef_j
  sigma2_j <- colSums(residual_j^2) / (n_time - ncol(D))
  direct_beta[j, ] <- coef_j[1L, ]
  direct_se[j, ] <- sqrt(gram_inv[1L, 1L] * sigma2_j)
}
se_oracle <- data.frame(
  Maximum_beta_error = max(abs(unname(fit_se$beta) - direct_beta)),
  Maximum_SE_error = max(abs(unname(fit_se$se) - direct_se))
)
```

```{r se-result, echo = FALSE}
knitr::kable(
  se_oracle,
  digits = 12,
  caption = "OASIS beta and SE agreement with independent full GLMs."
)
```

```{r se-contract, include = FALSE}
stopifnot(
  all(dim(fit_se$beta) == c(n_trials, n_voxels)),
  all(dim(fit_se$se) == c(n_trials, n_voxels)),
  max(as.matrix(se_oracle)) < 1e-10
)
```

These SEs are model-based, conditional on the design, and on the coefficient
scale. They assume spherical temporal errors (uncorrelated, with equal
variance) and are not robust to autocorrelation or heteroskedasticity. The
result does not come with a multiple-testing or population-inference
procedure, and estimated whitening cannot be combined with `return_se = TRUE`.

## FIR coefficients and their uncertainty

A finite impulse response (FIR) basis gives each post-onset time bin its own
coefficient, so FIR output, like SPMG3 output, has K rows per trial and one
column per voxel. The plot below shows one trial in one voxel, with a model SE
for each coefficient. These are not across-voxel or across-trial SEs for an
average HRF. The simulated data follow the fitted FIR model exactly (target
trial plus summed other trials, with iid Gaussian errors), and a hidden chunk
again checks betas and SEs against an independently built GLM for every trial.

```{r fir-data, include = FALSE}
K_fir <- 6L
fir_hrf <- fmrihrf::hrf_fir_generator(nbasis = K_fir, span = 24)
spec_fir <- design_spec
spec_fir$cond$hrf <- fir_hrf
spec_fir$cond$span <- 24
spec_fir$others[[1]]$hrf <- fir_hrf
spec_fir$others[[1]]$span <- 24

# Generate the oracle design independently through fmrihrf's public API.
fir_trial_set <- fmrihrf::regressor_set(
  onsets = onsets_a,
  fac = factor(seq_len(n_trials)),
  hrf = fir_hrf,
  duration = 0,
  amplitude = 1,
  span = 24,
  summate = TRUE
)
X_fir <- as.matrix(fmrihrf::evaluate(
  fir_trial_set, grid = times, precision = 0.1, method = "conv"
))
fir_other_reg <- fmrihrf::regressor(
  onsets = onsets_b,
  hrf = fir_hrf,
  duration = 0,
  amplitude = 1,
  span = 24,
  summate = TRUE
)
X_other_fir <- as.matrix(fmrihrf::evaluate(
  fir_other_reg, grid = times, precision = 0.1, method = "conv"
))
summed_fir <- vapply(seq_len(K_fir), function(k) {
  rowSums(X_fir[, seq(k, ncol(X_fir), by = K_fir), drop = FALSE])
}, numeric(n_time))
theta_fir <- outer(
  seq(0.8, -0.2, length.out = K_fir),
  seq(0.8, 1.2, length.out = n_voxels)
)
other_theta_fir <- outer(
  seq(0.3, -0.1, length.out = ncol(X_other_fir)),
  seq(0.9, 1.1, length.out = n_voxels)
)
set.seed(20260815)
Y_fir <- summed_fir %*% theta_fir +
  X_other_fir %*% other_theta_fir +
  Z %*% fixed_beta +
  matrix(rnorm(n_time * n_voxels), n_time, n_voxels)
```

```{r fir-fit}
fit_fir <- lss(
  Y_fir,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    design_spec = spec_fir,
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  )
)
dim(fit_fir$beta)
dim(fit_fir$se)
```

```{r fir-oracle, include = FALSE}
# Assemble every corresponding full GLM independently.
direct_fir_beta <- direct_fir_se <- matrix(
  NA_real_, n_trials * K_fir, n_voxels
)
for (j in seq_len(n_trials)) {
  target_rows <- ((j - 1L) * K_fir + 1L):(j * K_fir)
  other_trial_ids <- setdiff(seq_len(n_trials), j)
  other_trials <- vapply(seq_len(K_fir), function(k) {
    columns <- (other_trial_ids - 1L) * K_fir + k
    rowSums(X_fir[, columns, drop = FALSE])
  }, numeric(n_time))
  D <- cbind(
    X_fir[, target_rows, drop = FALSE], other_trials, Z, X_other_fir
  )
  gram_inv <- solve(crossprod(D))
  coef_j <- gram_inv %*% crossprod(D, Y_fir)
  residual_j <- Y_fir - D %*% coef_j
  sigma2_j <- colSums(residual_j^2) / (n_time - ncol(D))
  direct_fir_beta[target_rows, ] <- coef_j[seq_len(K_fir), , drop = FALSE]
  direct_fir_se[target_rows, ] <- sqrt(
    outer(diag(gram_inv)[seq_len(K_fir)], sigma2_j)
  )
}

fir_oracle <- data.frame(
  Maximum_beta_error = max(abs(unname(fit_fir$beta) - direct_fir_beta)),
  Maximum_SE_error = max(abs(unname(fit_fir$se) - direct_fir_se))
)

fir_rows <- seq_len(K_fir)
fir_example <- data.frame(
  basis = fir_rows,
  coefficient = fit_fir$beta[fir_rows, 1L],
  se = fit_fir$se[fir_rows, 1L]
)
```

```{r fir-oracle-result, echo = FALSE}
knitr::kable(
  fir_oracle,
  digits = 12,
  caption = "FIR beta and SE agreement with independent full GLMs."
)
```

```{r fir-plot, echo = FALSE, fig.cap="FIR basis coefficients plus or minus one conditional model SE for trial 1, voxel 1; the y-axis is coefficient scale, not a normalized HRF amplitude.", fig.alt="Six FIR estimates for one trial and voxel: the first is positive, the second is near zero, and the remaining four are negative; one-SE intervals are wide, with the second crossing zero."}
old_par <- par(bg = "white")
plot(
  fir_example$basis,
  fir_example$coefficient,
  type = "b",
  pch = 19,
  xlab = "FIR basis index",
  ylab = "FIR basis coefficient",
  ylim = range(
    fir_example$coefficient - fir_example$se,
    fir_example$coefficient + fir_example$se
  )
)
arrows(
  fir_example$basis,
  fir_example$coefficient - fir_example$se,
  fir_example$basis,
  fir_example$coefficient + fir_example$se,
  angle = 90,
  code = 3,
  length = 0.04
)
abline(h = 0, col = "grey70", lty = 2)
par(old_par)
```

```{r fir-contract, include = FALSE}
stopifnot(
  all(dim(fit_fir$beta) == c(n_trials * K_fir, n_voxels)),
  all(dim(fit_fir$se) == c(n_trials * K_fir, n_voxels)),
  all(is.finite(fir_example$se)),
  all(fir_example$se >= 0),
  max(as.matrix(fir_oracle)) < 1e-10
)
```

FIR models can be poorly conditioned because each trial contributes many
columns. Ridge may be scientifically useful, but it cannot be combined with
`return_se = TRUE`: the package does not report inferential FIR SEs for a
penalized fit.

## HRF-grid selection is exploratory

When `design_spec$hrf_grid` lists candidate HRFs, OASIS chooses one of them
for the target condition from the observed `Y`. It residualizes both the
response and each candidate regressor against the full common span (`Z`,
`Nuisance`, and modeled `others`), scores each candidate with a matched filter,
and fits the design with the winner. The call shown returns the beta matrix. Even when design diagnostics are requested,
the public `lss()` route does not expose the winning candidate or candidate
scores, and it does not return selection-adjusted uncertainty.

```{r hrf-grid}
grid_spec <- design_spec
grid_spec$hrf_grid <- list(fmrihrf::HRF_SPMG1, fmrihrf::HRF_GAMMA)

fit_grid <- lss(
  Y,
  X = NULL,
  Z = Z,
  method = "oasis",
  oasis = oasis_options(
    design_spec = grid_spec,
    ridge_mode = "fractional",
    ridge_x = 0.01,
    ridge_b = 0.01
  )
)
dim(fit_grid)
```

A known-truth check in a hidden chunk tests the residualization step. Its
target signal uses a Gaussian HRF, while a much stronger other-condition signal
uses SPMG1. The grid fit must match an otherwise identical fit in which the
Gaussian target HRF is fixed in advance.

```{r hrf-grid-contract, include = FALSE}
grid_target_onsets <- c(10, 40, 70, 100, 130)
grid_other_onsets <- c(20, 50, 80, 110, 140)
grid_aggregate <- function(onsets, hrf) {
  reg <- fmrihrf::regressor(
    onsets = onsets,
    hrf = hrf,
    duration = 0,
    amplitude = 1,
    span = 30,
    summate = TRUE
  )
  value <- fmrihrf::evaluate(
    reg, times, precision = 0.1, method = "conv"
  )
  if (is.matrix(value)) rowSums(value) else as.numeric(value)
}
Y_grid_contract <- matrix(
  grid_aggregate(grid_target_onsets, fmrihrf::HRF_GAUSSIAN) +
    8 * grid_aggregate(grid_other_onsets, fmrihrf::HRF_SPMG1),
  n_time,
  1L
)
grid_contract_spec <- list(
  sframe = sframe,
  cond = list(
    onsets = grid_target_onsets,
    duration = 0,
    amplitude = 1,
    span = 30
  ),
  others = list(list(
    onsets = grid_other_onsets,
    hrf = fmrihrf::HRF_SPMG1,
    duration = 0,
    amplitude = 1,
    span = 30
  )),
  precision = 0.1,
  method = "conv"
)
selected_spec <- grid_contract_spec
selected_spec$hrf_grid <- list(
  fmrihrf::HRF_SPMG1,
  fmrihrf::HRF_GAUSSIAN
)
gaussian_spec <- grid_contract_spec
gaussian_spec$cond$hrf <- fmrihrf::HRF_GAUSSIAN
grid_contract_options <- list(
  ridge_mode = "absolute",
  ridge_x = 0,
  ridge_b = 0
)
fit_selected_contract <- lss(
  Y_grid_contract,
  X = NULL,
  Z = matrix(1, n_time, 1L),
  method = "oasis",
  oasis = utils::modifyList(
    grid_contract_options,
    list(design_spec = selected_spec)
  )
)
fit_gaussian_contract <- lss(
  Y_grid_contract,
  X = NULL,
  Z = matrix(1, n_time, 1L),
  method = "oasis",
  oasis = utils::modifyList(
    grid_contract_options,
    list(design_spec = gaussian_spec)
  )
)
grid_contract_error <- max(abs(
  unname(fit_selected_contract) - unname(fit_gaussian_contract)
))
stopifnot(grid_contract_error < 1e-10)
```

```{r hrf-grid-contract-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_Gaussian_reference_error = grid_contract_error),
  digits = 12,
  caption = paste(
    "Grid fit versus the fixed-Gaussian reference after conditioning on a",
    "strong SPMG1 other condition."
  )
)
```

Because selection and estimation use the same data, treat this as exploratory
or validate the selection out of sample. On this route, `return_se = TRUE`
raises an error instead of returning SEs. For explicitly normalized voxel-specific shapes, continue to
`vignette("voxel-wise-hrf")`.

## Prewhitening changes the fitted space

To model temporally autocorrelated noise, pass `prewhiten_options()` to the
top-level `prewhiten` argument of `lss()`. The same estimated whitening
operator is applied to the response, target design, and common design before
OASIS fits coefficients.

```{r prewhiten}
set.seed(20260814)
ar_error <- apply(matrix(rnorm(n_time * n_voxels), n_time), 2, function(x) {
  as.numeric(stats::filter(x, filter = 0.6, method = "recursive"))
})
Y_ar <- mean_signal + ar_error

pw_ar <- prewhiten_options(method = "ar", p = 1, pooling = "global")

fit_ar <- lss(
  Y_ar,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0
  ),
  prewhiten = pw_ar
)
dim(fit_ar)
```

```{r prewhiten-contract, include = FALSE}
fit_ar_naive <- lss(
  Y_ar,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "naive",
  prewhiten = pw_ar
)
prewhiten_backend_error <- max(abs(unname(fit_ar) - unname(fit_ar_naive)))

se_after_whitening <- try(lss(
  Y_ar,
  X_trials,
  Z = Z,
  Nuisance = X_other,
  method = "oasis",
  oasis = oasis_options(
    ridge_mode = "absolute",
    ridge_x = 0,
    ridge_b = 0,
    return_se = TRUE
  ),
  prewhiten = pw_ar
), silent = TRUE)

stopifnot(
  prewhiten_backend_error < 1e-10,
  inherits(se_after_whitening, "try-error"),
  grepl(
    "not calibrated for estimated prewhitening",
    as.character(se_after_whitening),
    fixed = TRUE
  )
)
```

```{r prewhiten-result, echo = FALSE}
knitr::kable(
  data.frame(Maximum_backend_error = prewhiten_backend_error),
  digits = 12,
  caption = "Unpenalized OASIS versus the naive backend after the same estimated whitening."
)
```

This shows that OASIS and the naive backend compute the same coefficients
after the same whitening, not that AR(1) is adequate for every dataset. Choose
the order and pooling from residual diagnostics and the study design. Because
OASIS uses one design matrix shared by all voxels, it supports global or
run-level operators; it rejects voxel- and parcel-specific operators because
they cannot be applied to one shared design matrix.

The next chunk only builds run-aware options for two runs of 90 scans; it does
not fit them.

```{r run-aware-whitening}
run_lengths <- c(90L, 90L)
run_id <- rep(seq_along(run_lengths), times = run_lengths)
runwise <- prewhiten_options(
  method = "ar",
  p = 1,
  pooling = "run",
  runs = run_id
)
length(run_id)
```

Always provide run labels when data span multiple runs so filtering does not
cross run boundaries. `pooling = "global"` with `runs` uses one shared set of
noise coefficients while still respecting those boundaries; `pooling = "run"`
estimates a separate set for each run. Estimated-prewhitening uncertainty is
not calibrated, so a whitened OASIS fit cannot also request model SEs.

## When to use OASIS

- Use a standard LSS backend for a fixed one-basis `X` when you want the
  leanest unpenalized path.
- Use OASIS for event construction, modeled other conditions, multiple bases,
  ridge, blocked products, or fixed-design model SEs.
- Treat ridge, HRF selection, and estimated whitening as changes to what is
  estimated and which uncertainty is available, not harmless speed options.
- Preserve row names and trial/basis identity whenever $K>1$.

## Next steps

- `vignette("oasis_theory")` --- derivation, complexity, memory, and inference
  boundaries
- `vignette("lss_with_fmridesign")` --- run-aware event-table construction
- `vignette("voxel-wise-hrf")` --- normalized voxel-specific HRF shapes
- `vignette("sbhm")` --- library-constrained voxel-specific HRFs

## Reference

Mumford, J. A., Turner, B. O., Ashby, F. G., & Poldrack, R. A. (2012).
Deconvolving BOLD activation in event-related designs for multivoxel pattern
classification analyses. *NeuroImage*, 59(3), 2636--2643.
[https://doi.org/10.1016/j.neuroimage.2011.08.076](https://doi.org/10.1016/j.neuroimage.2011.08.076)
