---
title: 'Prewhitening: Where the Noise Model Comes From'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Prewhitening: Where the Noise Model Comes From}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
library(fmrilss)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

fMRI noise is temporally autocorrelated. Prewhitening estimates that
autocorrelation, builds a filter that removes it, and applies the same filter
to the data and to every design matrix, so that ordinary least squares on the
filtered system is generalized least squares on the original one. When the
noise model is right, trial estimates become more precise.

The filter is only as good as the autocorrelation estimate, and that estimate
has to come from *residuals of some fitted model*: the noise itself is never
observed. This article explains why the choice of that model matters a great
deal for single-trial analyses, what the three choices offered by
`prewhiten$residual_model` do, and how to pick one. Read it after
`vignette("fmrilss")`.

## Why the residual model matters

Let \(M = I - D(D^\top D)^{-1}D^\top\) be the residual-forming matrix of a
fitted design \(D\) with \(q\) columns and \(n\) scans. The residuals are
\(r = M y\). If the noise were white with variance \(\sigma^2\), the expected
lag-\(k\) autocovariance of the residuals would be

\[
\mathbb{E}\Big[\tfrac{1}{n}\textstyle\sum_t r_t r_{t+k}\Big] =
\tfrac{\sigma^2}{n}\operatorname{tr}(M L_k),
\]

where \(L_k\) is the lag-\(k\) shift matrix. That is not zero. Fitting \(D\)
absorbs part of the noise into the fitted values, and because HRF-convolved
trial regressors are smooth, the part they absorb is the slowly varying,
positively autocorrelated part. What is left looks rougher than the noise
really is, so the autocorrelation estimate is biased **downward**. The size of
the bias grows with \(q / n\).

That is harmless for a block design with a handful of regressors. It is not
harmless for single-trial models, where the natural "full" design, the
least-squares-all (LSA) model with one column per trial, can have a column for
every two or three scans.

The opposite mistake is to fit too little. If the residual model leaves real
signal in the residuals, for example trial-to-trial amplitude variability, that
smooth signal is mistaken for autocorrelated noise and the estimate is biased
**upward**.

## The three residual models

| `residual_model` | Trial part of the fitted design | Bias of the autocorrelation estimate | Cost | Pooling |
|---|---|---|---|---|
| `"aggregate"` (default) | One summed regressor per `trial_groups` level (or per basis function) | Upward when trial-to-trial variability is large relative to noise | Fast | All |
| `"full"` | Every trial regressor (LSA) | Downward, severe when trials are a sizeable fraction of scans | Fast | All |
| `"corrected"` | Every trial regressor, then fmriAR's residual-bias correction | Approximately unbiased | About \(O(n^2 \times \text{trials})\) | `"global"`, `"run"` |

In every case the confounds (`Z`, `Nuisance`, and an intercept if none is
present) are part of the fitted design.

* **`"aggregate"`** is the model the per-trial GLMs of Nilearn and
  NiBetaSeries effectively estimate their noise from: the condition-level
  response is removed, trial-by-trial deviations from it are not.
* **`"full"`** was the only behaviour before fmrilss 0.2.0.9000. Use it to
  reproduce earlier results.
* **`"corrected"`** fits the full model and then inverts the linear map
  \(\gamma \mapsto \operatorname{tr}(M L_k \Gamma)\) from true to expected
  residual autocovariance (`fmriAR::acvf_bias_matrix()`), so neither the
  over-fitting of the full model nor unmodelled trial variability distorts the
  estimate. fmrilss builds the correction design for you; supplying
  `prewhiten$design` yourself is equivalent and implies `"full"`.

## A simulation

The simulation below draws AR(1) noise with \(\phi = 0.4\), and trial
amplitudes with mean 1 and either modest or large trial-to-trial variability,
in a rapid design (inter-trial interval 1.5--4 s, about one trial per three
scans) and a slow design (8--12 s).

```{r simulate}
canonical_hrf <- function(t) {
  h <- dgamma(t, 6) - dgamma(t, 16) / 6
  h / max(h)
}

simulate_run <- function(iti, trial_sd, n = 300, n_voxels = 200, phi = 0.4,
                         seed = 1) {
  set.seed(seed)
  onsets <- cumsum(c(20, runif(1000, iti[1], iti[2])))
  onsets <- onsets[onsets < n - 25]
  X <- vapply(onsets, function(onset) {
    lag <- seq_len(n) - 1 - onset
    ifelse(lag >= 0, canonical_hrf(pmax(lag, 0)), 0)
  }, numeric(n))
  betas <- matrix(rnorm(ncol(X) * n_voxels, 1, trial_sd), ncol(X), n_voxels)
  innovations <- matrix(rnorm(n * n_voxels, sd = sqrt(1 - phi^2)), n, n_voxels)
  noise <- apply(innovations, 2, function(e) {
    as.numeric(stats::filter(e, phi, method = "recursive"))
  })
  list(Y = X %*% betas + noise, X = X, betas = betas)
}

compare_models <- function(label, run) {
  rmse <- function(estimate) sqrt(mean((estimate - run$betas)^2))
  fit <- function(model) {
    lss(run$Y, run$X,
        prewhiten = list(method = "ar", p = 1, residual_model = model))
  }
  fits <- lapply(c(aggregate = "aggregate", full = "full",
                   corrected = "corrected"), fit)
  phi_hat <- vapply(fits, function(f) attr(f, "whiten_plan")$phi[[1]], 1)
  data.frame(
    design = label,
    trials = ncol(run$X),
    phi_full = phi_hat[["full"]],
    phi_aggregate = phi_hat[["aggregate"]],
    phi_corrected = phi_hat[["corrected"]],
    rmse_ols = rmse(lss(run$Y, run$X)),
    rmse_full = rmse(fits$full),
    rmse_aggregate = rmse(fits$aggregate),
    rmse_corrected = rmse(fits$corrected)
  )
}

scenarios <- rbind(
  compare_models("rapid, trial SD 0.5", simulate_run(c(1.5, 4), 0.5)),
  compare_models("slow, trial SD 0.5", simulate_run(c(8, 12), 0.5)),
  compare_models("rapid, trial SD 2", simulate_run(c(1.5, 4), 2)),
  compare_models("slow, trial SD 2", simulate_run(c(8, 12), 2))
)
```

```{r simulate-table, echo = FALSE}
display <- scenarios
display[-(1:2)] <- lapply(display[-(1:2)], round, digits = 2)
knitr::kable(
  display,
  col.names = c("Design", "Trials", "phi: full", "phi: aggregate",
                "phi: corrected", "RMSE: OLS", "RMSE: full",
                "RMSE: aggregate", "RMSE: corrected"),
  caption = "AR(1) estimates (truth 0.4) and trial-beta RMSE for 300 scans and 200 voxels."
)
```

```{r simulate-check, include = FALSE}
rapid <- scenarios[1, ]
slow_noisy <- scenarios[4, ]
stopifnot(
  rapid$phi_full < 0,
  abs(rapid$phi_aggregate - 0.4) > abs(rapid$phi_corrected - 0.4),
  all(abs(scenarios$phi_corrected - 0.4) < 0.05),
  all(scenarios$phi_full < 0.35),
  slow_noisy$phi_aggregate > 0.6,
  slow_noisy$rmse_aggregate > slow_noisy$rmse_corrected,
  rapid$rmse_full > rapid$rmse_aggregate
)
```

Reading the table:

* **`"full"` is biased whenever the trial count is not small relative to the
  scan count.** In the rapid design (one trial column per three scans) it
  estimates a *negative* \(\phi\) for positively autocorrelated noise, and the
  resulting filter gives betas no better (slightly worse) than no
  prewhitening at all. Even the slow design, with about one column per eleven
  scans, loses a third of \(\phi\).
* **`"aggregate"` works well when trial-to-trial variability is modest
  relative to the noise**, which is the common situation for single-trial
  fMRI: its estimate is close in the slow design and somewhat high in the
  rapid one, and it gives the lowest or equal-lowest RMSE in both. When the
  variability is large, the deviations it leaves in the residuals inflate
  \(\phi\) substantially; in the slow design that over-whitening costs
  accuracy. (In the rapid, high-variability design the over-whitened LSS
  betas happen to have lower error, because LSS already mixes neighbouring
  trials' deviations into each estimate. Treat that as a coincidence of this
  design, not a reason to over-whiten.)
* **`"corrected"` recovers \(\phi\) in every scenario** and is never far from
  the best RMSE in each row.

## Choosing a residual model

1. Start with the default, `"aggregate"`. It is fast, works with every
   pooling mode, including voxel-adaptive pooling, and matches what Nilearn
   does.
2. If you can afford it, compare its estimate with `"corrected"` once on your
   data, for example on a subset of voxels or one run:

    ```{r compare-plans, eval = FALSE}
    est <- function(model) {
      fit <- lss(Y[, voxel_subset], X, Z, Nuisance,
                 prewhiten = list(method = "ar", p = 1, residual_model = model))
      attr(fit, "whiten_plan")$phi
    }
    est("aggregate")
    est("corrected")
    ```

   Close agreement means the default is fine. A clearly larger
   `"aggregate"` estimate means trial-to-trial variability is large relative
   to the noise; prefer `"corrected"` (with global or run pooling) for that
   dataset.
3. Avoid `"full"` except to reproduce earlier results.

The correction's cost grows with the square of the scan count times the
number of design columns, roughly a second for 400 scans and 150 trials and
around ten seconds for 1,200 scans and 300 trials, independent of the voxel
count. It is computed once per fit. The correction is designed for
high-pass-filtered designs; see `correction_max_lag` in `?lss` and
`fmriAR::fit_noise()`.

## How the aggregate columns are chosen

The aggregate model sums trial regressors so that the condition-level response
is removed from the residuals:

| Caller | Aggregate trial columns |
|---|---|
| `lss()` with `trial_groups` | One summed regressor per group |
| `lss()` without `trial_groups` | One regressor summing all trials |
| `lss(method = "oasis")`, multi-basis | One summed regressor per basis function |
| `lss_sbhm()`, SBHM amplitude fits | One summed regressor per shared basis function (plus other conditions) |

Passing `trial_groups` therefore also improves the noise model whenever
conditions differ in their mean response.

## Pooling and the residual model

`residual_model` decides which residuals are used; `pooling` decides how the
autocorrelation is shared across voxels:

* `"global"` and `"run"` estimate one model (per run) from all voxels. All
  three residual models are available.
* `"voxel"` (weight-matrix methods `r_optimized`, `cpp_optimized`, `cpp`)
  groups voxels into `voxel_bins` bins of similar residual autocorrelation,
  refits one AR model per bin, and fits each bin with its own filtered design.
  `"parcel"` does the same for user-supplied parcels. These use `"aggregate"`
  or `"full"` residuals; fmriAR's correction is not available for them.

Every prewhitened result carries the fitted plan:

```{r inspect-plan}
run <- simulate_run(c(1.5, 4), 0.5)
fit <- lss(run$Y, run$X, prewhiten = prewhiten_options(method = "ar", p = 1))
plan <- attr(fit, "whiten_plan")
plan$phi
```

```{r inspect-plan-check, include = FALSE}
stopifnot(inherits(plan, "fmriAR_plan"), abs(plan$phi[[1]] - 0.4) < 0.15)
```

## Where to go next

- `?prewhiten_options` and `?lss` list every prewhitening field.
- `vignette("oasis_method")` covers prewhitening with OASIS and multi-basis
  designs.
- `vignette("lss_with_fmridesign")` covers multi-run designs and run-aware
  pooling.
