---
title: 'Rank-1 GLM: Learning One HRF per Voxel While Estimating Trials'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Rank-1 GLM: Learning One HRF per Voxel While Estimating Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
library(fmrilss)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

The haemodynamic response differs between voxels, and a canonical HRF that is
wrong for a voxel biases every trial estimate in it. Estimating a free HRF per
trial is hopelessly noisy. The *rank-1 GLM* of Pedregosa et al. (2015) sits
in between: within a voxel, every trial shares one HRF, expressed in a small
basis, and each trial has its own amplitude. `lss_rank1()` implements that
model, including its least-squares-separate variant, with a solver designed
for whole-brain data.

This article explains the model, how fmrilss fits it and why that solver was
chosen over the quasi-Newton method of the original paper, and which variant
to use for which design. Read `vignette("voxel-wise-hrf")` first for HRF
bases and the `VoxelHRF` object.

## The model

For trial \(i\), let \(X_i\) (\(n \times K\)) be the trial's onsets convolved
with each of \(K\) basis functions, \(h \in \mathbb{R}^K\) the voxel's HRF
coefficients and \(T_X = \sum_j X_j\). `lss_rank1()` fits, per voxel, either

* `model = "joint"` (R1-GLM): one least-squares-all model,
  \[ \min_{h, \beta} \; \Big\| y - \sum_i \beta_i X_i h \Big\|^2, \]
* `model = "separate"` (R1-GLMS, the default): one least-squares-separate
  model per trial, all sharing \(h\),
  \[ \min_{h, \beta, r} \; \sum_i \big\| y - \beta_i X_i h - r_i (T_X - X_i) h \big\|^2 . \]

With `trial_groups`, the single "other trials" term of the separate model
becomes one term per group, \(\sum_g r_{ig} (A_g - [g = g_i] X_i) h\) with
\(A_g\) the summed blocks of group \(g\): the LSS-N design of `lss()`.

Confounds are projected out first, which gives each trial-wise model its own
confound coefficients as in `lss()`. The fitted HRF is oriented to correlate
positively with a reference (`ref_hrf`, the canonical HRF by default) and
scaled to a unit positive peak; the amplitudes absorb that scale.

## How it is fitted, and why not L-BFGS

The objectives are bilinear: linear in \(h\) for fixed amplitudes, and linear
in the amplitudes for fixed \(h\). Pedregosa et al. minimize them jointly with
L-BFGS-B, using a box constraint and a penalty to control the scale
ambiguity \(h \to c\,h,\ \beta \to \beta / c\). `lss_rank1()` instead uses
**exact alternating least squares (ALS)**:

1. *Amplitude step.* For fixed \(h\), the trial regressors \(X_i h\) are fixed,
   so the amplitudes are ordinary LSS (or LSS-N, or LSA) estimates: one
   \(2 \times 2\) (or \((G+1) \times (G+1)\)) solve per trial.
2. *HRF step.* For fixed amplitudes the objective is a least-squares problem in
   \(K\) unknowns: one \(K \times K\) solve.
3. Rescale \(h\) to unit norm (the objective is unchanged) and repeat until the
   relative change in the objective is below `tol`.

Each step minimizes the objective exactly over its block, so the objective
never increases, there is no step size or line search, and the scale
ambiguity is removed by step 3 instead of by constraints.

Both steps need only small Gram blocks. Writing \(G_{ii} = X_i^\top X_i\),
\(S_i = X_i^\top T_X\), \(G_{TT} = T_X^\top T_X\) and \(U_i = X_i^\top y\), the
amplitude step for trial \(i\) uses
\(h^\top G_{ii} h,\ h^\top S_i h,\ h^\top G_{TT} h,\ h^\top U_i\), and the HRF
step solves
\[
\Big[\sum_i a_i^2 G_{ii} + a_i r_i (S_i + S_i^\top) + r_i^2 G_{TT}\Big] h =
\sum_i a_i U_i + r_i\, T_X^\top y,
\qquad a_i = \beta_i - r_i .
\]
So after one matrix product \(X^\top Y\) for all voxels, an iteration of the
separate model costs \(O(T K^2)\) per voxel, independent of the number of
scans, and voxels are processed in parallel with OpenMP. The joint model's
amplitude step is a \(T \times T\) solve, \(O(T^2 K^2 + T^3)\) per voxel, so it
suits regions of interest.

### The comparison

To decide between the two solvers on evidence, fmrilss also ships the paper's
approach as `solver = "lbfgs"`: L-BFGS-B over \((h, \beta, r)\) with the
analytic gradient of the *same* objective, computed from the same Gram
blocks, and started from the same point. `bench/rank1/run_rank1_bench.R` in
the package sources compares them single-threaded on 1,000 voxels whose HRFs
peak between 4 and 7.5 s, for a moderate design (inter-trial interval 3--8 s)
and a rapid, noisy one (1.5--4 s), two bases (SPMG3 and a 12-bin FIR), and
both initializations:

```{r solver-table, echo = FALSE}
solver_summary <- data.frame(
  Basis = c("SPMG3", "SPMG3", "FIR (12)", "FIR (12)"),
  Tolerance = c("default", "tight", "default", "tight"),
  `ALS seconds` = c("0.12-0.15", "0.12-0.17", "0.18-0.30", "0.20-0.34"),
  `L-BFGS-B seconds` = c("0.31-8.5", "8.2-36", "1.3-20", "15-62"),
  `ALS iterations` = c("2-4", "3-6", "3-4", "4-6"),
  `L-BFGS-B evaluations` = c("9-168", "277-746", "9-171", "264-541"),
  `Worst objective gap, ALS` = c("1e-8", "1e-11", "8e-9", "1e-11"),
  `Worst objective gap, L-BFGS-B` = c("5e-5", "3e-8", "6e-3", "2e-11"),
  check.names = FALSE
)
knitr::kable(
  solver_summary,
  caption = paste(
    "Ranges over the two designs and two initializations (1,000 voxels,",
    "one thread). The objective gap is relative to the best objective any",
    "solver reached for that voxel."
  )
)
```

Run to tight tolerances, both solvers reach the same solution in every voxel.
ALS gets there in a handful of iterations: 2.5--7 times faster than
L-BFGS-B at its default tolerance when started from the data-driven
initialization, and roughly 30--200 times faster when started from the
reference HRF or run to a tight tolerance. At the default tolerance, ALS is also closer to the optimum. Two
properties explain this. The amplitude step solves for all \(2T\) amplitudes
exactly, which removes most of the coupling a gradient method must discover
iteration by iteration. And the scale ambiguity leaves the joint objective
with a flat, badly conditioned direction that slows quasi-Newton methods but
does not affect ALS, which renormalizes \(h\). ALS also parallelizes over voxels,
whereas R's L-BFGS-B is not thread-safe. `solver = "lbfgs"` remains
available, single-threaded, for comparison.

The small live check below confirms that the two solvers agree:

```{r simulate}
simulate_voxels <- function(n_voxels, iti = c(3, 8), amplitude = c("positive", "opposite"),
                            noise_sd = 0.8, n = 300, seed = 1) {
  amplitude <- match.arg(amplitude)
  set.seed(seed)
  sframe <- fmrihrf::sampling_frame(blocklens = n, TR = 1)
  onsets <- cumsum(c(12, runif(400, iti[1], iti[2])))
  onsets <- onsets[onsets < n - 30]
  events <- data.frame(
    onset = onsets, duration = 0,
    condition = rep(c("A", "B"), length.out = length(onsets))
  )
  # each voxel gets a double-gamma HRF peaking between 4 and 7.5 s
  peak <- runif(n_voxels, 4, 7.5)
  grid <- seq(0, 30, by = 0.1)
  hrf_of <- function(p) {
    a <- p / 0.9 + 1
    h <- dgamma(grid, a, scale = 0.9) - dgamma(grid, a + 10, scale = 0.9) / 6
    h / max(h)
  }
  stick <- function(o) {
    u <- numeric(n * 10)
    u[round(o * 10) + 1] <- 1
    u
  }
  sticks <- vapply(onsets, stick, numeric(n * 10))
  amp_mean <- if (amplitude == "positive") 1 else ifelse(events$condition == "A", 1, -1)
  betas <- matrix(rnorm(length(onsets) * n_voxels, amp_mean, 0.5),
                  length(onsets), n_voxels)
  Y <- vapply(seq_len(n_voxels), function(v) {
    drive <- drop(sticks %*% betas[, v])
    bold <- stats::filter(c(numeric(length(grid) - 1), drive), hrf_of(peak[v]),
                          sides = 1)[-seq_len(length(grid) - 1)]
    bold[seq(1, n * 10, by = 10)]
  }, numeric(n))
  noise <- apply(matrix(rnorm(n * n_voxels), n, n_voxels), 2, function(e) {
    as.numeric(stats::filter(e, 0.3, "recursive"))
  })
  list(Y = Y + noise_sd * noise, events = events, sframe = sframe,
       betas = betas, peak = peak)
}

sim <- simulate_voxels(60)
basis <- fmrihrf::HRF_SPMG3
time_als <- system.time(
  als <- lss_rank1(sim$Y, sim$events, basis, sim$sframe, tol = 1e-10)
)[["elapsed"]]
time_lbfgs <- system.time(
  lbfgs <- lss_rank1(sim$Y, sim$events, basis, sim$sframe, solver = "lbfgs",
                     tol = 10 * .Machine$double.eps)
)[["elapsed"]]
c(
  max_relative_objective_difference =
    max(abs(als$objective - lbfgs$objective) / als$objective),
  max_beta_difference = max(abs(als$beta - lbfgs$beta)),
  als_seconds = time_als,
  lbfgs_seconds = time_lbfgs
)
```

```{r simulate-check, include = FALSE}
stopifnot(
  max(abs(als$objective - lbfgs$objective) / als$objective) < 1e-8,
  all(als$converged)
)
```

## Which variant to use

The same benchmark compares the estimators by the mean per-voxel correlation
between estimated and true trial amplitudes, across four amplitude regimes.
The oracle rows know each voxel's true HRF.

```{r estimator-table, echo = FALSE}
estimators <- data.frame(
  Estimator = c(
    "Oracle LSS (true HRF)", "Oracle LSS-N (true HRF)",
    "LSS, canonical HRF", "LSS-N, canonical HRF",
    "estimate_voxel_hrf() + lss_with_hrf()",
    "lss_rank1(), separate", "lss_rank1(), separate + trial_groups",
    "lss_rank1(), joint"
  ),
  `Positive amplitudes` = c(0.633, 0.632, 0.559, 0.557, 0.620, 0.620, 0.616, 0.571),
  `Positive, rapid and noisy` = c(0.390, 0.375, 0.341, 0.329, 0.368, 0.369, 0.339, 0.118),
  `Zero-mean amplitudes` = c(0.807, 0.807, 0.729, 0.728, 0.453, 0.600, 0.592, 0.766),
  `Opposite-signed conditions` = c(0.649, 0.880, 0.505, 0.691, 0.147, 0.272, 0.876, 0.640),
  check.names = FALSE
)
knitr::kable(
  estimators, digits = 3,
  caption = "Trial-amplitude recovery (2,000 voxels, SPMG3 basis)."
)
```

* **Trials that share a response sign** (the common activation setting):
  the separate model recovers most of the gap between the canonical HRF and
  the oracle, in milliseconds per thousand voxels.
* **Conditions with different mean responses, especially opposite signs:**
  supply `trial_groups`. In the separate model the HRF is learned mainly
  through the "other trials" regressors. With one pooled regressor, opposite
  responses cancel and the HRF is learned poorly (0.27). With one regressor per
  condition, each condition informs the shared HRF with its own amplitude
  (0.88, matching the LSS-N oracle).
* **Responses that vary around zero within a condition:** the pooled
  regressors carry little information about the HRF, and the joint model,
  which learns from every trial's amplitude, is better (0.77). Prefer it when
  the design is not rapid and the region is small enough for its cost.
* **Rapid designs:** the joint model inherits LSA's variance and degrades
  sharply (0.12); use the separate model.
* **Basis size:** in these simulations the 3-function SPMG3 basis
  outperformed a 12-bin FIR basis for amplitude recovery. More flexible bases
  pay in variance, as Pedregosa et al. also found for decoding.

Two further cautions. The separate model inherits LSS's approximation: with
overlapping trials its pooled "other trials" regressor cannot represent
neighbouring trials' different amplitudes, so it is biased even without
noise. Second, the objectives are not convex. `init = "aggregate"` (the pooled
shape of `estimate_voxel_hrf()`) and `init = "reference"` usually reach the
same solution, but voxels with little signal can differ.

## Learning HRFs on one run and applying them to another

The `hrf` element is a `VoxelHRF` object, so the learned HRFs can estimate
trials in new data with `lss_with_hrf()`. Fitting the HRF on training data
and applying it to held-out data avoids circularity in encoding and decoding
analyses.

```{r held-out}
train <- simulate_voxels(40, seed = 2)
fit <- lss_rank1(train$Y, train$events, basis, train$sframe)
str(fit$hrf$coefficients)

# Re-estimating trials in the training data with the learned HRFs reproduces
# the separate model's amplitudes exactly.
refit <- lss_with_hrf(train$Y, train$events, fit$hrf, verbose = FALSE)
max(abs(unclass(refit)[seq_len(nrow(train$events)), ] - fit$beta))
```

```{r held-out-check, include = FALSE}
stopifnot(max(abs(unclass(refit)[seq_len(nrow(train$events)), ] - fit$beta)) < 1e-8)
```

For held-out data, call `lss_with_hrf(Y_test, events_test, fit$hrf)` with the
test run's data, events and sampling frame stored in the `VoxelHRF` object (or
supplied through `sframe`).

## Reading the result

```{r result-fields}
names(fit)
summary(fit$iterations)
all(fit$converged)
```

* `beta`: trial amplitudes in peak-response units (trials x voxels).
* `other`: the "other trials" amplitudes of each trial's model (trials x
  voxels, or trials x groups x voxels with `trial_groups`).
* `hrf`: the unit-peak HRF coefficients as a `VoxelHRF` object.
* `objective`, `iterations`, `converged`: per-voxel fit diagnostics.
* `degenerate`: voxels whose estimated HRF has no positive peak of at least
  5% of its largest absolute deflection after orientation, typically voxels
  without signal. They are scaled by their
  largest absolute value instead, so their amplitudes are not in
  peak-response units, and a warning reports how many there are. This is the
  same policy and normalization code as `estimate_voxel_hrf()`; the flags are
  also stored in `fit$hrf$degenerate` and passed on by `lss_with_hrf()`.

`prewhiten` accepts global or run pooling; the noise model is estimated from
one aggregate regressor per basis function (see `vignette("prewhitening")`).

## Reference

Pedregosa, F., Eickenberg, M., Ciuciu, P., Thirion, B., & Gramfort, A.
(2015). Data-driven HRF estimation for encoding and decoding models.
*NeuroImage*, 104, 209--220.
