---
title: 'Shared-basis HRF matching: shapes, score margins, and trial coefficients'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Shared-basis HRF matching: shapes, score margins, and trial coefficients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(family = params$family, preset = params$preset)
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)
set.seed(811)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## What this workflow estimates

This is the final specialist article in the learning path. Read
`vignette("fmrilss")`, `vignette("oasis_method")`, and
`vignette("voxel-wise-hrf")` first. The voxel-wise HRF article estimates a
continuous shape in a small basis. This article instead restricts each voxel to
a finite library, or to a declared blend of nearby library members.

SBHM separates two quantities:

1. a voxel-specific HRF **shape coordinate** in a shared basis; and
2. a trial-wise **coefficient on the event design built from that shape**.

The coefficient is not automatically a peak BOLD response. With
`normalize = TRUE`, `sbhm_build()` normalizes each candidate to discrete unit
\(L_2\) energy before rank truncation. The returned coefficient is therefore on
the scale of the selected rank-truncated waveform and the supplied event
duration and amplitude. Changing any of those changes the coefficient scale.

SBHM produces adaptive point estimates. Shape matching, optional blending,
ridge penalties, and reuse of the same data for shape selection and coefficient
estimation are not covered by a calibrated standard-error or test-statistic
contract. Every SBHM amplitude route rejects `return_se = TRUE`.

```{r imports}
suppressPackageStartupMessages({
  library(fmrihrf)
  library(fmrilss)
})
```

```{r contract-table}
contract <- data.frame(
  Object = c("Library", "Shared basis", "Shape match", "Trial output", "Uncertainty"),
  Contract = c(
    "Finite, uniquely named candidate waveforms; optional discrete-L2 normalization",
    "Rank-r SVD representation with total-energy and numerical-rank metadata",
    "Hard candidate or explicit soft blend; margin is an uncalibrated score difference",
    "Coefficient on the supplied event design using the selected rank-truncated shape",
    "Point estimates only; SBHM standard errors and tests are not calibrated"
  )
)
knitr::kable(contract, caption = "The estimands and boundaries used throughout this article.")
```

## Build one transparent library

For a compact computational example, we use a grid of positive gamma curves.
It is a demonstration library, not a package recommendation and not a claim
that the grid spans physiological HRFs. A scientific analysis must justify its
own candidates, preprocessing, and coverage.

```{r build-library}
TR <- 1
n_time <- 180L
sframe <- sampling_frame(blocklens = n_time, TR = TR)
time <- samples(sframe, global = TRUE)
span <- 32

library_grid <- expand.grid(
  shape = c(5, 6.5, 8, 9.5),
  rate = c(0.85, 1.00, 1.15)
)
library_H <- vapply(seq_len(nrow(library_grid)), function(i) {
  h <- dgamma(time, shape = library_grid$shape[i], rate = library_grid$rate[i])
  h[time > span] <- 0
  h
}, numeric(length(time)))
colnames(library_H) <- sprintf(
  "gamma_s%02d_r%03d",
  round(10 * library_grid$shape), round(100 * library_grid$rate)
)

rank_used <- 4L
sbhm <- sbhm_build(
  library_H = library_H,
  r = rank_used,
  tgrid = time,
  span = span,
  normalize = TRUE,
  baseline = NULL,
  ref = "mean"
)
```

### Rank is a declared approximation

`sbhm_build()` performs one decomposition and retains the full singular-value
spectrum as metadata. The cumulative fraction below uses total squared
singular-value energy—not the energy of the largest rank we happened to plot.
Rank four is a modeling choice for this fixture; the table makes its
approximation error inspectable rather than turning it into a generic rule.

```{r rank-court}
singular_values <- sbhm$meta$singular_values_full
rank_table <- data.frame(
  Rank = seq_along(singular_values),
  CumulativeEnergy = cumsum(singular_values^2) / sum(singular_values^2),
  RelativeResidual = sqrt(
    pmax(0, 1 - cumsum(singular_values^2) / sum(singular_values^2))
  )
)
rank_table <- head(rank_table, 6)
rank_table$CumulativeEnergy <- round(rank_table$CumulativeEnergy, 6)
rank_table$RelativeResidual <- round(rank_table$RelativeResidual, 6)
knitr::kable(
  rank_table,
  caption = "Total library energy retained and relative Frobenius residual by rank."
)
```

```{r rank-court-check, include=FALSE}
stopifnot(
  sbhm$meta$numerical_rank >= rank_used,
  abs(sbhm$meta$retained_energy_fraction -
        rank_table$CumulativeEnergy[rank_used]) < 1e-6,
  all(diff(rank_table$CumulativeEnergy) >= 0),
  all(diff(rank_table$RelativeResidual) <= 0)
)
```

The shared basis `B` is orthonormal on this sampled grid. The coordinates `A`
reconstruct the rank-four candidate waveforms as `B %*% A`; basis vectors are
algebraic modes, not direct estimates of latency, width, or physiology.

## A truth-referenced fixture

The fixture contains twelve voxels. Six use an exact library member and six use
an off-library convex mixture. This tests both interpolation and the limitation
of hard matching. Trial coefficients vary across trials and voxels. Motion-like
nuisance signals and independent Gaussian noise are added explicitly.

```{r simulate-fixture}
n_voxels <- 12L
voxel_names <- paste0("voxel_", sprintf("%02d", seq_len(n_voxels)))
onsets <- seq(12, 138, length.out = 8)
n_trials <- length(onsets)
trial_names <- paste0("event_", sprintf("%02d", seq_len(n_trials)))
design_spec <- list(
  sframe = sframe,
  cond = list(
    onsets = onsets,
    duration = 0,
    amplitude = 1,
    span = span,
    trial_names = trial_names
  ),
  precision = 0.1,
  method = "conv"
)

on_library_idx <- c(1L, 3L, 5L, 7L, 9L, 11L)
alpha_true <- matrix(NA_real_, rank_used, n_voxels)
alpha_true[, 1:6] <- sbhm$A[, on_library_idx, drop = FALSE]
mix_pairs <- rbind(c(1, 2), c(3, 4), c(5, 6), c(7, 8), c(9, 10), c(11, 12))
mix_weight <- c(0.25, 0.40, 0.55, 0.70, 0.35, 0.60)
for (v in 1:6) {
  alpha_true[, 6 + v] <-
    mix_weight[v] * sbhm$A[, mix_pairs[v, 1]] +
    (1 - mix_weight[v]) * sbhm$A[, mix_pairs[v, 2]]
}
colnames(alpha_true) <- voxel_names

hrf_B <- sbhm_hrf(sbhm$B, sbhm$tgrid, sbhm$span)
trial_regs <- lapply(seq_along(onsets), function(j) {
  rr <- regressor(
    onsets = onsets[j], hrf = hrf_B,
    duration = 0, amplitude = 1, span = span, summate = FALSE
  )
  as.matrix(evaluate(rr, grid = time, precision = 0.1, method = "conv"))
})

trial_index <- seq_len(n_trials)
voxel_index <- seq_len(n_voxels)
beta_true <- outer(trial_index, voxel_index, function(i, v) {
  1.6 + 0.35 * sin(i / 1.7 + v / 4) + 0.12 * cos(i * v / 5)
})
dimnames(beta_true) <- list(trial_names, voxel_names)

Nuisance <- cbind(
  linear = seq(-1, 1, length.out = n_time),
  slow_sine = sin(2 * pi * time / max(time))
)
nuisance_coef <- rbind(
  seq(-0.25, 0.25, length.out = n_voxels),
  seq(0.20, -0.20, length.out = n_voxels)
)
Y_mean <- matrix(2, n_time, n_voxels) + Nuisance %*% nuisance_coef
for (v in seq_len(n_voxels)) {
  for (j in seq_len(n_trials)) {
    Y_mean[, v] <- Y_mean[, v] +
      beta_true[j, v] * as.numeric(trial_regs[[j]] %*% alpha_true[, v])
  }
}
set.seed(812)
Y <- Y_mean + matrix(rnorm(n_time * n_voxels, sd = 0.04), n_time, n_voxels)
colnames(Y) <- voxel_names
```

The main call makes every consequential choice explicit. Zero penalties are
used so the later direct-GLM oracle has the same estimand. Hard matching is used
first; soft matching is evaluated separately rather than being hidden in a
default.

```{r fit-hard}
fit_hard <- lss_sbhm(
  Y = Y,
  sbhm = sbhm,
  design_spec = design_spec,
  Nuisance = Nuisance,
  prepass = list(ridge = list(mode = "absolute", lambda = 0)),
  match = list(topK = 1, soft_blend = FALSE, whiten = FALSE),
  oasis = list(ridge_mode = "absolute", ridge_x = 0, ridge_b = 0),
  amplitude = list(method = "global_ls", ridge = 0),
  return = "both"
)
```

```{r output-contract}
output_contract <- data.frame(
  Component = c(
    "alpha_coords", "amplitude", "coeffs_r", "matched_name", "shape_mode",
    "event_amplitude / event_duration"
  ),
  Shape = c(
    paste(dim(fit_hard$alpha_coords), collapse = " x "),
    paste(dim(fit_hard$amplitude), collapse = " x "),
    paste(dim(fit_hard$coeffs_r), collapse = " x "),
    as.character(length(fit_hard$matched_name)),
    as.character(length(fit_hard$shape_mode)),
    paste(length(fit_hard$event_amplitude), "/", length(fit_hard$event_duration))
  ),
  Meaning = c(
    "rank x voxel coordinates actually used by the amplitude refit",
    "trial x voxel event-design coefficients",
    "rank x trial x voxel OASIS coefficients",
    "top-scoring library candidate for each voxel",
    "hard, soft, or fallback shape policy actually used",
    "named event-design scale metadata aligned with amplitude rows"
  )
)
knitr::kable(output_contract, caption = "Reader-facing SBHM output contract for the fitted fixture.")
```

```{r output-contract-check, include=FALSE}
stopifnot(
  identical(dim(fit_hard$alpha_coords), c(rank_used, n_voxels)),
  identical(dim(fit_hard$amplitude), c(n_trials, n_voxels)),
  identical(dim(fit_hard$coeffs_r), c(rank_used, n_trials, n_voxels)),
  identical(colnames(fit_hard$alpha_coords), voxel_names),
  identical(dimnames(fit_hard$amplitude), list(trial_names, voxel_names)),
  identical(dimnames(fit_hard$coeffs_r)[2:3], list(trial_names, voxel_names)),
  identical(names(fit_hard$matched_name), voxel_names),
  identical(names(fit_hard$event_amplitude), trial_names),
  identical(names(fit_hard$event_duration), trial_names),
  identical(fit_hard$trial_basis_map$trial_name,
            rep(trial_names, each = rank_used))
)
```

## Evaluate the shape actually used

Exact library index is secondary: nearby candidates can have almost identical
waveforms. We therefore evaluate `B %*% alpha_coords` for every voxel. Angular
error is zero for identical waveform direction and ignores arbitrary positive
scale. Unit-norm waveform RMSE adds a time-domain discrepancy.

```{r shape-metrics, include=FALSE}
waveform_metrics <- function(alpha_est, alpha_truth, B) {
  estimated <- B %*% alpha_est
  truth <- B %*% alpha_truth
  est_norm <- sweep(estimated, 2, sqrt(colSums(estimated^2)), "/")
  truth_norm <- sweep(truth, 2, sqrt(colSums(truth^2)), "/")
  cosine <- pmin(1, pmax(-1, colSums(est_norm * truth_norm)))
  data.frame(
    AngularErrorRadians = acos(cosine),
    UnitWaveformRMSE = sqrt(colMeans((est_norm - truth_norm)^2))
  )
}

hard_shape_by_voxel <- waveform_metrics(fit_hard$alpha_coords, alpha_true, sbhm$B)
shape_group <- rep(c("On-library", "Off-library"), each = 6)
shape_summary <- do.call(rbind, lapply(split(hard_shape_by_voxel, shape_group), function(x) {
  data.frame(
    MeanAngularErrorRadians = mean(x$AngularErrorRadians),
    MaxAngularErrorRadians = max(x$AngularErrorRadians),
    MeanUnitWaveformRMSE = mean(x$UnitWaveformRMSE)
  )
}))
shape_summary$Truth <- rownames(shape_summary)
rownames(shape_summary) <- NULL
shape_summary <- shape_summary[, c("Truth", setdiff(names(shape_summary), "Truth"))]
```

```{r shape-summary}
shape_summary_display <- shape_summary
shape_summary_display[-1] <- lapply(shape_summary_display[-1], round, 5)
knitr::kable(
  shape_summary_display,
  caption = "All-voxel hard-match waveform error, separated by library coverage."
)
```

```{r shape-plot-prep, include=FALSE}
true_waveforms <- sbhm$B %*% alpha_true
hard_waveforms <- sbhm$B %*% fit_hard$alpha_coords
shape_alt <- sprintf(
  paste0(
    "Twelve panels compare true solid and hard-matched dashed rank-four waveforms. ",
    "Mean angular error is %.3f radians on-library and %.3f radians off-library."
  ),
  shape_summary$MeanAngularErrorRadians[shape_summary$Truth == "On-library"],
  shape_summary$MeanAngularErrorRadians[shape_summary$Truth == "Off-library"]
)
```

```{r shape-plot, fig.width=10, fig.height=7, fig.alt=shape_alt, fig.cap="True and hard-matched rank-four waveforms for every voxel. Solid blue is truth; dashed orange is the coordinate actually used by the amplitude refit. Panels marked '(off)' contain off-library mixtures and are intentionally harder."}
local({
  oldpar <- par(mfrow = c(3, 4), mar = c(2.7, 2.8, 2.1, 0.6), bg = "white")
  on.exit(par(oldpar), add = TRUE)
  keep <- time <= span
  for (v in seq_len(n_voxels)) {
    ylim <- range(true_waveforms[keep, v], hard_waveforms[keep, v])
    plot(time[keep], true_waveforms[keep, v], type = "l", lwd = 2,
         col = "#2c7fb8", ylim = ylim, xlab = "Seconds", ylab = "Shape",
         main = paste0(voxel_names[v], if (v <= 6) " (on)" else " (off)"))
    lines(time[keep], hard_waveforms[keep, v], lwd = 2, lty = 2, col = "#d95f02")
    abline(h = 0, col = "gray85")
  }
})
```

For the six on-library voxels, exact candidate identity is also judgeable:

```{r index-secondary}
exact_index_accuracy <- mean(
  fit_hard$matched_name[1:6] == colnames(sbhm$A)[on_library_idx]
)
data.frame(OnLibraryExactCandidateAccuracy = exact_index_accuracy)
```

## Hard and soft matching answer different questions

Soft matching uses cosine-score softmax weights; those weights are not mixture
fractions and the score margin is not calibrated confidence. We compare an
explicit soft policy with the explicit hard fit against the same truth. The
result is descriptive for this fixture, not a rule that soft matching reduces
variance or improves interpolation.

```{r fit-soft}
fit_soft <- lss_sbhm(
  Y = Y,
  sbhm = sbhm,
  design_spec = design_spec,
  Nuisance = Nuisance,
  prepass = list(ridge = list(mode = "absolute", lambda = 0)),
  match = list(
    topK = 3, soft_blend = TRUE, blend_margin = 2,
    whiten = FALSE
  ),
  oasis = list(ridge_mode = "absolute", ridge_x = 0, ridge_b = 0),
  amplitude = list(method = "global_ls", ridge = 0),
  return = "both"
)
```

```{r compare-policies}
soft_shape_by_voxel <- waveform_metrics(fit_soft$alpha_coords, alpha_true, sbhm$B)
policy_table <- rbind(
  data.frame(
    Policy = "Hard top-1",
    VoxelsUsingPolicy = sum(fit_hard$shape_mode == "hard"),
    MeanAngularError = mean(hard_shape_by_voxel$AngularErrorRadians),
    AmplitudeBias = mean(fit_hard$amplitude - beta_true),
    AmplitudeRMSE = sqrt(mean((fit_hard$amplitude - beta_true)^2))
  ),
  data.frame(
    Policy = "Soft top-3",
    VoxelsUsingPolicy = sum(fit_soft$shape_mode == "soft"),
    MeanAngularError = mean(soft_shape_by_voxel$AngularErrorRadians),
    AmplitudeBias = mean(fit_soft$amplitude - beta_true),
    AmplitudeRMSE = sqrt(mean((fit_soft$amplitude - beta_true)^2))
  )
)
policy_table[-1] <- lapply(policy_table[-1], round, 5)
knitr::kable(
  policy_table,
  caption = "Truth-referenced hard and soft results for this fixed fixture."
)
```

```{r policy-check, include=FALSE}
stopifnot(
  all(fit_hard$shape_mode == "hard"),
  all(fit_soft$shape_mode == "soft")
)
```

The returned `margin` is simply the top-one minus top-two cosine score. Low
margin can reveal near-ties, but no universal `min_margin` or `blend_margin`
follows from it. If gating is used, inspect `shape_mode` and
`fallback_low_conf`; `matched_name` remains the top-scoring candidate, whereas
`alpha_coords` records the shape actually used.

## Bind the coefficient estimator to an independent GLM

With `amplitude$method = "global_ls"` and zero ridge, the final stage is an
ordinary trial-wise GLM conditional on the selected shape. The following oracle
assembles every voxel model independently from the public `fmrihrf` regressors.
It proves implementation identity; it does not remove shape-selection bias.

```{r direct-oracle, include=FALSE}
direct_amplitude <- function(alpha) {
  out <- matrix(NA_real_, n_trials, n_voxels,
                dimnames = list(trial_names, voxel_names))
  common <- cbind(intercept = 1, Nuisance)
  for (v in seq_len(n_voxels)) {
    Xv <- do.call(cbind, lapply(
      trial_regs, function(Xt) as.numeric(Xt %*% alpha[, v])
    ))
    fit <- lm.fit(cbind(common, Xv), Y[, v])
    out[, v] <- tail(fit$coefficients, n_trials)
  }
  out
}
direct_hard <- direct_amplitude(fit_hard$alpha_coords)
known_shape <- direct_amplitude(alpha_true)
direct_error <- max(abs(direct_hard - fit_hard$amplitude))
amplitude_table <- data.frame(
  Estimator = c("Known truth shape", "SBHM hard-selected shape", "SBHM soft-selected shape"),
  Bias = c(
    mean(known_shape - beta_true),
    mean(fit_hard$amplitude - beta_true),
    mean(fit_soft$amplitude - beta_true)
  ),
  RMSE = c(
    sqrt(mean((known_shape - beta_true)^2)),
    sqrt(mean((fit_hard$amplitude - beta_true)^2)),
    sqrt(mean((fit_soft$amplitude - beta_true)^2))
  )
)
```

```{r direct-oracle-table}
amplitude_display <- amplitude_table
amplitude_display[-1] <- lapply(amplitude_display[-1], round, 5)
amplitude_display$DirectGLMMaxError <- c(NA, format(direct_error, scientific = TRUE), NA)
knitr::kable(
  amplitude_display,
  caption = "Trial-coefficient recovery and the independent all-cell GLM identity check."
)
```

```{r direct-oracle-check, include=FALSE}
stopifnot(
  direct_error < 1e-10,
  all(is.finite(as.matrix(amplitude_table[-1]))),
  identical(dimnames(fit_hard$amplitude), dimnames(beta_true))
)
```

```{r amplitude-plot-prep, include=FALSE}
amp_bias <- mean(fit_hard$amplitude - beta_true)
amp_rmse <- sqrt(mean((fit_hard$amplitude - beta_true)^2))
amp_alt <- sprintf(
  paste0(
    "Scatter of 96 true and hard-match event-design coefficients. ",
    "Bias is %.3f and RMSE is %.3f; a diagonal reference line is shown."
  ),
  amp_bias, amp_rmse
)
amp_range <- range(beta_true, fit_hard$amplitude)
```

```{r amplitude-plot, fig.alt=amp_alt, fig.cap="Hard-match event-design coefficients against truth for all trials and voxels. The red dashed line is equality; the displayed bias and RMSE are conditional on this fixed simulation."}
plot(
  as.vector(beta_true), as.vector(fit_hard$amplitude),
  pch = 19, col = grDevices::adjustcolor("#2c7fb8", alpha.f = 0.55),
  xlim = amp_range, ylim = amp_range,
  xlab = "True event-design coefficient",
  ylab = "Estimated event-design coefficient",
  main = sprintf("Hard matching: bias %.3f, RMSE %.3f", amp_bias, amp_rmse)
)
abline(0, 1, col = "#d7301f", lwd = 2, lty = 2)
grid()
```

## Factorization accelerates only the prepass

`data_fac` represents `Y` as `scores %*% loadings`, where scores are
\(T \times q\) and loadings are \(q \times V\). The full `Y` is still required by
`lss_sbhm()` for OASIS and the amplitude stage. The shortcut therefore reduces
only prepass multiplication; it does not make the full pipeline independent of
\(V\). Named factor axes must agree between scores and loadings; when `Y` has
voxel names, loadings must carry the same names and are aligned before use.
Active prewhitening with `data_fac` fails explicitly because the supplied
factorization is not a factorization of the estimated whitened data.

```{r factorized-prepass}
set.seed(813)
q <- 4L
scores <- matrix(rnorm(n_time * q), n_time, q)
loadings <- matrix(rnorm(q * n_voxels), q, n_voxels)
colnames(scores) <- rownames(loadings) <- paste0("factor_", seq_len(q))
colnames(loadings) <- voxel_names
Y_factorized <- scores %*% loadings
colnames(Y_factorized) <- voxel_names

pre_dense <- sbhm_prepass(
  Y_factorized, sbhm, design_spec,
  ridge = list(mode = "absolute", lambda = 0)
)
pre_factorized <- sbhm_prepass(
  Y_factorized, sbhm, design_spec,
  ridge = list(mode = "absolute", lambda = 0),
  data_fac = list(scores = scores, loadings = loadings)
)
factorized_error <- max(abs(pre_dense$beta_bar - pre_factorized$beta_bar))
data.frame(
  Scores = paste(dim(scores), collapse = " x "),
  Loadings = paste(dim(loadings), collapse = " x "),
  Voxels = ncol(Y_factorized),
  DenseVsFactorizedMaxError = factorized_error
)
```

```{r factorized-check, include=FALSE}
stopifnot(q < n_voxels, factorized_error < 1e-10)
factorized_whitening_error <- try(
  sbhm_prepass(
    Y_factorized, sbhm, design_spec,
    prewhiten = list(method = "ar", p = 1L),
    data_fac = list(scores = scores, loadings = loadings)
  ),
  silent = TRUE
)
stopifnot(
  inherits(factorized_whitening_error, "try-error"),
  grepl("does not support active prewhitening", as.character(factorized_whitening_error), fixed = TRUE)
)
```

For a PCA factorization, `prcomp(Y)$x` supplies scores and
`t(prcomp(Y)$rotation)` supplies the required \(q \times V\) loading
orientation. The evaluated path below keeps `Y` as the response for OASIS and
the amplitude stage while using a rank-four uncentered PCA approximation only
for the prepass. This is a computational approximation, not an accuracy claim.

```{r pca-end-to-end}
pca <- prcomp(Y, center = FALSE, scale. = FALSE, rank. = q)
pca_scores <- pca$x[, seq_len(q), drop = FALSE]
pca_loadings <- t(pca$rotation[, seq_len(q), drop = FALSE])

fit_pca_prepass <- lss_sbhm(
  Y = Y,
  sbhm = sbhm,
  design_spec = design_spec,
  Nuisance = Nuisance,
  prepass = list(
    ridge = list(mode = "absolute", lambda = 0),
    data_fac = list(scores = pca_scores, loadings = pca_loadings)
  ),
  match = list(topK = 1, soft_blend = FALSE, whiten = FALSE),
  oasis = list(ridge_mode = "absolute", ridge_x = 0, ridge_b = 0),
  amplitude = list(method = "global_ls", ridge = 0),
  return = "amplitude"
)
data.frame(
  ScoreShape = paste(dim(pca_scores), collapse = " x "),
  LoadingShape = paste(dim(pca_loadings), collapse = " x "),
  FullResponseShape = paste(dim(Y), collapse = " x "),
  OutputShape = paste(dim(fit_pca_prepass$amplitude), collapse = " x ")
)
```

```{r pca-end-to-end-check, include=FALSE}
stopifnot(
  q < n_voxels,
  identical(dim(fit_pca_prepass$amplitude), c(n_trials, n_voxels)),
  all(is.finite(fit_pca_prepass$amplitude))
)
```

## Runs, conditions, and whitening

Raw multi-run `design_spec` inputs use run-relative onsets and an explicit
`cond$run` vector. SBHM constructs every trial inside its own sampling-frame
run and uses run-specific intercepts, so an HRF tail cannot cross a run
boundary. `lss_sbhm_design()` obtains those run identities and heterogeneous
durations from the unique trialwise `fmridesign` term; non-trial event terms and
baseline-model columns enter the fixed nuisance span. See
`vignette("lss_with_fmridesign")` for the event-model identity contract.

If estimated prewhitening is requested, SBHM applies the same transformation to
the response, trial basis, other-condition span, and nuisance design in the
prepass, diagnostics, and final amplitude estimator. The whitening plan is
estimated once from the complete trial and common design, then reused at every
stage. For a multi-run sampling frame, omitted `prewhiten$runs` are filled from
the frame's run boundaries; contradictory labels fail rather than allowing a
temporal filter to bridge runs. This changes the conditional point estimator;
it does not create calibrated post-selection standard errors. Factorized
prepasses and active prewhitening cannot be combined.

## Advanced policies are explicit heuristics

`alpha_source = "trial_projection"` and `"oasis_rank1"`, low-score fallback,
condition gates, ridge penalties, and adaptive ridge controls are available for
specialized studies. They change the estimator. No rank, ridge fraction,
margin threshold, ISI boundary, or TR cutoff is generally recommended here.
Choose and validate such a policy against the design, signal scale, failure
modes, and loss function of the intended analysis.

The three scalar-coefficient engines also answer different model questions:

- `global_ls` fits all trial columns jointly for each selected voxel shape;
- `lss1` fits one target and one aggregate-other column per trial;
- `oasis_voxel` applies the K=1 OASIS operator per voxel.

Exactly coincident or otherwise unidentified trials cannot be recovered by any
of these labels. Ridge may return symmetric finite coefficients, but it does not
restore identification. None of the three engines returns calibrated SBHM
standard errors.

## Takeaways

SBHM is a constrained shape-selection and coefficient-estimation workflow. Its
useful compression comes with visible assumptions: the library defines the
shape space, rank truncation defines the approximation, matching or blending
defines the voxel waveform, and the supplied event design defines coefficient
units. Judge the waveform actually used (`alpha_coords`), retain the identity
maps, and evaluate hard, soft, fallback, ridge, and whitening policies against
truth or external calibration before using them in a scientific conclusion.

This is the end of the vignette learning path. Continue with `?sbhm_build`,
`?sbhm_match`, `?sbhm_prepass`, `?lss_sbhm`, and `?lss_sbhm_design` for the
complete API contracts.
