---
title: 'Shared-basis HRF matching: shapes, score margins, and trial coefficients'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Shared-basis HRF matching: shapes, score margins, and trial coefficients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(family = params$family, preset = params$preset)
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)
set.seed(811)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## What this workflow estimates

Shared-basis HRF matching (SBHM) gives each voxel an HRF shape taken from a
finite library of candidate waveforms, or a declared blend of nearby library
members, and then estimates one coefficient per trial using that shape. The
voxel-wise HRF article instead estimates a continuous shape in a small basis.
This is the final specialist article in the learning path; read
`vignette("fmrilss")`, `vignette("oasis_method")`, and
`vignette("voxel-wise-hrf")` first.

`sbhm_build()` compresses the library into a shared basis, a rank-`r` SVD of the
candidate waveforms, so that each candidate is described by `r` coordinates.
`lss_sbhm()` then works in that basis in four steps:

1. an OASIS fit returns `r` coefficients per trial and voxel (`coeffs_r`);
2. by default, a **prepass** (one aggregate GLM per voxel, with the trials
   summed) summarizes each voxel's response shape;
3. **matching** compares that summary with the library coordinates by cosine
   score and either keeps the top candidate (hard matching) or blends the top
   few (soft matching);
4. an **amplitude refit** estimates one scalar coefficient per trial on an
   event design built from the selected shape.

The workflow separates two quantities:

1. a voxel-specific HRF **shape coordinate** in a shared basis; and
2. a trial-wise **coefficient on the event design built from that shape**.

The coefficient is not automatically a peak BOLD response. With
`normalize = TRUE`, `sbhm_build()` normalizes each candidate to discrete unit
\(L_2\) energy before rank truncation. The returned coefficient is therefore on
the scale of the selected rank-truncated waveform and the supplied event
duration and amplitude. Changing any of those changes the coefficient scale.

SBHM produces adaptive point estimates. The package provides no calibrated
standard errors or test statistics that account for shape matching, optional
blending, ridge penalties, or the reuse of the same data for shape selection and
coefficient estimation. Every SBHM amplitude route rejects `return_se = TRUE`.

```{r imports}
suppressPackageStartupMessages({
  library(fmrihrf)
  library(fmrilss)
})
```

```{r contract-table}
contract <- data.frame(
  Object = c("Library", "Shared basis", "Shape match", "Trial output", "Uncertainty"),
  Contract = c(
    "Finite, uniquely named candidate waveforms; optional discrete-L2 normalization",
    "Rank-r SVD representation with total-energy and numerical-rank metadata",
    "Hard candidate or explicit soft blend; margin is an uncalibrated score difference",
    "Coefficient on the supplied event design using the selected rank-truncated shape",
    "Point estimates only; SBHM standard errors and tests are not calibrated"
  )
)
knitr::kable(contract, caption = "The estimands and boundaries used throughout this article.")
```

## Build a candidate HRF library

For a compact computational example, the library is a grid of twelve positive
gamma curves. It is a demonstration library, not a package recommendation and
not a claim that the grid spans physiological HRFs. A scientific analysis must
justify its own candidates, preprocessing, and coverage.

```{r build-library}
TR <- 1
n_time <- 180L
sframe <- sampling_frame(blocklens = n_time, TR = TR)
time <- samples(sframe, global = TRUE)
span <- 32

library_grid <- expand.grid(
  shape = c(5, 6.5, 8, 9.5),
  rate = c(0.85, 1.00, 1.15)
)
library_H <- vapply(seq_len(nrow(library_grid)), function(i) {
  h <- dgamma(time, shape = library_grid$shape[i], rate = library_grid$rate[i])
  h[time > span] <- 0
  h
}, numeric(length(time)))
colnames(library_H) <- sprintf(
  "gamma_s%02d_r%03d",
  round(10 * library_grid$shape), round(100 * library_grid$rate)
)

rank_used <- 4L
sbhm <- sbhm_build(
  library_H = library_H,
  r = rank_used,
  tgrid = time,
  span = span,
  normalize = TRUE,
  baseline = NULL,
  ref = "mean"
)
```

### Choose and inspect the basis rank

The rank `r` sets how closely the shared basis approximates the library.
`sbhm_build()` performs one decomposition and keeps the full singular-value
spectrum as metadata. The cumulative fraction below is relative to the total
squared singular-value energy of the library, not to the energy of the largest
rank shown in the table. Rank four is a modeling choice for this example, not a
general rule; the table shows how much approximation error it leaves.

```{r rank-court}
singular_values <- sbhm$meta$singular_values_full
rank_table <- data.frame(
  Rank = seq_along(singular_values),
  CumulativeEnergy = cumsum(singular_values^2) / sum(singular_values^2),
  RelativeResidual = sqrt(
    pmax(0, 1 - cumsum(singular_values^2) / sum(singular_values^2))
  )
)
rank_table <- head(rank_table, 6)
rank_table$CumulativeEnergy <- round(rank_table$CumulativeEnergy, 6)
rank_table$RelativeResidual <- round(rank_table$RelativeResidual, 6)
knitr::kable(
  rank_table,
  caption = "Total library energy retained and relative Frobenius residual by rank."
)
```

```{r rank-court-check, include=FALSE}
stopifnot(
  sbhm$meta$numerical_rank >= rank_used,
  abs(sbhm$meta$retained_energy_fraction -
        rank_table$CumulativeEnergy[rank_used]) < 1e-6,
  all(diff(rank_table$CumulativeEnergy) >= 0),
  all(diff(rank_table$RelativeResidual) <= 0)
)
```

The shared basis `B` is orthonormal on this sampled grid. The coordinates `A`
reconstruct the rank-four candidate waveforms as `B %*% A`. The basis vectors
are algebraic modes; they are not direct estimates of latency, width, or
physiology.

## Simulate data with known shapes and coefficients

The rest of the article uses a simulated dataset, the fixture, whose true shapes
and trial coefficients are known. It contains twelve voxels. Six use an exact
library member, and six use an off-library convex mixture of two adjacent
library members. The off-library voxels test interpolation and expose the
limitation of hard matching, which can only return a library member. Trial
coefficients vary across trials and voxels. Drift-like nuisance signals (a linear trend and a slow sine) and
independent Gaussian noise are added explicitly.

```{r simulate-fixture}
n_voxels <- 12L
voxel_names <- paste0("voxel_", sprintf("%02d", seq_len(n_voxels)))
onsets <- seq(12, 138, length.out = 8)
n_trials <- length(onsets)
trial_names <- paste0("event_", sprintf("%02d", seq_len(n_trials)))
design_spec <- list(
  sframe = sframe,
  cond = list(
    onsets = onsets,
    duration = 0,
    amplitude = 1,
    span = span,
    trial_names = trial_names
  ),
  precision = 0.1,
  method = "conv"
)

on_library_idx <- c(1L, 3L, 5L, 7L, 9L, 11L)
alpha_true <- matrix(NA_real_, rank_used, n_voxels)
alpha_true[, 1:6] <- sbhm$A[, on_library_idx, drop = FALSE]
mix_pairs <- rbind(c(1, 2), c(3, 4), c(5, 6), c(7, 8), c(9, 10), c(11, 12))
mix_weight <- c(0.25, 0.40, 0.55, 0.70, 0.35, 0.60)
for (v in 1:6) {
  alpha_true[, 6 + v] <-
    mix_weight[v] * sbhm$A[, mix_pairs[v, 1]] +
    (1 - mix_weight[v]) * sbhm$A[, mix_pairs[v, 2]]
}
colnames(alpha_true) <- voxel_names

hrf_B <- sbhm_hrf(sbhm$B, sbhm$tgrid, sbhm$span)
trial_regs <- lapply(seq_along(onsets), function(j) {
  rr <- regressor(
    onsets = onsets[j], hrf = hrf_B,
    duration = 0, amplitude = 1, span = span, summate = FALSE
  )
  as.matrix(evaluate(rr, grid = time, precision = 0.1, method = "conv"))
})

trial_index <- seq_len(n_trials)
voxel_index <- seq_len(n_voxels)
beta_true <- outer(trial_index, voxel_index, function(i, v) {
  1.6 + 0.35 * sin(i / 1.7 + v / 4) + 0.12 * cos(i * v / 5)
})
dimnames(beta_true) <- list(trial_names, voxel_names)

Nuisance <- cbind(
  linear = seq(-1, 1, length.out = n_time),
  slow_sine = sin(2 * pi * time / max(time))
)
nuisance_coef <- rbind(
  seq(-0.25, 0.25, length.out = n_voxels),
  seq(0.20, -0.20, length.out = n_voxels)
)
Y_mean <- matrix(2, n_time, n_voxels) + Nuisance %*% nuisance_coef
for (v in seq_len(n_voxels)) {
  for (j in seq_len(n_trials)) {
    Y_mean[, v] <- Y_mean[, v] +
      beta_true[j, v] * as.numeric(trial_regs[[j]] %*% alpha_true[, v])
  }
}
set.seed(812)
Y <- Y_mean + matrix(rnorm(n_time * n_voxels, sd = 0.04), n_time, n_voxels)
colnames(Y) <- voxel_names
```

The main call sets every consequential option explicitly. All ridge penalties
are zero, so that the direct-GLM check later in the article estimates the same
quantity. This first fit uses hard matching; soft matching is evaluated
separately rather than left to a default.

```{r fit-hard}
fit_hard <- lss_sbhm(
  Y = Y,
  sbhm = sbhm,
  design_spec = design_spec,
  Nuisance = Nuisance,
  prepass = list(ridge = list(mode = "absolute", lambda = 0)),
  match = list(topK = 1, soft_blend = FALSE, whiten = FALSE),
  oasis = list(ridge_mode = "absolute", ridge_x = 0, ridge_b = 0),
  amplitude = list(method = "global_ls", ridge = 0),
  return = "both"
)
```

```{r output-contract}
output_contract <- data.frame(
  Component = c(
    "alpha_coords", "amplitude", "coeffs_r", "matched_name", "shape_mode",
    "event_amplitude / event_duration"
  ),
  Shape = c(
    paste(dim(fit_hard$alpha_coords), collapse = " x "),
    paste(dim(fit_hard$amplitude), collapse = " x "),
    paste(dim(fit_hard$coeffs_r), collapse = " x "),
    as.character(length(fit_hard$matched_name)),
    as.character(length(fit_hard$shape_mode)),
    paste(length(fit_hard$event_amplitude), "/", length(fit_hard$event_duration))
  ),
  Meaning = c(
    "rank x voxel coordinates actually used by the amplitude refit",
    "trial x voxel event-design coefficients",
    "rank x trial x voxel OASIS coefficients",
    "top-scoring library candidate for each voxel",
    "hard, soft, or fallback shape policy actually used",
    "named event-design scale metadata aligned with amplitude rows"
  )
)
knitr::kable(output_contract, caption = "SBHM output components, their dimensions, and their meaning for the fitted fixture.")
```

```{r output-contract-check, include=FALSE}
stopifnot(
  identical(dim(fit_hard$alpha_coords), c(rank_used, n_voxels)),
  identical(dim(fit_hard$amplitude), c(n_trials, n_voxels)),
  identical(dim(fit_hard$coeffs_r), c(rank_used, n_trials, n_voxels)),
  identical(colnames(fit_hard$alpha_coords), voxel_names),
  identical(dimnames(fit_hard$amplitude), list(trial_names, voxel_names)),
  identical(dimnames(fit_hard$coeffs_r)[2:3], list(trial_names, voxel_names)),
  identical(names(fit_hard$matched_name), voxel_names),
  identical(names(fit_hard$event_amplitude), trial_names),
  identical(names(fit_hard$event_duration), trial_names),
  identical(fit_hard$trial_basis_map$trial_name,
            rep(trial_names, each = rank_used))
)
```

## Evaluate the shape actually used

Whether the match picked the exact library index matters less than the
waveform it produced, because nearby candidates can have almost identical
waveforms. We therefore evaluate the waveform `B %*% alpha_coords` for every
voxel with two measures. Angular error is the angle between the estimated and
true waveforms; it is zero when they point in the same direction and ignores
any positive scale factor. Unit-norm waveform RMSE is the root-mean-square
difference between the two waveforms after each is scaled to unit norm, a
time-domain measure of the discrepancy.

```{r shape-metrics, include=FALSE}
waveform_metrics <- function(alpha_est, alpha_truth, B) {
  estimated <- B %*% alpha_est
  truth <- B %*% alpha_truth
  est_norm <- sweep(estimated, 2, sqrt(colSums(estimated^2)), "/")
  truth_norm <- sweep(truth, 2, sqrt(colSums(truth^2)), "/")
  cosine <- pmin(1, pmax(-1, colSums(est_norm * truth_norm)))
  data.frame(
    AngularErrorRadians = acos(cosine),
    UnitWaveformRMSE = sqrt(colMeans((est_norm - truth_norm)^2))
  )
}

hard_shape_by_voxel <- waveform_metrics(fit_hard$alpha_coords, alpha_true, sbhm$B)
shape_group <- rep(c("On-library", "Off-library"), each = 6)
shape_summary <- do.call(rbind, lapply(split(hard_shape_by_voxel, shape_group), function(x) {
  data.frame(
    MeanAngularErrorRadians = mean(x$AngularErrorRadians),
    MaxAngularErrorRadians = max(x$AngularErrorRadians),
    MeanUnitWaveformRMSE = mean(x$UnitWaveformRMSE)
  )
}))
shape_summary$Truth <- rownames(shape_summary)
rownames(shape_summary) <- NULL
shape_summary <- shape_summary[, c("Truth", setdiff(names(shape_summary), "Truth"))]
```

```{r shape-summary}
shape_summary_display <- shape_summary
shape_summary_display[-1] <- lapply(shape_summary_display[-1], round, 5)
knitr::kable(
  shape_summary_display,
  caption = "All-voxel hard-match waveform error, separated by library coverage."
)
```

```{r shape-plot-prep, include=FALSE}
true_waveforms <- sbhm$B %*% alpha_true
hard_waveforms <- sbhm$B %*% fit_hard$alpha_coords
shape_alt <- sprintf(
  paste0(
    "Twelve panels compare true solid and hard-matched dashed rank-four waveforms. ",
    "Mean angular error is %.3f radians on-library and %.3f radians off-library."
  ),
  shape_summary$MeanAngularErrorRadians[shape_summary$Truth == "On-library"],
  shape_summary$MeanAngularErrorRadians[shape_summary$Truth == "Off-library"]
)
```

```{r shape-plot, fig.width=10, fig.height=7, fig.alt=shape_alt, fig.cap="True and hard-matched rank-four waveforms for every voxel. Solid blue is truth; dashed orange is the waveform from the coordinates actually used by the amplitude refit. Panels marked '(off)' contain off-library mixtures and are intentionally harder."}
local({
  oldpar <- par(mfrow = c(3, 4), mar = c(2.7, 2.8, 2.1, 0.6), bg = "white")
  on.exit(par(oldpar), add = TRUE)
  keep <- time <= span
  for (v in seq_len(n_voxels)) {
    ylim <- range(true_waveforms[keep, v], hard_waveforms[keep, v])
    plot(time[keep], true_waveforms[keep, v], type = "l", lwd = 2,
         col = "#2c7fb8", ylim = ylim, xlab = "Seconds", ylab = "Shape",
         main = paste0(voxel_names[v], if (v <= 6) " (on)" else " (off)"))
    lines(time[keep], hard_waveforms[keep, v], lwd = 2, lty = 2, col = "#d95f02")
    abline(h = 0, col = "gray85")
  }
})
```

For the six on-library voxels, we can also check whether the match returned
the exact true candidate:

```{r index-secondary}
exact_index_accuracy <- mean(
  fit_hard$matched_name[1:6] == colnames(sbhm$A)[on_library_idx]
)
data.frame(OnLibraryExactCandidateAccuracy = exact_index_accuracy)
```

## Compare hard and soft matching

Soft matching blends the top-scoring candidates using softmax weights computed
from their cosine scores. Those weights are not mixture fractions, and the score
margin is not a calibrated confidence. Below, an explicit soft policy (top
three candidates) and the hard fit are both compared with the same truth. The
soft fit sets `blend_margin = 2`; because a margin is a difference of cosine
scores and cannot exceed 2, this blends every voxel. The package default,
`blend_margin = 0.08`, blends only voxels whose top two candidates score within
0.08 of each other. The
result describes this fixture only; it is not a rule that soft matching reduces
variance or improves interpolation.

```{r fit-soft}
fit_soft <- lss_sbhm(
  Y = Y,
  sbhm = sbhm,
  design_spec = design_spec,
  Nuisance = Nuisance,
  prepass = list(ridge = list(mode = "absolute", lambda = 0)),
  match = list(
    topK = 3, soft_blend = TRUE, blend_margin = 2,
    whiten = FALSE
  ),
  oasis = list(ridge_mode = "absolute", ridge_x = 0, ridge_b = 0),
  amplitude = list(method = "global_ls", ridge = 0),
  return = "both"
)
```

```{r compare-policies}
soft_shape_by_voxel <- waveform_metrics(fit_soft$alpha_coords, alpha_true, sbhm$B)
policy_table <- rbind(
  data.frame(
    Policy = "Hard top-1",
    VoxelsUsingPolicy = sum(fit_hard$shape_mode == "hard"),
    MeanAngularError = mean(hard_shape_by_voxel$AngularErrorRadians),
    AmplitudeBias = mean(fit_hard$amplitude - beta_true),
    AmplitudeRMSE = sqrt(mean((fit_hard$amplitude - beta_true)^2))
  ),
  data.frame(
    Policy = "Soft top-3",
    VoxelsUsingPolicy = sum(fit_soft$shape_mode == "soft"),
    MeanAngularError = mean(soft_shape_by_voxel$AngularErrorRadians),
    AmplitudeBias = mean(fit_soft$amplitude - beta_true),
    AmplitudeRMSE = sqrt(mean((fit_soft$amplitude - beta_true)^2))
  )
)
policy_table[-1] <- lapply(policy_table[-1], round, 5)
knitr::kable(
  policy_table,
  caption = "Truth-referenced hard and soft results for this fixed fixture."
)
```

```{r policy-check, include=FALSE}
stopifnot(
  all(fit_hard$shape_mode == "hard"),
  all(fit_soft$shape_mode == "soft")
)
```

The returned `margin` is the top cosine score minus the second-highest score.
A low margin can reveal near-ties, but it does not imply a
universal `min_margin` or `blend_margin` threshold. If you use gating thresholds,
inspect `shape_mode` and `fallback_low_conf`: `matched_name` remains the
top-scoring candidate, whereas `alpha_coords` records the shape actually used.

## Check the coefficient stage against a direct GLM

With `amplitude$method = "global_ls"` and zero ridge, the final stage is an
ordinary trial-wise GLM conditional on the selected shape. The reference
computation below (the oracle) builds every voxel's GLM independently from the
public `fmrihrf` regressors and fits it with `lm.fit()`. It confirms, on this
simulated dataset, that for the selected shapes SBHM returns the same
coefficients as the direct GLM to numerical precision. It does not remove
shape-selection bias.

```{r direct-oracle, include=FALSE}
direct_amplitude <- function(alpha) {
  out <- matrix(NA_real_, n_trials, n_voxels,
                dimnames = list(trial_names, voxel_names))
  common <- cbind(intercept = 1, Nuisance)
  for (v in seq_len(n_voxels)) {
    Xv <- do.call(cbind, lapply(
      trial_regs, function(Xt) as.numeric(Xt %*% alpha[, v])
    ))
    fit <- lm.fit(cbind(common, Xv), Y[, v])
    out[, v] <- tail(fit$coefficients, n_trials)
  }
  out
}
direct_hard <- direct_amplitude(fit_hard$alpha_coords)
known_shape <- direct_amplitude(alpha_true)
direct_error <- max(abs(direct_hard - fit_hard$amplitude))
amplitude_table <- data.frame(
  Estimator = c("Known truth shape", "SBHM hard-selected shape", "SBHM soft-selected shape"),
  Bias = c(
    mean(known_shape - beta_true),
    mean(fit_hard$amplitude - beta_true),
    mean(fit_soft$amplitude - beta_true)
  ),
  RMSE = c(
    sqrt(mean((known_shape - beta_true)^2)),
    sqrt(mean((fit_hard$amplitude - beta_true)^2)),
    sqrt(mean((fit_soft$amplitude - beta_true)^2))
  )
)
```

```{r direct-oracle-table}
amplitude_display <- amplitude_table
amplitude_display[-1] <- lapply(amplitude_display[-1], round, 5)
amplitude_display$DirectGLMMaxError <- c(NA, format(direct_error, scientific = TRUE), NA)
knitr::kable(
  amplitude_display,
  caption = "Trial-coefficient recovery and the independent all-cell GLM identity check."
)
```

```{r direct-oracle-check, include=FALSE}
stopifnot(
  direct_error < 1e-10,
  all(is.finite(as.matrix(amplitude_table[-1]))),
  identical(dimnames(fit_hard$amplitude), dimnames(beta_true))
)
```

```{r amplitude-plot-prep, include=FALSE}
amp_bias <- mean(fit_hard$amplitude - beta_true)
amp_rmse <- sqrt(mean((fit_hard$amplitude - beta_true)^2))
amp_alt <- sprintf(
  paste0(
    "Scatter of 96 true and hard-match event-design coefficients. ",
    "Bias is %.3f and RMSE is %.3f; a diagonal reference line is shown."
  ),
  amp_bias, amp_rmse
)
amp_range <- range(beta_true, fit_hard$amplitude)
```

```{r amplitude-plot, fig.alt=amp_alt, fig.cap="Hard-match event-design coefficients against truth for all trials and voxels. The red dashed line is equality; the displayed bias and RMSE are conditional on this fixed simulation."}
plot(
  as.vector(beta_true), as.vector(fit_hard$amplitude),
  pch = 19, col = grDevices::adjustcolor("#2c7fb8", alpha.f = 0.55),
  xlim = amp_range, ylim = amp_range,
  xlab = "True event-design coefficient",
  ylab = "Estimated event-design coefficient",
  main = sprintf("Hard matching: bias %.3f, RMSE %.3f", amp_bias, amp_rmse)
)
abline(0, 1, col = "#d7301f", lwd = 2, lty = 2)
grid()
```

## Factorization accelerates only the prepass

The prepass can compute its cross-products from a factorization of the data
rather than from `Y` itself. `data_fac` represents `Y` as `scores %*% loadings`,
where scores are \(T \times q\) and loadings are \(q \times V\), for \(T\) time
points, \(V\) voxels, and \(q\) factors. The full `Y` is still required by
`lss_sbhm()` for OASIS and the amplitude stage. The shortcut therefore reduces
only prepass multiplication; it does not make the full pipeline independent of
\(V\). Named factor axes must agree between scores and loadings; when `Y` has
voxel names, loadings must carry the same names and are aligned before use.
Requesting active prewhitening with `data_fac` stops with an error, because the
supplied factorization is not a factorization of the estimated whitened data.

```{r factorized-prepass}
set.seed(813)
q <- 4L
scores <- matrix(rnorm(n_time * q), n_time, q)
loadings <- matrix(rnorm(q * n_voxels), q, n_voxels)
colnames(scores) <- rownames(loadings) <- paste0("factor_", seq_len(q))
colnames(loadings) <- voxel_names
Y_factorized <- scores %*% loadings
colnames(Y_factorized) <- voxel_names

pre_dense <- sbhm_prepass(
  Y_factorized, sbhm, design_spec,
  ridge = list(mode = "absolute", lambda = 0)
)
pre_factorized <- sbhm_prepass(
  Y_factorized, sbhm, design_spec,
  ridge = list(mode = "absolute", lambda = 0),
  data_fac = list(scores = scores, loadings = loadings)
)
factorized_error <- max(abs(pre_dense$beta_bar - pre_factorized$beta_bar))
data.frame(
  Scores = paste(dim(scores), collapse = " x "),
  Loadings = paste(dim(loadings), collapse = " x "),
  Voxels = ncol(Y_factorized),
  DenseVsFactorizedMaxError = factorized_error
)
```

```{r factorized-check, include=FALSE}
stopifnot(q < n_voxels, factorized_error < 1e-10)
factorized_whitening_error <- try(
  sbhm_prepass(
    Y_factorized, sbhm, design_spec,
    prewhiten = list(method = "ar", p = 1L),
    data_fac = list(scores = scores, loadings = loadings)
  ),
  silent = TRUE
)
stopifnot(
  inherits(factorized_whitening_error, "try-error"),
  grepl("does not support active prewhitening", as.character(factorized_whitening_error), fixed = TRUE)
)
```

For a PCA factorization, `prcomp(Y)$x` supplies scores and
`t(prcomp(Y)$rotation)` supplies the required \(q \times V\) loading
orientation. The example below keeps `Y` as the response for OASIS and
the amplitude stage while using a rank-four uncentered PCA approximation only
for the prepass. This is a computational approximation, not an accuracy claim.

```{r pca-end-to-end}
pca <- prcomp(Y, center = FALSE, scale. = FALSE, rank. = q)
pca_scores <- pca$x[, seq_len(q), drop = FALSE]
pca_loadings <- t(pca$rotation[, seq_len(q), drop = FALSE])

fit_pca_prepass <- lss_sbhm(
  Y = Y,
  sbhm = sbhm,
  design_spec = design_spec,
  Nuisance = Nuisance,
  prepass = list(
    ridge = list(mode = "absolute", lambda = 0),
    data_fac = list(scores = pca_scores, loadings = pca_loadings)
  ),
  match = list(topK = 1, soft_blend = FALSE, whiten = FALSE),
  oasis = list(ridge_mode = "absolute", ridge_x = 0, ridge_b = 0),
  amplitude = list(method = "global_ls", ridge = 0),
  return = "amplitude"
)
data.frame(
  ScoreShape = paste(dim(pca_scores), collapse = " x "),
  LoadingShape = paste(dim(pca_loadings), collapse = " x "),
  FullResponseShape = paste(dim(Y), collapse = " x "),
  OutputShape = paste(dim(fit_pca_prepass$amplitude), collapse = " x ")
)
```

```{r pca-end-to-end-check, include=FALSE}
stopifnot(
  q < n_voxels,
  identical(dim(fit_pca_prepass$amplitude), c(n_trials, n_voxels)),
  all(is.finite(fit_pca_prepass$amplitude))
)
```

## Runs, conditions, and whitening

Raw multi-run `design_spec` inputs use run-relative onsets and an explicit
`cond$run` vector. SBHM constructs every trial inside its own sampling-frame
run and uses run-specific intercepts, so an HRF tail cannot cross a run
boundary. `lss_sbhm_design()` obtains those run identities and heterogeneous
durations from the unique trialwise `fmridesign` term; non-trial event terms and
baseline-model columns are included in the fixed nuisance regressors. See
`vignette("lss_with_fmridesign")` for the event-model identity contract.

If estimated prewhitening is requested, SBHM applies the same transformation to
the response, trial basis, other-condition span, and nuisance design in the
prepass, diagnostics, and final amplitude estimator. The whitening plan is
estimated once from the complete trial and common design, then reused at every
stage. For a multi-run sampling frame, omitted `prewhiten$runs` are filled from
the frame's run boundaries; contradictory labels fail rather than allowing a
temporal filter to bridge runs. This changes the conditional point estimator;
it does not create calibrated post-selection standard errors. Factorized
prepasses and active prewhitening cannot be combined.

## Advanced shape and amplitude options

`alpha_source = "trial_projection"` and `"oasis_rank1"`, low-score fallback,
condition gates, ridge penalties, and adaptive ridge controls are available for
specialized studies. They are heuristics, and they change the estimator. No
rank, ridge fraction, margin threshold, ISI boundary, or TR cutoff is generally
recommended here.
Choose and validate such a policy against the design, signal scale, failure
modes, and loss function of the intended analysis.

The three scalar-coefficient engines, selected with `amplitude$method`, also
fit different models:

- `global_ls` fits all trial columns jointly for each selected voxel shape;
- `lss1` fits one target and one aggregate-other column per trial;
- `oasis_voxel` applies the K=1 OASIS operator per voxel.

Exactly coincident or otherwise unidentified trials cannot be recovered by any
of these engines. A ridge penalty can make the fit return finite values for
such trials (identical values for coincident trials), but it does not make them
identifiable. None of the three engines returns calibrated SBHM
standard errors.

## Takeaways

SBHM is a constrained shape-selection and coefficient-estimation workflow. Its
useful compression comes with visible assumptions: the library defines the
shape space, rank truncation defines the approximation, matching or blending
defines the voxel waveform, and the supplied event design defines coefficient
units. Judge the waveform actually used (`alpha_coords`), keep the
`trial_basis_map` and voxel names that identify each output row and column, and evaluate hard, soft, fallback, ridge, and whitening policies against
truth or external calibration before using them in a scientific conclusion.

This is the end of the vignette learning path. Continue with `?sbhm_build`,
`?sbhm_match`, `?sbhm_prepass`, `?lss_sbhm`, and `?lss_sbhm_design` for the
complete API contracts.
