---
title: 'Voxel-wise HRF shapes and trial coefficients'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Voxel-wise HRF shapes and trial coefficients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

`estimate_voxel_hrf()` estimates one HRF shape per voxel; `lss_with_hrf()`
then uses those shapes to estimate one LSS coefficient per trial and voxel.
Because each shape is normalized to a peak of one, the coefficient for a
unit-amplitude impulse event is the trial's peak-response amplitude. Read this
after `vignette("fmrilss")`, `vignette("oasis_method")`, and
`vignette("lss_with_fmridesign")`.

```{r packages}
suppressPackageStartupMessages({
  library(fmrilss)
  library(fmrihrf)
})
```

## Fix the HRF scale before interpreting a beta

HRF shape and response amplitude are not jointly identified without a scale
convention: multiplying an HRF by \(c\) and dividing its beta by \(c\) leaves
the fitted signal unchanged. `estimate_voxel_hrf()` and `lss_with_hrf()`
resolve that ambiguity with the conventions in this table.

```{r contract-table, echo = FALSE}
contract <- data.frame(
  Object = c(
    "VoxelHRF coefficients",
    "VoxelHRF amplitude_scale",
    "lss_with_hrf() beta; unit-amplitude impulse",
    "lss_with_hrf() beta; nonzero duration or non-unit amplitude"
  ),
  Meaning = c(
    "oriented HRF-shape weights normalized to a positive peak of one",
    "scale removed from the pooled calibration fit",
    "trial response on the unit-peak HRF scale",
    "coefficient on the supplied duration- and amplitude-coded event design"
  ),
  Units = c(
    "unit-peak shape weights",
    "calibration-response scale",
    "peak-response amplitude",
    "event-design coefficient"
  )
)
knitr::kable(
  contract,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "The normalization contract for the public voxel-HRF workflow."
)
```

`amplitude_scale` lets you reconstruct the raw pooled-fit coefficients, but
it is not a trial beta. With unit-amplitude impulses, the trial beta from
`lss_with_hrf()` can be read as a peak response because the HRF shape it
receives has a positive peak of one.

Shapes are oriented to correlate positively with a reference HRF (`ref_hrf`,
the canonical HRF by default). Some voxels cannot be meaningfully
peak-normalized: those whose oriented shape has no positive peak, or a positive
peak smaller than 5% of its largest absolute deflection. Such a voxel typically
has no signal, and dividing by a noise-level peak would inflate its betas
arbitrarily. `estimate_voxel_hrf()` instead scales it by its largest absolute
response, flags it in `degenerate`, and counts it in a warning; the rest of the
fit is unaffected. `lss_with_hrf()` carries the flags as a `degenerate`
attribute, because those voxels' betas are not in peak-response units.
`lss_rank1()` follows the same policy.

## Separate shape calibration from trial estimation

If the shape and the trial betas are estimated from the same response, the
shape adapts to that response, which can make an in-sample comparison
optimistic. This example therefore uses two independent responses with the same
scan timing: a calibration response for estimating shapes and an analysis
response for estimating trial betas. The example demonstrates how the
functions work and what their outputs mean; it is not a population-level
performance study.

```{r experiment}
set.seed(20260822)

n_time <- 240L
n_voxels <- 4L
TR <- 1.2
sframe <- sampling_frame(blocklens = n_time, TR = TR)
end_time <- max(samples(sframe, global = TRUE))

calibration_events <- data.frame(
  onset = seq(10, end_time - 35, length.out = 24),
  duration = 0,
  condition = "calibration"
)
analysis_events <- data.frame(
  onset = seq(12, end_time - 35, length.out = 16),
  duration = 0,
  condition = "analysis"
)
```

### Simulate HRFs that differ in latency and width

Stretching an HRF's time axis (dilation) changes its width; multiplying it by a
scalar does not. The helper below evaluates the canonical HRF at
`(time - delay) / dilation`, then renormalizes the resulting waveform to unit
peak. Dilation therefore changes the response's width without changing the
amplitude being estimated.

```{r true-hrfs}
make_unit_peak_hrf <- function(delay, dilation) {
  dense_time <- seq(0, 30, by = 0.01)
  peak <- max(HRF_SPMG1((dense_time - delay) / dilation))
  HRF(
    fun = function(time) HRF_SPMG1((time - delay) / dilation) / peak,
    name = sprintf("delay_%s_width_%s", delay, dilation),
    span = 30,
    nbasis = 1L
  )
}

delays <- c(-0.4, -0.1, 0.2, 0.5)
dilations <- c(0.85, 0.95, 1.05, 1.15)
true_hrfs <- Map(make_unit_peak_hrf, delays, dilations)
voxel_names <- paste0("voxel_", seq_len(n_voxels))
```

```{r waveform-helpers, include = FALSE}
shape_time <- seq(0, 30, by = 0.05)
true_shapes <- vapply(
  true_hrfs, function(hrf) hrf(shape_time), numeric(length(shape_time))
)
colnames(true_shapes) <- voxel_names

fwhm <- function(waveform, time = shape_time) {
  inside <- which(waveform >= 0.5 * max(waveform))
  time[max(inside)] - time[min(inside)]
}

true_shape_summary <- data.frame(
  Voxel = voxel_names,
  Delay_seconds = delays,
  Dilation = dilations,
  Peak_height = apply(true_shapes, 2, max),
  Peak_time_seconds = apply(true_shapes, 2, function(x) shape_time[which.max(x)]),
  FWHM_seconds = apply(true_shapes, 2, fwhm)
)
stopifnot(
  max(abs(true_shape_summary$Peak_height - 1)) < 1e-8,
  all(diff(true_shape_summary$FWHM_seconds) > 0)
)
```

```{r true-shape-table, echo = FALSE}
knitr::kable(
  true_shape_summary,
  row.names = FALSE,
  digits = 3,
  caption = "The simulated shapes keep unit peak while peak time and FWHM change."
)
```

```{r true-shape-plot, echo = FALSE, fig.cap="The four simulated HRFs all peak at one, while later voxels have later and wider responses.", fig.alt=sprintf("Four unit-peak HRF curves. Peak times progress from %.2f to %.2f seconds and FWHM increases from %.2f to %.2f seconds across voxels 1 through 4.", min(true_shape_summary$Peak_time_seconds), max(true_shape_summary$Peak_time_seconds), min(true_shape_summary$FWHM_seconds), max(true_shape_summary$FWHM_seconds))}
local({
  old_par <- par(bg = "white", mar = c(4, 4, 2, 1))
  on.exit(par(old_par), add = TRUE)
  matplot(
    shape_time, true_shapes,
    type = "l", lty = 1, lwd = 2,
    col = hcl.colors(n_voxels, "Dark 3"),
    xlab = "Time after event (seconds)",
    ylab = "HRF value (unit peak)",
    ylim = range(0, true_shapes)
  )
  legend(
    "topright", legend = voxel_names,
    col = hcl.colors(n_voxels, "Dark 3"), lty = 1, lwd = 2,
    bty = "n"
  )
})
```

### Generate calibration and analysis responses

The calibration events all have amplitude one, so the pooled-shape model is
correctly specified. Analysis amplitudes are 2, 3, 4, and 5 across voxels and
constant across trials. Constant amplitudes keep every LSS model (the target
trial plus one regressor for all other trials) correctly specified, and values
well above one test whether amplitudes far from one are recovered.

```{r simulation-helpers}
trial_design <- function(events, hrf, sframe) {
  regressors <- regressor_set(
    onsets = events$onset,
    fac = factor(seq_len(nrow(events))),
    hrf = hrf,
    duration = events$duration,
    span = 30,
    summate = FALSE
  )
  as.matrix(evaluate(
    regressors,
    grid = samples(sframe, global = TRUE),
    precision = 0.1,
    method = "conv"
  ))
}

fixed_regs <- cbind(trend = seq(-1, 1, length.out = n_time))
nuisance_regs <- cbind(
  sine = sin(seq_len(n_time) / 13),
  cosine = cos(seq_len(n_time) / 17)
)
common_design <- cbind(Intercept = 1, fixed_regs, nuisance_regs)
```

```{r simulate-responses}
calibration_response <- common_design %*%
  matrix(rnorm(ncol(common_design) * n_voxels, sd = 0.2),
         ncol(common_design), n_voxels)
for (voxel in seq_len(n_voxels)) {
  calibration_response[, voxel] <- calibration_response[, voxel] +
    rowSums(trial_design(calibration_events, true_hrfs[[voxel]], sframe))
}
calibration_response <- calibration_response +
  matrix(rnorm(n_time * n_voxels, sd = 0.02), n_time, n_voxels)
colnames(calibration_response) <- voxel_names

true_amplitude <- c(2, 3, 4, 5)
analysis_response <- common_design %*%
  matrix(rnorm(ncol(common_design) * n_voxels, sd = 0.2),
         ncol(common_design), n_voxels)
true_trial_designs <- vector("list", n_voxels)
for (voxel in seq_len(n_voxels)) {
  true_trial_designs[[voxel]] <- trial_design(
    analysis_events, true_hrfs[[voxel]], sframe
  )
  analysis_response[, voxel] <- analysis_response[, voxel] +
    rowSums(true_trial_designs[[voxel]]) * true_amplitude[voxel]
}
analysis_response <- analysis_response +
  matrix(rnorm(n_time * n_voxels, sd = 0.02), n_time, n_voxels)
colnames(analysis_response) <- voxel_names
```

## Estimate normalized voxel HRFs

`estimate_voxel_hrf()` takes event onsets and durations in seconds, plus a
sampling frame (`sframe`) that defines the scan timing. `HRF_SPMG3` supplies canonical, temporal-derivative, and
dispersion-derivative basis functions; the returned weights define a
reconstructed curve, not three directly interpretable physiological
measurements.

The basis must also cover the whole response. fmrilss evaluates a basis only
over its declared span, and fmrihrf's built-in `HRF_SPMG1` and `HRF_SPMG3`
declare a 24-second span, so their post-stimulus undershoot is truncated at
24 seconds. A truncated basis cannot represent the late undershoot, and the
missing tail biases trial amplitudes, most for the widest responses. The
simulated responses here last 30 seconds, so the analysis rebuilds the same
bases with a 30-second span.

```{r full-span-basis}
spmg3_30 <- gen_hrf(HRF_SPMG3, span = 30)
spmg1_30 <- gen_hrf(HRF_SPMG1, span = 30)
c(builtin_span = attr(HRF_SPMG3, "span"), analysis_span = attr(spmg3_30, "span"))
```

```{r estimate-hrf}
estimated_hrf <- estimate_voxel_hrf(
  calibration_response,
  calibration_events,
  basis = spmg3_30,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe
)

relabeled_events <- calibration_events
relabeled_events$condition <- rep(c("A", "B"), length.out = nrow(relabeled_events))
relabeled_hrf <- estimate_voxel_hrf(
  calibration_response,
  relabeled_events,
  basis = spmg3_30,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe
)
relabeling_error <- max(abs(
  relabeled_hrf$coefficients - estimated_hrf$coefficients
))

c(
  class = class(estimated_hrf),
  coefficient_shape = paste(dim(estimated_hrf$coefficients), collapse = " x "),
  normalization = estimated_hrf$normalization,
  condition_pooling = estimated_hrf$condition_pooling,
  condition_relabeling_error = format(relabeling_error, scientific = TRUE)
)
```

`estimate_voxel_hrf()` pools all supplied events into one shape per voxel.
Condition labels are retained as metadata but do not request
condition-specific HRFs. The second fit above checks this in this example:
relabeling the events as alternating conditions `A` and `B` leaves the
coefficients unchanged (`condition_relabeling_error` is zero).

### Inspect reconstructed shapes

```{r estimated-shapes, include = FALSE}
basis_impulse <- regressor(
  onsets = 0, hrf = spmg3_30, duration = 0, span = 30
)
basis_shapes <- as.matrix(evaluate(
  basis_impulse, shape_time, precision = 0.05, method = "conv"
))
basis_shapes <- sweep(basis_shapes, 2, basis_shapes[1, ], "-")
estimated_shapes <- basis_shapes %*% estimated_hrf$coefficients

shape_summary <- data.frame(
  Voxel = voxel_names,
  Correlation = vapply(seq_len(n_voxels), function(voxel) {
    cor(true_shapes[, voxel], estimated_shapes[, voxel])
  }, numeric(1)),
  Peak_time_error_seconds = vapply(seq_len(n_voxels), function(voxel) {
    shape_time[which.max(estimated_shapes[, voxel])] -
      shape_time[which.max(true_shapes[, voxel])]
  }, numeric(1)),
  FWHM_error_seconds = vapply(seq_len(n_voxels), function(voxel) {
    fwhm(estimated_shapes[, voxel]) - fwhm(true_shapes[, voxel])
  }, numeric(1))
)
stopifnot(
  min(shape_summary$Correlation) > 0.95,
  max(abs(shape_summary$Peak_time_error_seconds)) < 0.5,
  max(abs(shape_summary$FWHM_error_seconds)) < 1
)
```

```{r shape-summary-table, echo = FALSE}
knitr::kable(
  shape_summary,
  row.names = FALSE,
  digits = 3,
  caption = "Reconstructed-shape agreement in the calibration simulation."
)
```

```{r estimated-shape-plot, echo = FALSE, fig.cap=sprintf("Estimated SPMG3 curves closely follow the independently generated unit-peak HRFs; the largest width discrepancy is %.2f seconds (%s).", max(abs(shape_summary$FWHM_error_seconds)), shape_summary$Voxel[which.max(abs(shape_summary$FWHM_error_seconds))]), fig.alt=sprintf("Four panels compare solid true and dashed estimated unit-peak HRFs. Shape correlations range from %.3f to %.3f; the largest width difference is %.2f seconds.", min(shape_summary$Correlation), max(shape_summary$Correlation), max(abs(shape_summary$FWHM_error_seconds)))}
local({
  old_par <- par(
    bg = "white", mfrow = c(2, 2),
    mar = c(2.5, 2.8, 2, 1), oma = c(3, 3.5, 0, 0),
    mgp = c(1.6, 0.5, 0), tcl = -0.3
  )
  on.exit(par(old_par), add = TRUE)
  for (voxel in seq_len(n_voxels)) {
    plot(
      shape_time, true_shapes[, voxel],
      type = "l", lwd = 2, col = "#1b1b1b",
      xlab = "", ylab = "",
      ylim = range(0, true_shapes[, voxel], estimated_shapes[, voxel]),
      main = voxel_names[voxel]
    )
    lines(shape_time, estimated_shapes[, voxel], lwd = 2, lty = 2,
          col = "#2166ac")
    legend("topright", c("true", "estimated"),
           col = c("#1b1b1b", "#2166ac"), lty = c(1, 2), lwd = 2,
           bty = "n", cex = 0.85)
  }
  mtext("Time after event (seconds)", side = 1, outer = TRUE, line = 1.5)
  mtext("HRF value (unit peak)", side = 2, outer = TRUE, line = 2)
})
```

## Estimate trial amplitudes with `lss_with_hrf()`

```{r fit-public}
fit_r <- lss_with_hrf(
  analysis_response,
  analysis_events,
  estimated_hrf,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe,
  engine = "R",
  verbose = FALSE
)

fit_cpp <- lss_with_hrf(
  analysis_response,
  analysis_events,
  estimated_hrf,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe,
  engine = "C++",
  chunk_size = 2,
  verbose = FALSE
)
dense_cpp <- as.matrix(fit_cpp)
```

With `engine = "R"`, `lss_with_hrf()` returns a dense matrix. With
`engine = "C++"`, it returns an `LSSBeta` object backed by `bigmemory`;
`as.matrix()` is the documented way to extract a dense matrix from it. Both
results record the engine actually used, so if the function falls back to a
different backend, the output shows it.

```{r output-contract}
engine_contract <- data.frame(
  Requested = c(attr(fit_r, "engine_requested"), fit_cpp$engine_requested),
  Used = c(attr(fit_r, "engine_used"), fit_cpp$engine_used),
  Returned_class = c(paste(class(fit_r), collapse = "/"), class(fit_cpp)[1]),
  Shape = c(paste(dim(fit_r), collapse = " x "),
            paste(dim(dense_cpp), collapse = " x ")),
  Units = c(attr(fit_r, "units"), attr(dense_cpp, "units"))
)
knitr::kable(
  engine_contract,
  row.names = FALSE,
  align = c("l", "l", "l", "l", "l"),
  caption = "Return types, realized engines, dimensions, and coefficient units."
)
```

```{r engine-contract-check, include = FALSE}
engine_error <- max(abs(fit_r - dense_cpp))
permuted_hrf <- estimated_hrf
permuted_hrf$coefficients <-
  permuted_hrf$coefficients[, rev(voxel_names), drop = FALSE]
permutation_fit <- lss_with_hrf(
  analysis_response, analysis_events, permuted_hrf,
  nuisance_regs = nuisance_regs, fixed_regs = fixed_regs,
  sframe = sframe, engine = "R", verbose = FALSE
)
stopifnot(
  engine_error < 1e-10,
  max(abs(permutation_fit - fit_r)) < 1e-10,
  identical(attr(dense_cpp, "normalization"), "positive-peak"),
  identical(attr(dense_cpp, "units"), "peak-response amplitude"),
  identical(attr(dense_cpp, "engine_used"), fit_cpp$engine_used),
  identical(attr(dense_cpp, "chunk_size"), 2L)
)
```

## Compare estimators on the same amplitude scale

The tables below compare three estimators. The reference (oracle) fits every
trial and voxel with the known unit-peak HRF and the same fixed and nuisance
regressors. The second is `lss_with_hrf()` with the estimated voxel HRFs. The
third is standard LSS with the canonical HRF, also scaled to unit peak. All
three therefore target peak-response amplitude; no derivative-basis
coefficient is relabeled as an amplitude.

```{r amplitude-oracles, include = FALSE}
truth <- matrix(
  rep(true_amplitude, each = nrow(analysis_events)),
  nrow = nrow(analysis_events),
  ncol = n_voxels,
  dimnames = dimnames(fit_r)
)

true_hrf_oracle <- matrix(NA_real_, nrow(analysis_events), n_voxels)
for (voxel in seq_len(n_voxels)) {
  X <- true_trial_designs[[voxel]]
  for (trial in seq_len(ncol(X))) {
    D <- cbind(
      Intercept = 1,
      fixed_regs,
      nuisance_regs,
      target = X[, trial],
      others = rowSums(X) - X[, trial]
    )
    true_hrf_oracle[trial, voxel] <-
      lm.fit(D, analysis_response[, voxel])$coefficients[ncol(D) - 1L]
  }
}

canonical_hrf <- make_unit_peak_hrf(0, 1)
canonical_design <- trial_design(analysis_events, canonical_hrf, sframe)
canonical_fit <- lss(
  analysis_response,
  canonical_design,
  Z = cbind(Intercept = 1, fixed_regs),
  Nuisance = nuisance_regs,
  method = "r_optimized"
)

rmse <- function(estimate, reference) {
  sqrt(mean((unname(estimate) - unname(reference))^2))
}
amplitude_comparison <- data.frame(
  Estimator = c(
    "known unit-peak HRF oracle",
    "estimated unit-peak voxel HRF",
    "unit-peak canonical HRF"
  ),
  RMSE_to_true_amplitude = c(
    rmse(true_hrf_oracle, truth),
    rmse(fit_r, truth),
    rmse(canonical_fit, truth)
  ),
  RMSE_to_known_HRF_oracle = c(
    0,
    rmse(fit_r, true_hrf_oracle),
    rmse(canonical_fit, true_hrf_oracle)
  )
)

voxel_amplitudes <- data.frame(
  Voxel = voxel_names,
  True_amplitude = true_amplitude,
  Mean_estimated_amplitude = colMeans(fit_r),
  Mean_error = colMeans(fit_r) - true_amplitude
)

worst_voxel <- which.max(abs(voxel_amplitudes$Mean_error / true_amplitude))
worst_relative_error <- voxel_amplitudes$Mean_error[worst_voxel] /
  true_amplitude[worst_voxel]
stopifnot(
  max(abs(true_hrf_oracle - truth)) < 0.05,
  # the estimated shape at least halves the canonical-HRF error ...
  amplitude_comparison$RMSE_to_true_amplitude[2] <
    0.5 * amplitude_comparison$RMSE_to_true_amplitude[3],
  # ... and recovers every voxel's absolute scale to within 10%
  abs(worst_relative_error) < 0.10
)
```

```{r amplitude-comparison-table, echo = FALSE}
knitr::kable(
  amplitude_comparison,
  row.names = FALSE,
  digits = 4,
  caption = "Amplitude errors of the three estimators, all on the peak-response-amplitude scale."
)
```

```{r voxel-amplitude-table, echo = FALSE}
knitr::kable(
  voxel_amplitudes,
  row.names = FALSE,
  digits = 4,
  caption = "Absolute scale recovery for amplitudes 2 through 5."
)
```

In this deliberately constructed four-voxel example, the estimated-shape fit
is closer to truth than the canonical fit. The largest remaining error is in
`r voxel_names[worst_voxel]`, whose mean amplitude is off by
`r sprintf("%.1f%%", 100 * worst_relative_error)` although its reconstructed
shape correlates `r sprintf("%.3f", shape_summary$Correlation[worst_voxel])`
with the truth. The likely reason is that the three SPMG3 functions
approximate a time-dilated HRF only locally, and LSS amplitudes are sensitive
to small shape errors, especially when neighbouring trial responses overlap.
With the true shape, the oracle recovers the amplitudes to within the noise
(RMSE `r sprintf("%.3f", amplitude_comparison$RMSE_to_true_amplitude[1])`), so
the remaining error comes from the estimated shape, not from trial
estimation.
These results describe this simulated example; they do not rank the methods in
general. Real performance depends on calibration quality, basis adequacy,
event timing, noise, and nuisance modeling.

## Check other TRs and run boundaries

The example above uses a TR of 1.2 seconds. The table below repeats a small,
correctly specified, noise-free experiment at TRs of 0.8 and 2 seconds, with a
one-function basis, some nonzero event durations, and two voxels whose true
coefficients are 2 and 4. Because of the nonzero durations, these values are
event-design coefficients rather than peak responses. The table also compares
the R output with chunked C++ output.

```{r tr-contract, include = FALSE}
check_tr <- function(TR) {
  set.seed(round(1000 * TR))
  n_time_check <- 140L
  sframe_check <- sampling_frame(blocklens = n_time_check, TR = TR)
  events_check <- data.frame(
    onset = c(0, 16.3, 38.7, 65.1, 92.4),
    duration = c(0, 1.2, 0, 2.1, 0.6),
    condition = "A"
  )
  hrf_check <- make_unit_peak_hrf(0, 1)
  X_check <- trial_design(events_check, hrf_check, sframe_check)
  truth_check <- cbind(rep(2, nrow(events_check)), rep(4, nrow(events_check)))
  fixed_check <- cbind(trend = seq(-1, 1, length.out = n_time_check))
  nuisance_check <- cbind(
    sine = sin(seq_len(n_time_check) / 11),
    cosine = cos(seq_len(n_time_check) / 17)
  )
  common_check <- cbind(1, fixed_check, nuisance_check)
  response_check <- X_check %*% truth_check +
    common_check %*% matrix(
      rnorm(ncol(common_check) * 2), ncol(common_check), 2
    )
  colnames(response_check) <- c("voxel_2", "voxel_4")
  estimate_check <- estimate_voxel_hrf(
    response_check, events_check, spmg1_30,
    nuisance_regs = nuisance_check, fixed_regs = fixed_check,
    sframe = sframe_check
  )
  r_check <- lss_with_hrf(
    response_check, events_check, estimate_check,
    nuisance_regs = nuisance_check, fixed_regs = fixed_check,
    engine = "R", verbose = FALSE
  )
  cpp_check <- as.matrix(lss_with_hrf(
    response_check, events_check, estimate_check,
    nuisance_regs = nuisance_check, fixed_regs = fixed_check,
    engine = "C++", chunk_size = 1, verbose = FALSE
  ))
  data.frame(
    TR_seconds = TR,
    RMSE_to_truth = rmse(r_check, truth_check),
    Maximum_engine_difference = max(abs(r_check - cpp_check)),
    Mean_coefficient_2 = mean(r_check[, 1]),
    Mean_coefficient_4 = mean(r_check[, 2])
  )
}

tr_contract <- do.call(rbind, lapply(c(0.8, 2), check_tr))
stopifnot(
  max(tr_contract$RMSE_to_truth) < 1e-3,
  max(tr_contract$Maximum_engine_difference) < 1e-10,
  max(abs(tr_contract$Mean_coefficient_2 - 2)) < 5e-4,
  max(abs(tr_contract$Mean_coefficient_4 - 4)) < 5e-4
)
```

```{r tr-contract-table, echo = FALSE}
knitr::kable(
  tr_contract,
  row.names = FALSE,
  digits = 12,
  caption = "Event-design coefficient recovery and R/C++ agreement at TR 0.8 and 2 seconds."
)
```

For multiple runs, add a `run` column of exact integer run indices and give
each event's onset relative to the start of its run. The trial design is built
separately within each run, so an HRF tail cannot leak across a run boundary.
The table below checks this for two runs of unequal length.

```{r run-boundary-contract, include = FALSE}
multi_sframe <- sampling_frame(blocklens = c(50L, 70L), TR = 1.2)
multi_events <- data.frame(
  onset = c(54, 6), duration = c(0, 1), condition = "A", run = c(1L, 2L)
)
multi_design <- fmrilss:::.voxhrf_trial_basis(
  multi_events, HRF_SPMG1, multi_sframe
)$X
multi_blocks <- fmrihrf::blockids(multi_sframe)
run_boundary <- data.frame(
  Check = c("run-1 event inside run 2", "run-2 event inside run 1"),
  Maximum_absolute_design_value = c(
    max(abs(multi_design[multi_blocks == 2, 1])),
    max(abs(multi_design[multi_blocks == 1, 2]))
  )
)
stopifnot(max(run_boundary$Maximum_absolute_design_value) < 1e-12)
```

```{r run-boundary-table, echo = FALSE}
knitr::kable(
  run_boundary,
  row.names = FALSE,
  digits = 12,
  caption = "Run-boundary check: largest design value that crosses into the other run, for two runs of unequal length."
)
```

## Assumptions and limitations

- `condition` labels are metadata in this API. All supplied events estimate
  one pooled HRF shape per voxel. Fit separate calibration subsets only when
  distinct condition-specific shapes are scientifically justified.
- The calibration model assumes one pooled event amplitude per voxel. Strongly
  heterogeneous calibration amplitudes can distort the estimated shape.
- The returned trial betas do not include uncertainty for estimated HRF
  shapes. Use them as point estimates conditional on those shapes; this
  vignette makes no claim about their standard errors or \(t\) reference
  distributions. `estimate_voxel_hrf()` and `lss_with_hrf()` also do not
  implement prewhitening with an estimated noise model.
- Pass the same fixed and nuisance regressors, those your design requires,
  to both shape estimation and trial fitting. The functions add run intercepts
  when they are absent. If the column space of these common regressors,
  including the intercepts, contains the pooled HRF basis, the shape is not
  identifiable and estimation fails explicitly.
- Run-aware convolution prevents an HRF tail from crossing a run boundary, but
  in a joint LSS fit the aggregate other-trials regressor still sums trials
  from all runs. When the estimand must be run-local, fit runs separately, then
  map each trial back to its original event.
- Event onsets and durations are finite times in seconds. Empty event tables,
  out-of-run onsets, malformed run indices, non-finite coefficients, and
  incomplete or duplicated voxel names fail explicitly.
- The main example uses unit-amplitude impulses, for which the coefficient is a
  peak response. With a nonzero duration or non-unit `events$amplitude`, the
  result is a coefficient on the supplied event design; `units`,
  `event_duration`, and `event_amplitude` record that change. For a
  zero-duration event, multiplying by event amplitude recovers the modeled
  peak. For a nonzero duration, evaluate the duration-coded regressor when a
  peak conversion is needed. A zero-amplitude trial is not identifiable and
  fails explicitly.
- OASIS with a multi-basis HRF returns \(K\) coefficients per trial. Those
  coefficients are not interchangeable with the scalar peak-response
  amplitudes above; retain the full basis result unless you define and verify a
  separate shape-to-amplitude reduction.

## Next step

`vignette("rank1_hrf")` estimates each voxel's HRF jointly with the trial
amplitudes (the rank-1 GLM of Pedregosa et al., 2015) instead of from one
pooled amplitude, which matters when conditions differ in their mean response.
`vignette("sbhm")` presents a library-constrained alternative for estimating
voxel-specific shapes, score margins (how far the best library shape scores
above the runner-up), and trial event-design coefficients.
