---
title: "Voxel-wise HRF Modeling"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voxel-wise HRF Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates voxel-wise HRF estimation and LSS analysis using
`fmrilss`. The workflow has two stages:

1. estimate HRF basis coefficients for each voxel.
2. compute trial-wise betas using those HRFs.

We will simulate a small dataset to show the process.

## 1. Simulate Data

```{r simulate}
library(fmrilss)
library(fmrihrf)

set.seed(123)
n_time <- 120
events <- data.frame(
  onset = c(10, 40, 70, 100),
  duration = 1,
  condition = "A"
)

basis <- fmrihrf::HRF_FIR(length = 8)
X_basis <- fmrihrf::regressor_set(events, basis = basis, n = n_time)$X
n_vox <- 3
true_coef <- matrix(rnorm(ncol(X_basis) * n_vox, 0, 1), ncol(X_basis), n_vox)
Y <- X_basis %*% true_coef + matrix(rnorm(n_time * n_vox, 0, 0.1), n_time, n_vox)
```

## 2. Estimate Voxel-wise HRFs

```{r estimate-hrf}
hrf_est <- estimate_voxel_hrf(Y, events, basis)
hrf_est
```

## 3. Inspect HRF Shapes

```{r hrf-plots}
time_pts <- seq(0, fmrihrf::hrf_length(basis), by = 1)
hrf_kernels <- apply(hrf_est$coefficients, 2, function(co) {
  fmrihrf::hrf_from_coefficients(basis, co)
})
matplot(time_pts, hrf_kernels, type = "l", lty = 1,
        xlab = "Time", ylab = "Amplitude",
        main = "Estimated HRFs")
legend("topright", legend = paste0("Voxel ", 1:n_vox),
       col = 1:n_vox, lty = 1)

# Time-to-peak for each voxel
ttp <- apply(hrf_kernels, 2, function(h) {
  (which.max(h) - 1)
})
barplot(ttp, names.arg = paste0("V", 1:n_vox),
        ylab = "Time to peak (TR)", main = "HRF Time-to-Peak")
```

## 4. Run LSS Using Voxel-wise HRFs

```{r lss-run}
lss_res <- lss_with_hrf(Y, events, hrf_est, verbose = FALSE, chunk_size = n_vox)
betas <- as.matrix(lss_res$betas[])
head(betas)
```

## 5. Analyze Trial Betas

```{r beta-analysis}
matplot(betas, type = "l", lty = 1,
        xlab = "Trial", ylab = "Beta",
        main = "Trial-wise Betas by Voxel")
legend("topright", legend = paste0("Voxel ", 1:n_vox),
       col = 1:n_vox, lty = 1)
```

This simple example shows how voxel-wise HRF modeling can be
integrated into an LSS analysis pipeline.
