---
title: 'Voxel-wise HRF shapes and trial coefficients'
output:
  rmarkdown::html_vignette:
    mathjax: default
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Voxel-wise HRF shapes and trial coefficients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  warning = TRUE,
  message = TRUE
)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

`estimate_voxel_hrf()` estimates one HRF shape per voxel; `lss_with_hrf()`
then estimates one LSS coefficient per trial and voxel. For unit-amplitude
impulses, unit-peak HRF normalization makes that coefficient a peak-response
amplitude. Read this after
`vignette("fmrilss")`, `vignette("oasis_method")`, and
`vignette("lss_with_fmridesign")`.

```{r packages}
suppressPackageStartupMessages({
  library(fmrilss)
  library(fmrihrf)
})
```

## Fix the scale before interpreting a beta

HRF shape and response amplitude are not jointly identified without a scale
convention: multiplying an HRF by \(c\) and dividing its beta by \(c\) leaves
the fitted signal unchanged. The public workflow resolves that ambiguity as
follows.

```{r contract-table, echo = FALSE}
contract <- data.frame(
  Object = c(
    "VoxelHRF coefficients",
    "VoxelHRF amplitude_scale",
    "lss_with_hrf() beta; unit-amplitude impulse",
    "lss_with_hrf() beta; nonzero duration or non-unit amplitude"
  ),
  Meaning = c(
    "oriented HRF-shape weights normalized to a positive peak of one",
    "scale removed from the pooled calibration fit",
    "trial response on the unit-peak HRF scale",
    "coefficient on the supplied duration- and amplitude-coded event design"
  ),
  Units = c(
    "unit-peak shape weights",
    "calibration-response scale",
    "peak-response amplitude",
    "event-design coefficient"
  )
)
knitr::kable(
  contract,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "The normalization contract for the public voxel-HRF workflow."
)
```

`amplitude_scale` is useful for reconstructing the raw pooled-fit
coefficients, but it is not a trial beta. With unit-amplitude impulses, the
downstream beta is interpretable as a peak response because the supplied HRF
shape has unit positive peak.

## Separate shape calibration from trial estimation

Estimating a shape and trial betas from the same response is adaptive and can
make an in-sample comparison optimistic. This example uses independent
calibration and analysis responses with the same scan timing. It is a
mechanism and contract demonstration, not a population-level performance
study.

```{r experiment}
set.seed(20260822)

n_time <- 240L
n_voxels <- 4L
TR <- 1.2
sframe <- sampling_frame(blocklens = n_time, TR = TR)
end_time <- max(samples(sframe, global = TRUE))

calibration_events <- data.frame(
  onset = seq(10, end_time - 35, length.out = 24),
  duration = 0,
  condition = "calibration"
)
analysis_events <- data.frame(
  onset = seq(12, end_time - 35, length.out = 16),
  duration = 0,
  condition = "analysis"
)
```

### Simulate genuine latency and width differences

Time dilation changes width; scalar multiplication does not. The helper below
evaluates the canonical HRF at `(time - delay) / dilation`, then renormalizes
the resulting waveform to unit peak. Dilation therefore changes duration
without changing the amplitude estimand.

```{r true-hrfs}
make_unit_peak_hrf <- function(delay, dilation) {
  dense_time <- seq(0, 30, by = 0.01)
  peak <- max(HRF_SPMG1((dense_time - delay) / dilation))
  HRF(
    fun = function(time) HRF_SPMG1((time - delay) / dilation) / peak,
    name = sprintf("delay_%s_width_%s", delay, dilation),
    span = 30,
    nbasis = 1L
  )
}

delays <- c(-0.4, -0.1, 0.2, 0.5)
dilations <- c(0.85, 0.95, 1.05, 1.15)
true_hrfs <- Map(make_unit_peak_hrf, delays, dilations)
voxel_names <- paste0("voxel_", seq_len(n_voxels))
```

```{r waveform-helpers, include = FALSE}
shape_time <- seq(0, 30, by = 0.05)
true_shapes <- vapply(
  true_hrfs, function(hrf) hrf(shape_time), numeric(length(shape_time))
)
colnames(true_shapes) <- voxel_names

fwhm <- function(waveform, time = shape_time) {
  inside <- which(waveform >= 0.5 * max(waveform))
  time[max(inside)] - time[min(inside)]
}

true_shape_summary <- data.frame(
  Voxel = voxel_names,
  Delay_seconds = delays,
  Dilation = dilations,
  Peak_height = apply(true_shapes, 2, max),
  Peak_time_seconds = apply(true_shapes, 2, function(x) shape_time[which.max(x)]),
  FWHM_seconds = apply(true_shapes, 2, fwhm)
)
stopifnot(
  max(abs(true_shape_summary$Peak_height - 1)) < 1e-8,
  all(diff(true_shape_summary$FWHM_seconds) > 0)
)
```

```{r true-shape-table, echo = FALSE}
knitr::kable(
  true_shape_summary,
  row.names = FALSE,
  digits = 3,
  caption = "The simulated shapes keep unit peak while peak time and FWHM change."
)
```

```{r true-shape-plot, echo = FALSE, fig.cap="The four simulated HRFs all peak at one, while later voxels have later and wider responses.", fig.alt="Four unit-peak HRF curves. Peak times progress from 3.85 to 6.25 seconds and FWHM increases from 4.45 to 6.05 seconds across voxels 1 through 4."}
local({
  old_par <- par(bg = "white", mar = c(4, 4, 2, 1))
  on.exit(par(old_par), add = TRUE)
  matplot(
    shape_time, true_shapes,
    type = "l", lty = 1, lwd = 2,
    col = hcl.colors(n_voxels, "Dark 3"),
    xlab = "Time after event (seconds)",
    ylab = "HRF value (unit peak)",
    ylim = range(0, true_shapes)
  )
  legend(
    "topright", legend = voxel_names,
    col = hcl.colors(n_voxels, "Dark 3"), lty = 1, lwd = 2,
    bty = "n"
  )
})
```

### Generate calibration and analysis responses

The calibration events all have amplitude one, so the pooled-shape model is
correctly specified. Analysis amplitudes are 2, 3, 4, and 5 across voxels and
constant across trials; this keeps every LSS target-plus-other-trials model
correctly specified while testing scale far away from one.

```{r simulation-helpers}
trial_design <- function(events, hrf, sframe) {
  regressors <- regressor_set(
    onsets = events$onset,
    fac = factor(seq_len(nrow(events))),
    hrf = hrf,
    duration = events$duration,
    span = 30,
    summate = FALSE
  )
  as.matrix(evaluate(
    regressors,
    grid = samples(sframe, global = TRUE),
    precision = 0.1,
    method = "conv"
  ))
}

fixed_regs <- cbind(trend = seq(-1, 1, length.out = n_time))
nuisance_regs <- cbind(
  sine = sin(seq_len(n_time) / 13),
  cosine = cos(seq_len(n_time) / 17)
)
common_design <- cbind(Intercept = 1, fixed_regs, nuisance_regs)
```

```{r simulate-responses}
calibration_response <- common_design %*%
  matrix(rnorm(ncol(common_design) * n_voxels, sd = 0.2),
         ncol(common_design), n_voxels)
for (voxel in seq_len(n_voxels)) {
  calibration_response[, voxel] <- calibration_response[, voxel] +
    rowSums(trial_design(calibration_events, true_hrfs[[voxel]], sframe))
}
calibration_response <- calibration_response +
  matrix(rnorm(n_time * n_voxels, sd = 0.02), n_time, n_voxels)
colnames(calibration_response) <- voxel_names

true_amplitude <- c(2, 3, 4, 5)
analysis_response <- common_design %*%
  matrix(rnorm(ncol(common_design) * n_voxels, sd = 0.2),
         ncol(common_design), n_voxels)
true_trial_designs <- vector("list", n_voxels)
for (voxel in seq_len(n_voxels)) {
  true_trial_designs[[voxel]] <- trial_design(
    analysis_events, true_hrfs[[voxel]], sframe
  )
  analysis_response[, voxel] <- analysis_response[, voxel] +
    rowSums(true_trial_designs[[voxel]]) * true_amplitude[voxel]
}
analysis_response <- analysis_response +
  matrix(rnorm(n_time * n_voxels, sd = 0.02), n_time, n_voxels)
colnames(analysis_response) <- voxel_names
```

## Estimate normalized voxel HRFs

The public estimator takes physical-time onsets and durations plus an explicit
sampling frame. `HRF_SPMG3` supplies canonical, temporal-derivative, and
dispersion-derivative basis functions; the returned weights define a
reconstructed curve, not three directly interpretable physiological
measurements.

```{r estimate-hrf}
estimated_hrf <- estimate_voxel_hrf(
  calibration_response,
  calibration_events,
  basis = HRF_SPMG3,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe
)

relabeled_events <- calibration_events
relabeled_events$condition <- rep(c("A", "B"), length.out = nrow(relabeled_events))
relabeled_hrf <- estimate_voxel_hrf(
  calibration_response,
  relabeled_events,
  basis = HRF_SPMG3,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe
)
relabeling_error <- max(abs(
  relabeled_hrf$coefficients - estimated_hrf$coefficients
))

c(
  class = class(estimated_hrf),
  coefficient_shape = paste(dim(estimated_hrf$coefficients), collapse = " x "),
  normalization = estimated_hrf$normalization,
  condition_pooling = estimated_hrf$condition_pooling,
  condition_relabeling_error = format(relabeling_error, scientific = TRUE)
)
```

The estimator pools all supplied events into one shape per voxel. Condition
labels are retained as metadata but do not request condition-specific HRFs;
the zero relabeling discrepancy above binds that contract in this example.

### Inspect reconstructed shapes

```{r estimated-shapes, include = FALSE}
basis_impulse <- regressor(
  onsets = 0, hrf = HRF_SPMG3, duration = 0, span = 30
)
basis_shapes <- as.matrix(evaluate(
  basis_impulse, shape_time, precision = 0.05, method = "conv"
))
basis_shapes <- sweep(basis_shapes, 2, basis_shapes[1, ], "-")
estimated_shapes <- basis_shapes %*% estimated_hrf$coefficients

shape_summary <- data.frame(
  Voxel = voxel_names,
  Correlation = vapply(seq_len(n_voxels), function(voxel) {
    cor(true_shapes[, voxel], estimated_shapes[, voxel])
  }, numeric(1)),
  Peak_time_error_seconds = vapply(seq_len(n_voxels), function(voxel) {
    shape_time[which.max(estimated_shapes[, voxel])] -
      shape_time[which.max(true_shapes[, voxel])]
  }, numeric(1)),
  FWHM_error_seconds = vapply(seq_len(n_voxels), function(voxel) {
    fwhm(estimated_shapes[, voxel]) - fwhm(true_shapes[, voxel])
  }, numeric(1))
)
stopifnot(
  min(shape_summary$Correlation) > 0.95,
  max(abs(shape_summary$Peak_time_error_seconds)) < 0.5,
  max(abs(shape_summary$FWHM_error_seconds)) < 1
)
```

```{r shape-summary-table, echo = FALSE}
knitr::kable(
  shape_summary,
  row.names = FALSE,
  digits = 3,
  caption = "Reconstructed-shape agreement in the calibration simulation."
)
```

```{r estimated-shape-plot, echo = FALSE, fig.cap="Estimated SPMG3 curves closely follow the independently generated unit-peak HRFs; the largest discrepancy is the width of voxel 4.", fig.alt="Four panels compare solid true and dashed estimated unit-peak HRFs. Shape correlations range from about 0.99 to 1.00; voxel 4 has the largest width difference, about 0.85 seconds."}
local({
  old_par <- par(
    bg = "white", mfrow = c(2, 2),
    mar = c(2.5, 2.8, 2, 1), oma = c(3, 3.5, 0, 0),
    mgp = c(1.6, 0.5, 0), tcl = -0.3
  )
  on.exit(par(old_par), add = TRUE)
  for (voxel in seq_len(n_voxels)) {
    plot(
      shape_time, true_shapes[, voxel],
      type = "l", lwd = 2, col = "#1b1b1b",
      xlab = "", ylab = "",
      ylim = range(0, true_shapes[, voxel], estimated_shapes[, voxel]),
      main = voxel_names[voxel]
    )
    lines(shape_time, estimated_shapes[, voxel], lwd = 2, lty = 2,
          col = "#2166ac")
    legend("topright", c("true", "estimated"),
           col = c("#1b1b1b", "#2166ac"), lty = c(1, 2), lwd = 2,
           bty = "n", cex = 0.85)
  }
  mtext("Time after event (seconds)", side = 1, outer = TRUE, line = 1.5)
  mtext("HRF value (unit peak)", side = 2, outer = TRUE, line = 2)
})
```

## Estimate trial amplitudes with the public API

```{r fit-public}
fit_r <- lss_with_hrf(
  analysis_response,
  analysis_events,
  estimated_hrf,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe,
  engine = "R",
  verbose = FALSE
)

fit_cpp <- lss_with_hrf(
  analysis_response,
  analysis_events,
  estimated_hrf,
  nuisance_regs = nuisance_regs,
  fixed_regs = fixed_regs,
  sframe = sframe,
  engine = "C++",
  chunk_size = 2,
  verbose = FALSE
)
dense_cpp <- as.matrix(fit_cpp)
```

The R engine returns a dense matrix. The C++ request returns an `LSSBeta`
object backed by `bigmemory`; `as.matrix()` is its documented dense extraction
path. Both paths report the engine actually used, so a backend fallback is not
silent.

```{r output-contract}
engine_contract <- data.frame(
  Requested = c(attr(fit_r, "engine_requested"), fit_cpp$engine_requested),
  Used = c(attr(fit_r, "engine_used"), fit_cpp$engine_used),
  Returned_class = c(paste(class(fit_r), collapse = "/"), class(fit_cpp)[1]),
  Shape = c(paste(dim(fit_r), collapse = " x "),
            paste(dim(dense_cpp), collapse = " x ")),
  Units = c(attr(fit_r, "units"), attr(dense_cpp, "units"))
)
knitr::kable(
  engine_contract,
  row.names = FALSE,
  align = c("l", "l", "l", "l", "l"),
  caption = "Return types, realized engines, dimensions, and coefficient units."
)
```

```{r engine-contract-check, include = FALSE}
engine_error <- max(abs(fit_r - dense_cpp))
permuted_hrf <- estimated_hrf
permuted_hrf$coefficients <-
  permuted_hrf$coefficients[, rev(voxel_names), drop = FALSE]
permutation_fit <- lss_with_hrf(
  analysis_response, analysis_events, permuted_hrf,
  nuisance_regs = nuisance_regs, fixed_regs = fixed_regs,
  sframe = sframe, engine = "R", verbose = FALSE
)
stopifnot(
  engine_error < 1e-10,
  max(abs(permutation_fit - fit_r)) < 1e-10,
  identical(attr(dense_cpp, "normalization"), "positive-peak"),
  identical(attr(dense_cpp, "units"), "peak-response amplitude"),
  identical(attr(dense_cpp, "engine_used"), fit_cpp$engine_used),
  identical(attr(dense_cpp, "chunk_size"), 2L)
)
```

## Compare only estimators of the same amplitude

The reference below fits every trial and voxel with the known unit-peak HRF,
the same fixed regressors, and the same nuisance span. A canonical comparator
also uses a unit-peak HRF. All rows therefore target peak-response amplitude;
no derivative-basis coefficient is relabeled as an amplitude.

```{r amplitude-oracles, include = FALSE}
truth <- matrix(
  rep(true_amplitude, each = nrow(analysis_events)),
  nrow = nrow(analysis_events),
  ncol = n_voxels,
  dimnames = dimnames(fit_r)
)

true_hrf_oracle <- matrix(NA_real_, nrow(analysis_events), n_voxels)
for (voxel in seq_len(n_voxels)) {
  X <- true_trial_designs[[voxel]]
  for (trial in seq_len(ncol(X))) {
    D <- cbind(
      Intercept = 1,
      fixed_regs,
      nuisance_regs,
      target = X[, trial],
      others = rowSums(X) - X[, trial]
    )
    true_hrf_oracle[trial, voxel] <-
      lm.fit(D, analysis_response[, voxel])$coefficients[ncol(D) - 1L]
  }
}

canonical_hrf <- make_unit_peak_hrf(0, 1)
canonical_design <- trial_design(analysis_events, canonical_hrf, sframe)
canonical_fit <- lss(
  analysis_response,
  canonical_design,
  Z = cbind(Intercept = 1, fixed_regs),
  Nuisance = nuisance_regs,
  method = "r_optimized"
)

rmse <- function(estimate, reference) {
  sqrt(mean((unname(estimate) - unname(reference))^2))
}
amplitude_comparison <- data.frame(
  Estimator = c(
    "known unit-peak HRF oracle",
    "estimated unit-peak voxel HRF",
    "unit-peak canonical HRF"
  ),
  RMSE_to_true_amplitude = c(
    rmse(true_hrf_oracle, truth),
    rmse(fit_r, truth),
    rmse(canonical_fit, truth)
  ),
  RMSE_to_known_HRF_oracle = c(
    0,
    rmse(fit_r, true_hrf_oracle),
    rmse(canonical_fit, true_hrf_oracle)
  )
)

voxel_amplitudes <- data.frame(
  Voxel = voxel_names,
  True_amplitude = true_amplitude,
  Mean_estimated_amplitude = colMeans(fit_r),
  Mean_error = colMeans(fit_r) - true_amplitude
)

stopifnot(
  max(abs(true_hrf_oracle - truth)) < 0.05,
  amplitude_comparison$RMSE_to_true_amplitude[2] < 0.15,
  amplitude_comparison$RMSE_to_true_amplitude[2] <
    amplitude_comparison$RMSE_to_true_amplitude[3]
)
```

```{r amplitude-comparison-table, echo = FALSE}
knitr::kable(
  amplitude_comparison,
  row.names = FALSE,
  digits = 4,
  caption = "A fair comparison on the common peak-response-amplitude scale."
)
```

```{r voxel-amplitude-table, echo = FALSE}
knitr::kable(
  voxel_amplitudes,
  row.names = FALSE,
  digits = 4,
  caption = "Absolute scale recovery for amplitudes 2 through 5."
)
```

In this deliberately constructed four-voxel example, the estimated-shape fit
is closer to truth than the canonical fit. That result diagnoses this fixture;
it is not a universal method ranking. Real performance depends on calibration
quality, basis adequacy, event timing, noise, and nuisance modeling.

## Check physical time and run boundaries

The same public pipeline also runs at non-unit TR. The table below repeats a
correctly specified one-basis coefficient experiment at TR 0.8 and 2 seconds,
including nonzero durations, with true coefficients 2 and 4. These rows are
event-design coefficients rather than peak responses. The court also compares
R with chunked C++ output.

```{r tr-contract, include = FALSE}
check_tr <- function(TR) {
  set.seed(round(1000 * TR))
  n_time_check <- 140L
  sframe_check <- sampling_frame(blocklens = n_time_check, TR = TR)
  events_check <- data.frame(
    onset = c(0, 16.3, 38.7, 65.1, 92.4),
    duration = c(0, 1.2, 0, 2.1, 0.6),
    condition = "A"
  )
  hrf_check <- make_unit_peak_hrf(0, 1)
  X_check <- trial_design(events_check, hrf_check, sframe_check)
  truth_check <- cbind(rep(2, nrow(events_check)), rep(4, nrow(events_check)))
  fixed_check <- cbind(trend = seq(-1, 1, length.out = n_time_check))
  nuisance_check <- cbind(
    sine = sin(seq_len(n_time_check) / 11),
    cosine = cos(seq_len(n_time_check) / 17)
  )
  common_check <- cbind(1, fixed_check, nuisance_check)
  response_check <- X_check %*% truth_check +
    common_check %*% matrix(
      rnorm(ncol(common_check) * 2), ncol(common_check), 2
    )
  colnames(response_check) <- c("voxel_2", "voxel_4")
  estimate_check <- estimate_voxel_hrf(
    response_check, events_check, HRF_SPMG1,
    nuisance_regs = nuisance_check, fixed_regs = fixed_check,
    sframe = sframe_check
  )
  r_check <- lss_with_hrf(
    response_check, events_check, estimate_check,
    nuisance_regs = nuisance_check, fixed_regs = fixed_check,
    engine = "R", verbose = FALSE
  )
  cpp_check <- as.matrix(lss_with_hrf(
    response_check, events_check, estimate_check,
    nuisance_regs = nuisance_check, fixed_regs = fixed_check,
    engine = "C++", chunk_size = 1, verbose = FALSE
  ))
  data.frame(
    TR_seconds = TR,
    RMSE_to_truth = rmse(r_check, truth_check),
    Maximum_engine_difference = max(abs(r_check - cpp_check)),
    Mean_coefficient_2 = mean(r_check[, 1]),
    Mean_coefficient_4 = mean(r_check[, 2])
  )
}

tr_contract <- do.call(rbind, lapply(c(0.8, 2), check_tr))
stopifnot(
  max(tr_contract$RMSE_to_truth) < 1e-3,
  max(tr_contract$Maximum_engine_difference) < 1e-10,
  max(abs(tr_contract$Mean_coefficient_2 - 2)) < 5e-4,
  max(abs(tr_contract$Mean_coefficient_4 - 4)) < 5e-4
)
```

```{r tr-contract-table, echo = FALSE}
knitr::kable(
  tr_contract,
  row.names = FALSE,
  digits = 12,
  caption = "Non-unit-TR event-design coefficient and R/C++ identity court."
)
```

For multiple runs, add an exact-integer `run` column and keep event onsets
relative to their run. The design is built separately inside each run, so an
HRF tail cannot leak across an acquisition boundary.

```{r run-boundary-contract, include = FALSE}
multi_sframe <- sampling_frame(blocklens = c(50L, 70L), TR = 1.2)
multi_events <- data.frame(
  onset = c(54, 6), duration = c(0, 1), condition = "A", run = c(1L, 2L)
)
multi_design <- fmrilss:::.voxhrf_trial_basis(
  multi_events, HRF_SPMG1, multi_sframe
)$X
multi_blocks <- fmrihrf::blockids(multi_sframe)
run_boundary <- data.frame(
  Check = c("run-1 event inside run 2", "run-2 event inside run 1"),
  Maximum_absolute_design_value = c(
    max(abs(multi_design[multi_blocks == 2, 1])),
    max(abs(multi_design[multi_blocks == 1, 2]))
  )
)
stopifnot(max(run_boundary$Maximum_absolute_design_value) < 1e-12)
```

```{r run-boundary-table, echo = FALSE}
knitr::kable(
  run_boundary,
  row.names = FALSE,
  digits = 12,
  caption = "Unequal-run HRF convolution boundary court."
)
```

## Know the boundaries of the workflow

- `condition` labels are metadata in this API. All supplied events estimate
  one pooled HRF shape per voxel. Fit separate calibration subsets only when
  distinct condition-specific shapes are scientifically justified.
- The calibration model assumes one pooled event amplitude per voxel. Strongly
  heterogeneous calibration amplitudes can distort the estimated shape.
- The returned trial betas do not include uncertainty for estimated HRF
  shapes. Use them as conditional point estimates; this article makes no
  standard-error or \(t\)-reference claim. These helpers also do not implement
  an estimated-prewhitening path.
- Pass the same scientifically required fixed and nuisance spans to shape
  estimation and trial fitting. The functions add run intercepts when they are
  absent. If that complete common span contains the pooled HRF basis, shape is
  not identifiable and estimation fails explicitly.
- Run-aware convolution prevents an HRF tail from crossing a run boundary, but
  a joint LSS fit still pools the other-trial aggregate across runs. Fit runs
  separately, then restore global event identities, when the estimand must be
  run-local.
- Event onsets and durations are finite physical times. Empty event tables,
  out-of-run onsets, malformed run indices, non-finite coefficients, and
  incomplete or duplicated voxel names fail explicitly.
- The main example uses unit-amplitude impulses, for which the coefficient is a
  peak response. With a nonzero duration or non-unit `events$amplitude`, the
  result is a coefficient on the supplied event design; `units`,
  `event_duration`, and `event_amplitude` record that change. For a
  zero-duration event, multiplying by event amplitude recovers the modeled
  peak. For a nonzero duration, evaluate the duration-coded regressor when a
  peak conversion is needed. A zero-amplitude trial is not identifiable and
  fails explicitly.
- OASIS with a multi-basis HRF returns \(K\) coefficients per trial. Those
  coefficients are not interchangeable with the scalar peak-response
  amplitudes above; retain the full basis result unless you define and verify a
  separate shape-to-amplitude reduction.

## Next step

`vignette("sbhm")` presents a library-constrained alternative for estimating
voxel-specific shapes, score margins, and trial event-design coefficients.
